---
output: hugodown::md_document
title: "S05E05: R for Data Science (2e) - Ch. 8 - Data Import, Part II"
subtitle: "Today, we'll continue with the R4DS on data import (and export)"
summary: "Today, we'll continue with the R4DS on data import (and export)"
authors: [stephen-opiyo]
tags: [codeclub, r4ds]
date: "`r Sys.Date()`"
lastmod: "`r Sys.Date()`"
toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE, message = TRUE, cache = FALSE)
knitr::opts_knit$set(root.dir = here::here("content/codeclub/S05E06"))
```

<br>

--------------------------------------------------------------------------------

## Introduction

Today we will continue with the
[R for Data Science chapter 8 on importing data](https://r4ds.hadley.nz/).

We will cover a few more tricks to import data with _readr_,
and will also cover _exporting_ data.
We will talk about:

1) Controlling column types when reading data,

2) Reading data from multiple files, and

3) Writing to a file.

We will again be using the `tidyverse` and `janitor` packages,
so we first need make sure these packages are installed,
and then load them for the current session using `library()` commands:

```{r}
# If you don't have these installed:
#install.packages("tidyverse")
#install.packages("janitor")

# To load the packages
library(tidyverse)
library(janitor)
```

--------------------------------------------------------------------------------

<br>

## Book Chapter 8.3 and 8.4

Let's switch to [the book](https://r4ds.hadley.nz/data-import.html#sec-col-types)
for this part.

--------------------------------------------------------------------------------

<br>

## Breakout Rooms I

We'll use the `03-sales.csv` file from the examples in the book.
You can download it as follows to your current R working directory:

```{r, eval=FALSE}
url_csv <- "https://github.com/biodash/biodash.github.io/raw/master/content/codeclub/S05E06/03-sales.csv"
download.file(url = url_csv, destfile = "03-sales.csv")
```

<div class="puzzle"><div>

### Exercise 1

Read the `03-sales.csv` file, but with the following twist:
read _all_ columns as _factors_.

<details><summary><b>Hints</b>(click here)</summary>
<br>

Use the `col_types` argument and within that,
call the `cols` function  and specify a `.default`.

</details>

<br>

<details><summary><b>Solution</b>(click here)</summary>
<br>

```{r}
read_csv("03-sales.csv",
         col_types = cols(.default = col_factor()))
```

</details>
</div></div>

--------------------------------------------------------------------------------

<br>

## Book Chapter 8.5 and 8.6

Let's switch back to [the book](https://r4ds.hadley.nz/data-import.html#sec-writing-to-a-file)
for this part.

--------------------------------------------------------------------------------

<br>

## Breakout Rooms I

<div class="puzzle"><div>

### Exercise 2

Write the dataframe that you read in Exercise 1 to a CSV file.
Recall that all columns in the dataframe are stored as _factors_.

Then, read the CSV file you just created back in,
without specifying any additional arguments to `read_csv()`.

Check whether the columns are still factors, and explain why.

<details><summary><b>Hints</b>(click here)</summary>
<br>

- Assign the initial `read_csv()` output to a dataframe,
  then use `write_csv()` to write it to a CSV file.
  
- Recall that a CSV file is a plain text file.
  Can a plain text file store "metadata" about column types?

</details>

<br>

<details><summary><b>Solution</b>(click here)</summary>
<br>

```{r}
sales <- read_csv("03-sales.csv",
                  col_types = cols(.default = col_factor()))

write_csv(sales, "sales.csv")

read_csv("sales.csv")
```

When we read the file back in, the columns are no longer factors but
characters and numerics,
because this sort of information is lost when writing to a plain text file.

</details>
</div></div>

<br>

<div class="puzzle"><div>

### Exercise 3

Repeat what you did in Exercise 2, but now write to and read from an `RDS`
file.

Again, check whether the columns are still factors, and explain why.

<details><summary><b>Hints</b>(click here)</summary>
<br>

Use the `write_rds()` and `read_rds()` functions.

</details>
<br>

<details><summary><b>Solution</b>(click here)</summary>
<br>

```{r}
sales <- read_csv("03-sales.csv",
                  col_types = cols(.default = col_factor()))

write_rds(sales, "sales.rds")

read_rds("sales.rds")
```

The columns are still factors because RDS files preserve all information about
R objects, including column type information.

</details>
</div></div>
