---
output: hugodown::md_document
title: "S04E16: R for Data Science - Exploratory Data Analysis II"
subtitle: "Chapter 7.4 - 7.5: Missing Values and Covariation"
summary: "We continue with this chapter on Exploratory Data Analysis (EDA), now covering missing values (and the ifelse function to turn outliers into missing values) and covariation among variables, with a couple of new plot types: boxplots and heatmaps."
authors: [admin]
tags: [codeclub, r4ds]
date: 2022-12-01
lastmod: 2022-12-01
toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE, message = TRUE, cache = FALSE)
```

--------------------------------------------------------------------------------

## Setting up

Like last time,
we'll mostly use _tidyverse_ tools to explore the `diamonds` dataset,
which is also part of the _tidyverse_.

We'll also have one look at the `flights` dataset,
for which we'll need to load the _nycflights13_ package:

```{r}
## You only need to install packages if you haven't previously done so
# install.packages("nycflights13")
# install.packages("tidyverse")

## But you'll have to load packages for every R session:
library(nycflights13)
library(tidyverse)
```

Let's again take a quick look at the `diamonds` dataset before we begin:

```{r}
diamonds
```

On each row, we have information about one individual diamond,
such as its `carat` and `price`
(`x`, `y`, and `z` represent the diamond's length, width, and depth,
respectively).

Finally,
we'll again set a _ggplot2_ "theme" that is a little better-looking than the
default one
(this setting will apply until you restart R/RStudio):

```{r}
# This changes two things:
# - theme_minimal() gives an overall different look, with a white background
# - base_size = 14 will make the text relatively bigger
theme_set(theme_minimal(base_size = 14))
```

<br>

--------------------------------------------------------------------------------

## Chapter 7.4: Missing values

### Removing outliers

Let's assume you have established that certain outlier values in your data
are untrustworthy.
For instance, see the plot below for the `diamonds` data,
a scatterplot of diamond width (`y`) versus depth (`z`):

```{r}
ggplot(data = diamonds,
       mapping = aes(x = z, y = y)) +
  geom_point()
```

Three rather extreme outliers deviate not just in their absolute values,
but also in their relative values:
overall, depth and width are strongly correlated,
yet the extremely large `z` value does not correspond to a large `y` value at all
(and so on for the other outliers).

To get rid of outliers in your dataset, you have two main options.
First, you could **completely remove rows that contain outliers**,
for example with _dplyr_'s `filter()` function:

```{r}
# Remove rows for which column y is smaller than 3 or larger than 20: 
diamonds_no_outliers <- diamonds %>% filter(y < 3 | y > 20)
```

But you may not want throw out entire rows,
because the values for the _other variables_ (columns) in these rows
might be valid and valuable.
Therefore, and alternative is to **convert outliers to `NA`s (missing values)**,
and a convenient way to do that is with the `ifelse()` function:

```{r}
diamonds_NA_outliers <- diamonds %>%
  mutate(y = ifelse(test = y < 3 | y > 20, yes = NA, no = y))
```

<div class="alert alert-note"><div>

### More on `ifelse()`

To better understand `ifelse()`, a simple example may help.
First, we create a vector with integers from 1 to 10:

```{r}
x <- 1:10
x
```

The following expression will return a logical vector that indicates,
for each value of `x`, whether it is smaller than 5:

```{r}
x < 5
```

We can use that expression as the `test` in `ifelse()`,
and turn values smaller than 5 into `NA` (`yes = NA`),
while leaving the other values unchanged (`no = x`):

```{r}
ifelse(test = x < 5, yes = NA, no = x)
```

</div></div>

### Comparing observations with and without missing data 

It can be useful to compare distributions among rows with and without missing values.
To do that,
we can first create a new column that indicates whether a value for a variable
of interest is missing or not.
Then, we can map an aesthetic like `color` to this missing-or-not column
to show the two groups separately.

Below, we'll compare flights with and without missing values for departure time
(`dep_time`), i.e. cancelled and not-cancelled flights,
using the geom `geom_freqpoly` that we also saw last time:

```{r}
flights %>% 
  mutate(cancelled = is.na(dep_time)) %>% 
  ggplot(mapping = aes(x = sched_dep_time, color = cancelled)) +
  geom_freqpoly(binwidth = 100) # (100 = 1 hour, so we plot by hour)
```

However, the large difference in the absolute counts of cancelled versus
not-cancelled flights makes it hard to see relative differences along the
x-axis.

We can use `geom_density()` to produce a **density plot**,
where the height of the lines is only determined by the relative counts,
allowing us to see if cancelled flights have a different distribution:

```{r}
flights %>% 
  mutate(cancelled = is.na(dep_time)) %>% 
  ggplot(mapping = aes(x = sched_dep_time, color = cancelled)) + 
  geom_density()
```

It looks like flights at the end of the day are much more commonly cancelled
than those early on, which is what we might have expected.

<br>

--------------------------------------------------------------------------------

## Chapter 7.5: Covariation

This section of the book covers the exploration of covariation among two variables.
For example,
is there a relationship between the cut and the price of a diamond
(a categorical and continuous variable)?
Or an association between the cut and color of a diamond
(two categorical variables)?
Or between the carat and the price of a diamond (two continuous variables)?

### 7.5.1: A categorical and continuous variable

In out last plot above,
we already explored the relationship between a categorical variable
(cancelled & not-cancelled flights) and a continuous one (departure time),
and we did so with so with a frequency polygon (`geom_freqpoly`).

Let's see another example, this time for the `diamonds` dataset,
to examine whether prices differ among diamond cuts:

```{r}
ggplot(data = diamonds,
       mapping = aes(x = price, color = cut)) + 
  geom_density()
```

A **boxplot** is another classic way to show the relationship between a
categorical and a continuous variable.
[The book](https://r4ds.had.co.nz/exploratory-data-analysis.html#cat-cont)
has a good explanation of what the components of a boxplot
(box, median line, whiskers, outliers) represent. 
Let's make a boxplot of diamond `price` by `cut`:

```{r}
# To make the plot better-looking, and to see the same colors as above,
# we'll also map the fill aesthetic to cut:
ggplot(data = diamonds,
       mapping = aes(x = cut, y = price, fill = cut)) +
  geom_boxplot() +
  guides(fill = "none") # Just to turn the legend for fill off
```

A less formal, but sometimes more informative variant of this type of plot
is a **violin plot**, where the width represents the number of data points:

```{r}
ggplot(data = diamonds,
       mapping = aes(x = cut, y = price, fill = cut)) +
  geom_violin() +
  guides(fill = "none") # Just to turn the legend for fill off
```

<br>

### 7.5.2: Two categorical variables

As the books says,

> To visualise the covariation between categorical variables,
> you’ll need to count the number of observations for each combination. 

A quick way to do that is with `geom_count()`:

```{r}
ggplot(data = diamonds,
       mapping = aes(x = color, y = cut)) +
  geom_count()
```

A **heatmap** is a slightly more visually appealing way to plot this.
We can create one using **`geom_tile()`**
by first calculating the counts ourselves,
and then mapping these counts to the `fill` color of the tiles:

```{r}
diamonds %>% 
  count(color, cut) %>%  # This will create a column 'n' with the counts
  ggplot(mapping = aes(x = color, y = cut, fill = n)) +
  geom_tile()
```

<br>

### 7.5.3: Two continuous variables

Visualizing the relationship between two continuous variables is perhaps the
most intuitive of the variable type combinations.
It can be done with simple scatterplots (`geom_point()`),
of which we have already seen a couple of examples.

The books covers a few strategies that can be useful when dealing with large datasets,
when relationships may be hidden due to overplotting.
Consider the relationship between the `carat` (weight) and `price` of diamonds:

```{r}
ggplot(data = diamonds,
       mapping = aes(x = carat, y = price)) +
  geom_point()
```

There is clearly some overplotting going on here,
with areas of solid black --
though this type of thing can get a lot worse.
Here, the overall pattern is still apparent.

Making points transparent is one strategy to more clearly see patterns in the
data in the face of overplotting:

```{r}
# An alpha of 1 (the default) is opaque and an alpha of 1 is transparent 
ggplot(data = diamonds,
       mapping = aes(x = carat, y = price)) + 
  geom_point(alpha = 0.01)
```

Two other strategies covered in the book are:

- Use a geom that does 2-dimensional binning for you (e.g. `geom_bin2d`).

- Bin _one_ of the continuous variables,
  effectively turning it into a categorical variable,
  so that we can use plot types like boxplots.
  You'll try that in the exercises.

<br>

--------------------------------------------------------------------------------

## Breakout Rooms

All the exercises use the `diamonds` dataset.
After loading the _tidyverse_, this dataset should be available to you.

<div class="puzzle">
<div>

### Exercise 1

Use the function `cut_number()`
to divide the `carat` values into 10 bins,
and create a boxplot of diamond prices for each of these bins.

_Tip: If you can't read the x-axis labels (bin names) in the resulting plot_
_because they overlap, consider flipping the plot:_
_simply swap the `x` and `y` aesthetic assignments_
_(alternatively, use the stand-alone function `coord_flip()`)._

<details><summary><b>Hints</b> (click here)</summary>
<br>

- Run `?cut_number` to see the documentation for this function.

- You can start by creating a binned column with `mutate()` and
  `cut_number(carat, n = 10)`,
  _or_ you can create the bins "on the fly",
  by simply using `cut_number(carat, n = 10)` as the `x` or `y` aesthetic.

</details>

<br>

<details><summary><b>Solution</b> (click here)</summary>
<br>

To be able to read the axis labels, I moved `carat` to the y axis
(and I also added a y-axis label):

```{r}
diamonds %>%
  ggplot(mapping = aes(y = cut_number(carat, n = 10), x = price)) + 
  geom_boxplot() +
  labs(y = "carat (binned)")
```

The book has a different way of doing this, using the `group` aesthetic.
This is less intuitive but has the advantage of keeping the carat axis labels
as if it still were a regular continuous variable.
It also makes the width of the boxes represent the width of the bins,
which you may or may not like:

```{r}
diamonds %>% 
  ggplot(mapping = aes(x = carat, y = price)) + 
  geom_boxplot(mapping = aes(group = cut_number(carat, n = 10)))
```

Alternatively, if you want to have control over the ranges that the bins cover
(but not over the number of points in each bin),
you can use the `cut_width()` function:

```{r}
diamonds %>%
  ggplot(mapping = aes(y = cut_width(carat, width = 0.4), x = price)) + 
  geom_boxplot() +
  labs(y = "carat (binned)")
```

In that case, consider using `varwidth = TRUE` to make the width of the boxes
to reflect the number of data points:

```{r}
diamonds %>%
  ggplot(mapping = aes(y = cut_width(carat, width = 0.4), x = price)) + 
  geom_boxplot(varwidth = TRUE) +
  labs(y = "carat (binned)")
```

</details>
</div></div>

<br>

<div class="puzzle">
<div>

### Exercise 2

- Create a heatmap (`geom_tile()`) that shows the _mean diamond price_
  for each diamond color and cut combination
  (you'll have to compute the mean price first).

- From your heatmap,
  would you say that going from color `D` => `J` is associated with an _increase_
  or a _decrease_ in the mean price?

_Tip: add `+ scale_fill_viridis_c()` to your code for a much nicer color scale._

<details><summary><b>Hints</b> (click here)</summary>
<br>

- In the heatmap, you'll want `color` along the `x` axis and `cut` along the `y`
  axis (or vice versa), and you'll want to `fill` the tiles by the mean price.
  
- You'll first have to compute the mean diamond price for each of the `color`-`cut`
  combinations: use `group_by()` and then `summarize()`.

</details>

<br>

<details><summary><b>Solution just for getting the mean price</b> (click here)</summary>
<br>

```{r}
diamonds %>%
  group_by(color, cut) %>%
  summarize(price = mean(price))
```  

</details>

<br>

<details><summary><b>Full solution</b> (click here)</summary>
<br>

```{r}
diamonds %>%
  group_by(color, cut) %>%
  summarize(price = mean(price)) %>%
  ggplot(mapping = aes(x = color, y = cut, fill = price)) +
  geom_tile() +
  scale_fill_viridis_c()
```

It looks like going from color `D` to `J` is associated with an overall
_increase_ in the mean price of diamonds.

</details>
</div></div>

<br>

<div class="puzzle">
<div>

### Exercise 3

- Make a plot to visualize the relationship between `color` and `carat`.

- Seeing this relationship, and knowing that `carat` and `price` are strongly related
  (see the scatterplots above),
  do you think this might have influenced (confounded)
  the apparent relationship between `color` and `price`?

- How could you quickly create a price index that controls for `carat`?
  Make a heatmap with that price index instead of the raw price. 

<details><summary><b>Hints</b> (click here)</summary>
<br>

- Use a boxplot, violin plot and/or a density plot to visualize the relationship
  between `color` and `carat`.

- If higher carats are causally associated with higher prices,
  and certain colors have higher mean carats than others,
  it is not fair to look at the effect of color on price without
  somehow taking carat into account.
  
- A simple way of taking carat into account is by using "price per carat"
  (price divided by carat) rather than the raw price in your heatmap.

</details>

<br>

<details><summary><b>Solution</b> (click here)</summary>
<br>

- To visualize the relationship between `color` and `carat`,
  you could for example use a boxplot and/or a density plot:

```{r}
diamonds %>%
  ggplot(aes(x = color, y = carat)) +
  geom_boxplot()
```

```{r}
diamonds %>%
  ggplot(aes(x = carat, color = color)) +
  geom_density()
```

- It looks like going from `D` => `J`, `carat` is typically higher.

- Since `carat` is strongly positively associated with `price`,
  it is not fair to compare prices among colors without controlling for
  `carat`.
  
- A simple way to do so is dividing `price` by `carat`
  to create an index that represents the "price per carat".
  Then, you can use that index instead of the raw price in your heatmap:

```{r}
diamonds %>%
  group_by(color, cut) %>%
  mutate(price_per_carat = price / carat) %>% 
  summarize(price_per_carat = mean(price_per_carat)) %>%
  ggplot(mapping = aes(x = color, y = cut, fill = price_per_carat)) +
  geom_tile() +
  scale_fill_viridis_c()
```

- Now, it looks like going from `D` => `J` is associated with a **decrease** rather
  than an increase in the mean price!

</details>
</div></div>

<br>

<div class="puzzle">
<div>

### Exercise 4 (bonus)

To get another perspecective on the relationship between `color`, `carat`, and
`price` (see the previous exercise),
modify the earlier scatterplot of `carat` and `price` simply by mapping
diamond `color` to the color aesthetic.
  
<details><summary><b>Solution</b> (click here)</summary>
<br>

A scatterplot of `carat` and `price` that includes diamond color confirms
that `D` diamonds are more expensive than `J` diamonds (and so on) once you
take `carat` into account:

```{r}
diamonds %>%
  ggplot(aes(x = carat, y = price, color = color)) +
  geom_point()
```

</details>
</div></div>

<br>

--------------------------------------------------------------------------------

## Bonus: (re)ordering factor levels

In the plots with diamond `cut`,
you might have noticed that the cuts are ordered in a custom, sensible way
rather than alphabetically.
This is possible because the `cut` column has the data type `factor`.

If we convert `cut` to a regular character data type,
the custom order disappears (it is now ordered alphabetically):

```{r}
diamonds %>%
  mutate(cut = as.character(cut)) %>% 
  ggplot(mapping = aes(x = cut, y = price)) +
  geom_boxplot()
```

We could set a different custom order using the `levels` argument of the `factor`
function
(the same code would work if `cut` would not yet have been a factor at all):

```{r}
cut_order <- c("Very Good", "Fair", "Good", "Ideal", "Premium")

diamonds %>%
  mutate(cut = factor(cut, levels = cut_order)) %>% 
  ggplot(mapping = aes(x = cut, y = price)) +
  geom_boxplot()
```

Or, you could make the order of the factor levels depend on the data (!):

```{r}
diamonds %>%
  mutate(cut = reorder(cut, price, FUN = median)) %>% 
  ggplot(mapping = aes(x = cut, y = price)) +
  geom_boxplot()
```

<br>
