---
output: hugodown::md_document
title: "S05E01: R for Data Science (2e) - Ch. 5 - Pipes"
subtitle: "Introducing a new season of Code Club, in which we will continue to read the book R for Data Science (R4DS), and start with a short chapter on pipes."
summary: "We'll continue with the R for Data Science book in the 5th season of Code Club, now switching to the brand-new second edition! We'll talk about pipes in R, which is covered in the short Chapter 5 of the book."
authors: [admin]
tags: [codeclub, r4ds]
date: "`r Sys.Date()`"
lastmod: "`r Sys.Date()`"
toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE, message = FALSE, cache = FALSE)
knitr::opts_knit$set(root.dir = here::here("content/codeclub/S05E01_pipes"))

library(tidyverse)
```

<p align="center">
<img src=img/r4ds_cover.png width="45%">
</p>

------------------------------------------------------------------------

## Intro to this Code Club Season

#### Organizers

- _Michael Broe_ -- Evolution, Ecology and Organismal Biology (EEOB)
- _Jessica Cooperstone_ -- Horticulture & Crop Science (HCS) / Food Science & Technology (FST)
- _Stephen Opiyo_ -- Molecular & Cellular Imaging Center (MCIC) - Columbus
- _Jelmer Poelstra_ -- Molecular & Cellular Imaging Center (MCIC) - Wooster

#### Code Club practicalities

- In-person (Columbus & Wooster) and Zoom hybrid

- Mix of instruction/discussion with the entire group,
  and exercises in groups of 3-4 people.

- When doing **exercises in breakout groups**, we encourage you:
  - To briefly introduce yourselves and to do the exercises as a group
  - On Zoom, to turn your cameras on and to have someone share their screen
    (use the `Ask for help` button in Zoom to get help from an organizer)
  - To let a _less_ experienced person do the screen sharing and coding
  
- You can **ask a question** at any time, by speaking or typing in the Zoom chat.

- You can come up to 15 minutes early or stay late for
  **troubleshooting** and perhaps a question related to your research.

More general notes:

- If you can, read or skim the relevant (part of the) chapter before each session,
  especially if you're very new to the material.
  But we'll always try to present it in such a way that does not assume
  you've read it.

- We try to make each session **as stand-alone as possible**,
  and don't require you to know anything.
  That said, if you missed one or more sessions,
  you'll get more out of the next ones if you try to catch up with the material.
  
- We **record** the whole-group parts of the Zoom call,
  and share the recordings only with Code Club participants.

<div class="alert alert-note"><div>

#### New to Code Club or R?

Take a look at these pages on our website:

- [Computer setup for Code Club](/codeclub-setup/)
- [Resources and tips to get started with R](/tutorials/r-resources-tips/)
- [List of all previous Code Club session topics](/codeclub-schedule/#previous-semesters)

</div></div>

Before moving on the the chapter on pipes,
I will start with a very brief overview of the book, the RStudio interface,
and how to load R packages.

<br>

------------------------------------------------------------------------

## R for Data Science (R4DS)

This excellent book by Hadley Wickham
(also author of many of the R packages used in the book!)
and Garret Grolemund
is [freely available online](https://r4ds.hadley.nz/).

The book focuses on the so-called **"_tidyverse_" ecosystem** in R.
The _tidyverse_ can be seen as a **modern dialect of R**.
Most of its functionality is also contained in "base R"
(that which comes shipped with R by default),
but it has an improved and more consistent programming interface or "syntax".

Last year in Code Club, we worked through the material of a number of chapters
of the [first edition of the book](https://r4ds.had.co.nz/),
which was published in 2016.

Since 2016, quite some R development has taken place.
A **second edition** has been online since a couple of months,
with completely updated and also restructured contents --
we thought it has improved a lot!

This new edition is not _completely_ finished yet,
so you'll find notifications like these at the top of each chapter:

<p align="center">
<img src=img/work-in-progress-warning.png width="95%">
</p>

We decided not to restart at the beginning of the book for this semester.
We hope this won't make it too challenging for beginners to join us.
Especially in the first sessions,
we'll make sure to explain all code,
including things that were covered last semester.

#### What's in the book

The introductory chapter of the book has this figure to show the data science
process and what the book will cover:

<p align="center">
<img src=img/data-science.png width="80%">
</p>

In terms of what the book does _not_ cover,
it may especially be surprising for a book about data science
that it contains very little material on _statistics_
(even less so in the second edition, now that there is a companion book
["Tidy Modeling with R"](https://www.tmwr.org/) that focuses on this).

<br>

------------------------------------------------------------------------

## Getting Up and Running

### RStudio interface

R itself simply provides a "*console*" (command-line interface) where you can type your commands.
RStudio, on the other hand, allows you to see the R console side-by-side with your scripts, plots, and more.

Once you have a running instance of RStudio,
**create a new R script** by clicking `File` > `New File` > `R Script`.
Now, you should see all 4 "panes" that the RStudio window is divided into:

- _Top-left_: The **Editor** for your scripts and other documents (_hidden when no file is open_).
- _Bottom-left_: The **R Console** to interactively run your code (+ other tabs).
- _Top-right_: Your **Environment** with R objects you have created (+ other tabs).
- _Bottom-right_: Tabs for **Files**, **Plots**, **Help**, and others.

<p align="center">
<img src=img/rstudio-layout-ed.png width="95%">
</p>

<br>

<div class="puzzle"><div>

### Your turn: Check your R version

Take a look at your _**version of R**_:
this was printed in the console when you started RStudio
(see the RStudio screenshot above).

The most recent version of R is `4.2.2`.
To use the current functionality of the "base R pipe",
**you'll need at least version `4.2.0`**
(and to use it at all, you need at least R version `4.1`).
If you have a lower version of R,
I would recommend that you update at the end or after this session following
[these instructions](/tutorials/r-resources-tips/#updating-r).
  
</div></div>

<br>

### R packages

You can think of packages as "add-ons" / "extensions" to base R.

#### Installation versus loading

Packages have to be separately **installed** (usually _from within R_, using R code)
and once you have done this,
you don't need to redo it unless you have switched to a different version of R.

Unlike installation, **loading** a package is necessary again and again,
in every R session that you want to use it.

#### The tidyverse

The _tidyverse_ is unusual in that it is a _collection_ of packages
that can still be installed and loaded with a single command.
The individual core _tidyverse_ packages are the focus of several chapters in the book,
for instance:

| Package         | Functionality                     | Main chapter
|-----------------|-----------------------------------|-----------------|
| **_ggplot2_**   | Creating plots                    | Ch. 2
| **_tidyr_** & **_dplyr_** | Manipulating dataframes | Ch. 4 & 6
| **_readr_**     | Reading in data                   | Ch. 8
| **_stringr_**   | Working with "strings" (text)     | Ch. 16
| **_forcats_**   | Working with "factors" <br> (categorical variables) | Ch. 18
| **_purrr_**     | Iteration with functions          | Ch. 28

<div class="puzzle"><div>

### Your turn: Load the tidyverse

To check if you can load the _tidyverse_,
run the following and see if you get similar output as printed below:

```{r, message=TRUE, warning=TRUE}
library(tidyverse)
```

<p align="center">
<img src=img/load_tidyverse.png width="90%">
</p>

If instead, you got something like...

```{r}
#> Error in library(tidyverse) : there is no package called ‘tidyverse’
```

...then you still need to install it (`install.packages("tidyverse")`).

</div></div>

<br>

----

## Chapter 5: Pipes

### What is a pipe?

A pipe is a programming tool that takes the output of one command (in R, a _function_)
and passes it on to be used as the input for another command.

Pipes prevent you from having to save intermediate output to a file or object.
They also make your code shorter and easier to understand.

### An example

In R, we work a lot with "dataframes",
rectangular data structures like spreadsheets --
and in particular, the R4DS book and the _tidyverse_ focus on this very heavily.

So let's see an example of using a pipe with the `diamonds` dataframe,
which is automatically loaded along with the _tidyverse_.
It contains information on almost 54,000 diamonds (one diamond per row):

```{r}
diamonds

# If you get 'Error: object 'diamonds' not found', then the tidyverse isn't loaded.
# Use 'library(tidyverse)' to do so.
```

Let's say we want to subset this dataframe so it only contains diamonds
that cost over $1,000, and only shows the columns `carat`, `cut`, and `price`.

Without using pipes,
we could start by selecting the columns of interest with the `select()` function,
and saving the output in a new dataframe called `diamonds_simple`:

```{r}
# The first argument is the input dataframe, the others are the columns we want
diamonds_simple <- select(diamonds, carat, cut, price)
```

Next, we take `diamonds_simple` and use the `filter()` function
to only select the diamonds (_rows_) that we want:

```{r}
filter(diamonds_simple, price > 1000)
```

But this can be done much more elegantly,
and without wasting computer memory on an intermediate object, using the pipe:

```{r}
diamonds |>
  select(carat, cut, price) |>
  filter(price > 1000)
```

(Above, we used the pipe not just in between `select()` and `filter()`,
but also _before_ `select()`.
That adds a line but also makes it even easier to see what's being done.)

#### Two Unix & R examples

Pipes originate in Unix terminals, and are ubiquitous there.
So for those of you that are curious,
I've included two examples of using the Unix pipe,
and the corresponding commands in R, in the dropdown box below.

<details><summary><b>See the examples</b> (click here)</summary>
<br>

_(If you're trying to follow along yourself:_
_the Unix/terminal examples will only work natively on Mac and Linux,_
_where you can simply click the `Terminal` tab in the bottom-right RStudio_
_panel, and issue Unix commands.)_

**Counting files**

You might want to _count the number of files in a folder_,
which involve two distinct processes: obtaining a list of files, and counting them.

We can get a **l**i**s**t of files in the current folder with `ls`,
perform the counting with `wc -l` (**w**ord**c**ount -**l**ines),
and connect these processes with the pipe `|`:

```{bash}
# The output of 'ls' is piped (with '|') to 'wc -l':
ls | wc -l
```

We can do the same in R, where the function `dir()` lists files,
while the function `length()` counts the number of elements:

```{r}
dir() |>  length()
```

<br>

----------

**Counting word frequencies**

As another example,
let's say we have a file `words.txt` that contains one word per line:

```
table
chair
desk
chair
desk
table
chair
```

In a terminal,
we can get a list of unique words and their number of occurrences using:

```{bash}
# 'cat' prints the contents of the file
# 'sort' sorts alphabetically
# 'uniq -c' counts the number of occurrences for each entry 
cat words.txt | sort | uniq -c
```

And to do the same thing in R:

```{r}
# 'readLines()' reads the contents of a file into R
# 'table()' counts the number of occurrences for each entry
readLines("words.txt") |> table()
```

</details>

### The other pipe and a keyboard shortcut

Those of you who've worked with R for a bit are likely familiar with
**another pipe symbol: `%>%`**.

This pipe is is loaded as part of the _tidyverse_,
and until recently was very widely used including in the
the previous edition of R4DS.
There has been a gradual switch to the base R pipe since it was introduced
in May 2021, mainly because it does not rely on a package.
In addition, it's convenient that the base R pipe `|>` is more similar to the
Unix pipe `|`, and is one fewer character to type than `%>%`.

The number of characters shouldn't make much of a difference, though,
because it remains even quicker to use the
**RStudio keyboard shortcut for the pipe,**
**which is <kbd>Ctrl</kbd>+<kbd>Shift</kbd>+<kbd>M</kbd>.**

<div class="puzzle"><div>

### Your turn: Use the `|>` pipe by default

To make that keyboard shortcut _map to the base R pipe_ (instead of to `%>%`),
go to `Tools` in the top menu bar,
click `Global Options`,
click `Code` in the left menu,
and check the box `Use native pipe operator, |> (requires R 4.1+)`:

<p align="center">
<img src=img/rstudio-pipe-options.png width="60%">
</p>

</div></div>

<br>

There are some differences in the behavior of the `|>` and `%>%` pipes
in more advanced use cases, which the book chapter goes into.

If you ever need to use the pipe with a function where the piped data is not
the first argument,
see the Bonus section below about using the `_` placeholder.

<div class="puzzle"><div>

### Your turn: Use the pipe

With one single "pipeline" (operations connected by a pipe `|>`),
manipulate the `diamonds` dataframe such that you:

- print only the columns `cut`, `color`, `depth`, and `price`, and
- print only diamonds (rows) with a `depth` of less than 50.

How many diamonds have a `depth` of less than 50?

<details><summary><b>Hints</b>(click here)</summary>
<br>

This is quite similar to the example given above:
use the `select()` function to select certain columns,
and the `filter()` function to select certain rows.

</details>

<br>

<details><summary><b>Solution</b>(click here)</summary>
<br>

```{r}
diamonds |>
  select(.data = _, cut, color, depth, price) |>
  filter(.data = _, depth < 50)
```

</details>
</div></div>

<br>

### Bonus: Using the `_` placeholder

The R pipe passes its contents to the _first argument_ of a function.
In other words, the code we saw above is shorthand for the following,
where **`_`** is the placeholder for the data that the pipe passes on:

```{r}
diamonds |>
  select(.data = _, carat, cut, price) |>
  filter(.data = _, price > 1000)

diamonds %>%
  select(., carat, cut, price) %>%
  filter(., price > 1000)
```

What if we need our piped data to go to another argument
than the function's first one?

Let's see an example with the `gsub()` function,
which can be used to replace characters in text strings as follows:

```{r}
gsub(pattern = "N", replacement = "-", x = "ACCGNNT")

# Or, without naming the arguments:
# gsub("N", "-", "ACCGNNT")
```

As you could see above,
the input data (the string `x`) is not the first but the third argument to `gsub()`.
To use the pipe with `gsub()`, use the `_` placeholder:

```{r}
"ACCGNNT" |> gsub(pattern = "N", replacement = "-", x = _)
```

If you're wondering how you can know the order of the arguments of a function,
watch the pop-up box when you type a function's name and opening parenthesis
(see the screenshot below),
or check the help with e.g. `?gsub`.

<p align="center">
<img src=img/gsub.png width="45%">
</p>

----------

Above, I mentioned that the pipe passes its contents to the _first argument_
of a function.
But to be more precise,
the pipe passes the object to _the first argument that you didn't mention by name_.
Therefore, the following also works:

```{r}
# The piped data is being passed to the 3rd argument, 'x',
# which is the first of the function's arguments that we don't refer to below: 
"ACCGNNT" |> gsub(pattern = "N", replacement = "-")
```

Additionally, make sure you always _name_ the argument that you pass `_` to:

```{r, error=TRUE}
# Using '_' without the argument name ('x=') doesn't work:
"ACCGNNT" |> gsub(pattern = "N", replacement = "-", _)
```

<br>

----
