---
output: hugodown::md_document
title: "Session 12: Vectorization and loops in R"
subtitle: "Need for repeat."
summary: "This is be the first of several sessions broadly about iterating in order to avoid copying-and-pasting of code, and today we will talk about vectorization and for loops."
authors: [admin]
tags: [codeclub, markdown, rmarkdown]
date: 2021-03-01
lastmod: 2021-03-01
toc: true

# Featured image
# To use, add an image named `featured.jpg/png` to your page's folder.
# Focal points: Smart, Center, TopLeft, Top, TopRight, Left, Right, BottomLeft, Bottom, BottomRight.
image:
  caption: "Artwork by @allison_horst"
  focal_point: ""
  preview_only: false

# Projects (optional).
#   Associate this post with one or more of your projects.
#   Simply enter your project's folder or file name without extension.
#   E.g. `projects = ["internal-project"]` references `content/project/deep-learning/index.md`.
#   Otherwise, set `projects = []`.
projects: []
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, eval=TRUE, message=FALSE, cache=TRUE)
```

<br> <br> <br>

------------------------------------------------------------------------

## Setup

#### New to Code Club?

-   If you didn't already do this, please follow the
    [Code Club Computer Setup](/codeclub-setup/04_ggplot2/) instructions,
    which also have pointers for if you're new to R or RStudio.

-   If you're able to do so, please open RStudio a bit before Code Club starts --
    and in case you run into issues, please join the Zoom call early and we'll help
    you troubleshoot.

#### Session goals

Today, you will learn:

- That you should avoid copying your code.
- What different strategies for *iteration* exist in R.
- What *vectorization* is and how to make use of it.
- How to write a `for` loop.
- Best practices when using `for` loops.
- When you should (not) use `for` loops.
- Bonus: `if` statements.

<br>

------------------------------------------------------------------------

## Introduction

### Don't Repeat Yourself

Sometimes, you have a block of code and you need to repeat the operations in that code *almost* exactly. For instance, you may want to rerun a statistical model with different parameter values, rerun an analysis for a different batch of samples, or extract the same information for many different genes.

Your first instinct may be to copy-and-paste the block of code, and make the necessary slight adjustments in the pasted block. However, ***iterating*** and ***writing your own functions*** are strategies that are clearer, less error-prone, and more flexible (and these two can also be combined). When the number of repeats are high, *iteration* is needed. When the code that needs to be repeated is more than a line or two, *writing your own functions* becomes useful.

### Iteration

***Loops*** are the most universal iteration tool and the one we will focus on today.
However, R has "functional programming" iteration methods that are less verbose and that can also be quicker to execute. These are the `apply` family of functions, and a more recent *tidyverse* approach implemented in the *purrr* package: we will learn more about those in the two upcoming Code Club sessions.

Loops are still a very good place to start using iteration because they they make the iteration explicit and are therefore more intuitive. Moreover, they can easily accommodate longer blocks of code without the need to *also* write your own function.

We will talk about the most common type of loop: the `for` loops. (Other types of loops in R are `while` loops and `repeat` loops.)

#### But first...

Before we talk about loops we should take a step back and explore ***vectorization*** a bit more,
which was briefly introduced by Michael in [Code Club session 9](https://biodash.github.io/codeclub/09_subsetting/).
Besides functional programming methods, vectorization is the other reason that loops are not as widely used in R as in other programming languages.

------------------------------------------------------------------------

## I: Vectorization

Let's say we have a *vector* (i.e., a collection of values) that consists of distances in miles:

```{r}
dists_miles <- c(24, 81, 48, 29, 177, 175, 20, 11, 62, 156)
```

Of course, we can't science with miles, so we'll have to convert these distances to kilometers by multiplying each value in the vector by 1.61. You may or may not know that this can be done *really easily* in R:

```{r}
dists_km <- dists_miles * 1.61
dists_km
```

What is happening here is called a ***vectorized operation***: `1.61` is automatically recycled as many times as needed to be multiplied with *each individual value* in the `dist_miles` vector. This is a pretty unique and very useful feature of R!

In many other languages, we would need a ***loop*** or a similar construct to iterate over each value in the vector and multiply by 1.61. In fact, under the hood, R *also* uses a loop to do this! So does it even make a difference? Yes -- the advantages of using vectorization in R are:

-   You don't have to write the loop, saving you a fair bit of typing and making the code clearer.
-   The under-the-hood-loop is being executed *much* faster than a loop that you would write. This is because it is written in `C`/`C++` code which only has to be called once rather than at least as many times as there are iterations in our loop.

<br>

### Other vectorization patterns

Above, we saw an example of multiplying a vector by a single number. We can also use vectorized operations when both objects contain multiple items. For instance, say we have a vector with corresponding values for two dates:

```{r}
dists_Mar4 <- c(17, 93, 56, 19, 175, 40, 69, 267, 4, 91)
dists_Mar5 <- c(87, 143, 103, 223, 106, 18, 87, 72, 59, 5)
```

To get the sum of these values at each position (index) of the two vectors (`17 + 87`, `93 + 143`, etc.), we can simply do the following:

```{r}
dists_Mar4 + dists_Mar5
```

<br>

<div class="alert alert-note"><div>
The two vectors don't need to be of equal length, either:  
in the example below, we negate **_every other_** value in a vector:

```{r}
c(17, 93, 56, 19, 175, 40, 69, 267, 4, 91) * c(-1, 1)
```

</div></div>

<br>

This also works for columns of a data frame, which we can extract using the `dataframe_name$column_name` notation (see [Code Club session 9's section on
data frames](https://biodash.github.io/codeclub/09_subsetting/#data-frames),
and the Base R data frame indexing summary [below](#base-r-data-frame-indexing)).
Let's say we wanted the mean distance this time:

```{r}
dist_df <- data.frame(dists_Mar4, dists_Mar5)

dist_df$dists_mean = (dist_df$dists_Mar4 + dist_df$dists_Mar5) / 2
head(dist_df)
```

<br>

### Vectorization with matrices

Furthermore, we can also perform vectorized operations on *entire matrices*.
With the following matrix:

```{r}
# We use the "sample" function to get 25 random values between 1 and a 100,
# and put those in a 5*5 matrix:
mat <- matrix(sample(1:100, 25), nrow = 5, ncol = 5)
mat
```

...we could multiple all values by 10 or get the square of each value simply as follows:

```{r}
mat_more <- mat * 10
mat_more

mat_squared <- mat * mat
mat_squared
```

<br>

### Vectorization with indices

We can also use vectorized solutions when we want to operate only on elements
that satisfy a certain condition.

Let's say we consider any distance in one of our vectors that is below 50 to be
insufficient, and we want to turn those values into negatives.

To do so, we make use of R's ability to index a vector with a logical vector:

```{r}
# "not_far_enough" will be a vector of logicals:
not_far_enough <- dists_Mar4 < 50
not_far_enough

# When we index the original vector with a logical vector,
# we get only those values for which "not_far_enough" is TRUE:
dists_Mar4[not_far_enough]
```

With the following syntax, we can replace just those low distances in our
original vector:

```{r}
dists_Mar4[not_far_enough] <- dists_Mar4[not_far_enough] * -1
dists_Mar4
```

<div class="alert alert-note"><div>
In a simple case like this, we could also use the vectorized `ifelse()` function:

```{r}
ifelse(dists_Mar5 < 50, dists_Mar5 * -1, dists_Mar5)
```

</div></div>

<br>

------------------------------------------------------------------------

## II: For loops

While it is important to use vectorization whenever possible, it can only be applied to a specific set of problems. A more universal solution when you need to repeat operations is the `for` loop. `for` loops iterate over a collection of values, allowing you to perform one or more actions for each value in the collection.

The basic syntax is as follows:

``` r
for (variable_name in collection_name) {
  #...do things with variable_name...
}
```

On the first line, you initialize the `for` loop, telling it to assign each value in the collection to a variable (here, `variable_name`) **one at a time**.

The *variable name* is arbitrary, and the *collection* is whatever you want to loop over. However, `for`, the parentheses `()`, `in`, and the curly braces `{}` are all fixed elements of `for` loops. A simple example will help to understand the synax:

```{r}
# A loop to print negated values:
for (one_number in c(1, 2, 3, 4)) {
  print(one_number * -1)
}
```

Note that we don't *have to* use the variable that we are looping over: we could also use a `for` loop as a roundabout way to simply repeat something as many times as there are values in our collection:

```{r}
for (dummy in c(1, 2, 3, 4)) {
  print("Yes!")
}
```

As mentioned, the **variable name** that we assign is arbitrary: we could use anything, as long as we reference it with the same name inside the loop:

```{r}
# Example 1 with a different variable name: "positive_number"
for (positive_number in c(1, 2, 3, 4)) {
  print(positive_number * -1)
}
```

```{r}
# Example 2 with a different variable name: "i"
for (i in c(1, 2, 3, 4)) {
  print(i * -1)
}
```

Note that the variable as it was last assigned in the loop does **persist in your environment**:

```{r}
i
```

The ***curly braces*** are not strictly necessary for one-liners like this:

```{r}
for (i in 1:4) print(i * -1)
```

<br>

### `for` loop output

Note that we need the `print()` function to print anything to screen -- nothing will be printed if we omit this:

```{r}
for (i in 1:4) {
  i * -1
}
```

Similarly, if we want the output to be saved in an object of some kind, we need to explicitly make an assignment in each iteration of the loop. *This is where we need to start paying attention to the design of our loop.* Unless computational speed is of no concern at all, you should **avoid growing an object in each iteration of the loop**.

For example, you might be inclined to do the following if you wanted to compute the median of each column in a data frame:

```{r}
column_medians <- vector()

for (column_number in 1:ncol(dist_df)) {
  # We can extract a column using "dataframe_name[[column_number]]":
  column_median <- median(dist_df[[column_number]])
  column_medians <- c(column_medians, column_median)
}

column_medians
```

Similarly, you may be adding a column (with `cbind()`) or a row (with `rbind()`) to a data frame in each iteration of the loop. However, the problem with these approaches is that R has to create an entirely new object in each iteration of the loop, because the object's memory requirements keep increasing.

Instead, you'll want to give the final vector (here, `column_medians`) the appropriate size before you start the loop:

```{r}
column_medians <- vector(length = ncol(dist_df))

for (column_number in 1:ncol(dist_df)) {
  column_median <- median(dist_df[[column_number]])
  column_medians[column_number] <- column_median
}
```

Note that for very small problems, such as the example above, there will not be a noticeable difference in computation time between pre-assigning a properly sized object versus growing an object inside the loop. However, it is still good to get into the habit of pre-assigning an object of the right size.

<br>

### Summary guidelines

-   Don't use a loop when you can instead use vectorized operations.
-   Don't grow objects inside the loop. Instead, pre-assign an object large enough to contain all output of the loop and fill it in inside the loop.
-   When you write a loop, avoid doing things *inside* the loop that don't need to be repeated.

Learning about how to create your own functions and/or to use functional programming techniques like *purrr* and the `apply` family of functions (upcoming Code Club sessions!) will likely reduce your reliance on loops. For instance, as we'll see next week, computing the median of each column in a data frame can be done *much* more succinctly with `apply()`.

**Even for more experienced users, loops remain a more viable option when longer blocks of code need to be repeated: we will practice with that in the exercises.** 

<br>

------------------------------------------------------------------------

## Breakout rooms!

For the exercises, you can download an R Markdown file with some code to get
set up (I recommend coding in that document to get a nice overview of the
plots that you produce):

```{r, eval=FALSE}
dir.create('S12')
todays_rmd <- 'https://raw.githubusercontent.com/biodash/biodash.github.io/master/content/codeclub/12_loops/exercises.Rmd'
download.file(url = todays_rmd, destfile = 'S12/exercises.Rmd')
```

The following code is already in your R Markdown file,
which will download and read the bird dataset and the necessary packages:

```{r, eval=FALSE}
## Download the file with bird data:
birds_url <- 'https://raw.githubusercontent.com/biodash/biodash.github.io/master/assets/data/birds/backyard-birds_sample_error.tsv'
birds_file <- 'backyard-birds_sample_error.tsv'
download.file(url = birds_url, destfile = birds_file)
```

```{r, echo=FALSE}
birds_file <- 'backyard-birds_sample_error.tsv'
```

```{r}
## Load the tidyverse:
library(tidyverse)

## Read the file with bird data:
birds <- read_tsv(birds_file)

## Load the maps package and get the state map:
# install.packages('maps')   # first install if necessary
library(maps)
states <- map_data("state")
```

Last week, we learned about making maps. If you attended one of the first few Code Club sessions, you'll recall our Great Backyard Birdcount dataset. Here, we'll use a country-wide random subset of this data (the full file is over 4 GB) to see where Carolina Chickadees were seen:  

```{r}
## With this line, we select only the rows where the column "species_en"
## (English species name) equals "Carolina Chickadee",
## i.e. we are getting just the records for the Carolina Chickadee:
caro_chickadee <- birds[birds$species_en == 'Carolina Chickadee', ]

# Or in tidyverse-speak:
# caro_chickadee <- birds %>% filter(species_en == 'Carolina Chickadee')

# Next, we create a map much like we did last week:
ggplot(data = states,
       mapping = aes(x = long, y = lat, group = group)) +
  geom_polygon(color = "black", fill = "white") +
  geom_point(data = caro_chickadee,
             aes(x = long, y = lat, group = NULL),
             color = "green4", alpha = 0.5) +
  coord_fixed(1.3) +
  labs(title = 'Carolina Chickadee')
```

Uh-oh! Something appears to have gone wrong.
In the first exercise, you'll use vectorization to fix the coordinates.

In the second exercise, you'll use a loop to quickly produce similar plots for several other species.

<div class="puzzle">
<div>

### Exercise 1: Vectorization

Try to fix the coordinates using vectorized operations, and recreate the map
to see if it worked.

- Start with the **latitude**, which is wrong for all points.

<details>
<summary>
Hints (click here)
</summary>
<p>

- You can modify the `caro_chickadee` data frame and keep the plotting code
  exactly the same.

- Simply prepending the latitude column with a minus sign (`-`) will negate the values.

- Equivalent base R ways to refer to (and replace) the column with latitudes
  are `caro_chickadee$lat` and `caro_chickadee[['lat']]`.

</p>
</details>

<br>

<details>
<summary>
Solution (click here)
</summary>
<br>

First we fix the latitude, which was simply negated:

```{r}
caro_chickadee$lat <- -caro_chickadee$lat

## Or equivalently:
# caro_chickadee[['lat']] <- -caro_chickadee[['lat']]

## Or a tidyverse way of doing this:
# caro_chickadee <- caro_chickadee %>% mutate(lat = -lat)
```

Create the first map with the same code as the example:

```{r}
ggplot(data = states,
       mapping = aes(x = long, y = lat, group = group)) +
  geom_polygon(color = "black", fill = "white") +
  geom_point(data = caro_chickadee,
             aes(x = long, y = lat, group = NULL),
             color = "green4", alpha = 0.5) +
  coord_fixed(1.3) +
  labs(title = 'Carolina Chickadee')
```

</p>
</details>

<br>

- Once you have fixed the latitude, you should notice that for one state,
  there is a problem with the **longitude** (the offset is 10 decimal degrees).

<details>
<summary>
Hints (click here)
</summary>
<p>

- The displaced state is North Carolina.

- The states are in the `stateProvince` column, and North Carolina's name is
  simply "North Carolina" in that column.
  
- It may help to first create a logical vector indicating whether for each row
  in the `caro_chickadee` data frame, `stateProvincefor` equals "North Carolina".

- Your final map will look nicer if you get rid of the plotting canvas by
  adding <code>+ theme_void()</code> to the code for the plot.

</p>
</details>

<br>

<details>
<summary>
Solution (click here)
</summary>
<br>

It turns out that North Carolina's chickadees are above the Atlantic.
Let's perform a rescue operation by fixing the longitudes,
which are offset by 10 degrees, **just for North Carolina**:

```{r}
NC_rows <- caro_chickadee$stateProvince == "North Carolina"
caro_chickadee$long[NC_rows] <- caro_chickadee$long[NC_rows] - 10
```

And we create the final map:

```{r}
ggplot(data = states,
       mapping = aes(x = long, y = lat, group = group)) +
  geom_polygon(color = "black", fill = "white") +
  geom_point(data = caro_chickadee,
             aes(x = long, y = lat, group = NULL),
             color = "green4", alpha = 0.5) +
  coord_fixed(1.3) +
  labs(title = 'Carolina Chickadee') +
  theme_void()
```

Nice!

<br>
</details>

</div></div>

<div class="puzzle"><div>

### Exercise 2: for loops

Find the 10 most commonly observed bird species in the data set,
and save their English names (found in the `species_en` column) in a vector.

Feel free to check out the solution if you're not sure how,
because the focus here is on the next step: trying to create a loop.


<details>
<summary>
Solution (click here)
</summary>
<p>

```{r}
top10 <- birds %>%
  count(species_en, sort = TRUE) %>%  # Produces a sorted count table for "species_en"
  pull(species_en) %>%                # Extracts the "species_en" column
  head(n = 10)                        # Take the top 10
```

</p>
</details>

<br>

Next, loop over all the top-10 species to produce a plot for each one of them.
Start with the code for the Carolina Chickadee, including the subsetting operation,
and modify that.

<details>
<summary>
Hints (click here)
</summary>
<p>

- In the subsetting operation where you select data for the focal species,
  replace "Carolina Chickadee" with whatever you name the variable
  (indicating an individual species) that you loop over.
  
  Because this is a variable name, and not a string like "Carolina Chickadee",
  don't forget to omit the quotes.
  
- You'll also need to change the title with the looping variable.

</p>
</details>

<br>

<details>
<summary>Solution (click here)</summary>
<br>

```{r}
for (one_species in top10) {

# Select just the data for one species:
one_bird_data <- birds[birds$species_en == one_species, ]
# Or in tidyverse-speak:
# one_bird_data <- birds %>% filter(species_en == one_species)

p <- ggplot(data = states,
            mapping = aes(x = long, y = lat, group = group)) +
  geom_polygon(color = "black", fill = "white") +
  geom_point(data = one_bird_data,
             aes(x = long, y = lat, group = NULL),
             color = "green4", alpha = 0.5) +
  coord_fixed(1.3) +
  labs(title = one_species) +  # Make sure to change this to the looping variable
  theme_void()
print(p)
}
```

<br>
</details>

</div></div>

### Bonus exercise: if statements

`if` statements are similar in syntax to `for` loops, and are also considered
a "flow control" structure. But their purpose is different from loops:
instead of iterating, they do something once and they *only* do it when a condition
is fulfilled.

For instance, we may want to check in a script whether a certain directory (folder)
exists on our computer, and if it doesn't, then we create the directory:

```{r,eval=FALSE}
if (dir.exists('path/to/my/dir')) {
  warning("Oh my, the output directory doesn't exist yet!")
  dir.create('path/to/my/dir')
}
```

Inside the parentheses `()` after `if` should be a statement that evaluates to either
`TRUE` or `FALSE` (`dir.exists()` will be `TRUE` if the directory exists,
and `FALSE` if it does not).
If it is `TRUE`, whatever is inside the curly braces `{}` will be executed,
and if it is `FALSE`, what is inside the curly braces will be ignored.

`if` statements are commonly combined with `for` loops: we may want to only execute
the functions in our loop for items in our collection that fulfill a certain condition:

```{r}
for (one_number in 1:10) {
  if(one_number > 7) {
    print(one_number)
  }
}
```

In the example above, `one number > 7` will only be `TRUE` for numbers larger than 7.
This example is quite contrived, as it would have been easier (and faster!)
to remove these items from the vector before the loop, but it hopefully gets the
point across.

<div class="puzzle"><div>

Many of the maps we produced in the previous exercise looked quite similar,
with most species very widespread and a few restricted to the east of the US.
Maybe if we select species that haven't been seen in Ohio, we can find some
other distributional patterns.

First, select the the top 50 most observed bird species, just like you did in
exercise 2.

Then, use an `if` statement to create plots only for those top-50 birds
that have _not_ been seen in Ohio.

<details>
<summary>
Solution (click here)
</summary>
<br>

- Select the top-50 birds:

```{r}
all_species <- birds %>%
  count(species_en, sort = TRUE) %>%
  pull(species_en) %>%
  head(n = 50)
```

- Loop over the species:

```{r}
for (one_species in all_species) {

  # Select the focal species:
  one_bird <- birds[birds$species_en == one_species, ]
  
  # Create a data frame with only records from Ohio:
  one_bird_ohio <- one_bird[one_bird$stateProvince == 'Ohio', ]

  # Test whether the data frame with only records from Ohio has any rows.
  # If it does not, we create the map for the species in question: 
  if(nrow(one_bird_ohio) == 0) {
  
    p <- ggplot(data = states,
           mapping = aes(x = long, y = lat, group = group)) +
      geom_polygon(color = "black", fill = "white") +
      geom_point(data = one_bird,
                 aes(x = long, y = lat, group = NULL),
                 color = "green4", alpha = 0.5) +
      coord_fixed(1.3) +
      labs(title = one_species) +
      theme_void()
    print(p)
    }
}
```

<br>
</details>

</div></div>

<br>

------------------------------------------------------------------------

## Going further

<div class="alert alert-note"><div>

### Base R data frame indexing

Extract a column *as a vector*:

```{r, eval=FALSE}
# By name:
birds$lat
birds[['lat']]   # Equivalent, $ notation is shorthand

# By index (column number):
birds[[8]]
```

Extract one or more columns *as a data frame* using `[row, column]` notation,  
with a leading comma (`[, column]`) meaning all rows:

```{r, eval=FALSE}
# By name:
birds[, 'lat']   # dataframe['row_name', 'column_name']
birds[, c('lat', 'long')]

# By index (column numbers):
birds[, 8]       # dataframe[row_number, column_number]
birds[, c(8, 9)]

```

Subset rows by a condition, with a trailing comma (`[row, ]`) meaning all columns:

```{r, eval=FALSE}
birds[birds$lat > 25, ]
birds[birds$species_en == 'Carolina Chickadee', ]
```

</div></div>

<br>

<div class="alert alert-note"><div>

#### `seq_along()`

To loop over column indices, we have used `1:ncol()` above,
and to loop over vector indices, you could similarly use `1:length()`.

However, an alternative is `seq_along()`, which will create an index for you.

```{r}
birds <- c('titmouse', 'chickadee', 'cardinal')
seq_along(birds)
```

The advantage of `seq_along()` is thtat it will behave better when your vector
accidentally has length 0.

</div></div>

#### Further reading

-   [Vectorization in R: Why?](http://www.noamross.net/archives/2014-04-16-vectorization-in-r-why/) (Noam Ross, 2014)
-   The [Iteration chapter](https://r4ds.had.co.nz/iteration.html) in Hadley Wickham's R for Data Science (2017)
