---
output: hugodown::md_document
title: "S04E08: R for Data Science - Chapters 4, 6, and 8"
subtitle: "Covering some R Basics, how to work with R scripts, and introducing RStudio Projects"
summary: "In this first session of Code Club for Fall '22, we will continue working our way through the book R for Data Science. Today, we'll look at three very short chapters on some R and RStudio basics."
authors: [jessica-cooperstone]
tags: [codeclub, r4ds]
date: "`r Sys.Date()`"
lastmod: "`r Sys.Date()`"
toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, eval=TRUE, message=FALSE, cache=FALSE)
knitr::opts_knit$set(root.dir = here::here("content/codeclub/S04E01_r4ds-ch1"))

library(tidyverse)
```

<p align="center">
<img src=img/r4ds_cover.png width="45%">
</p>

------------------------------------------------------------------------

## I -- Brief Code Club Introduction

### Organizers

- _Michael Broe_ -- Evolution, Ecology and Organismal Biology (EEOB)
- _Jessica Cooperstone_ -- Horticulture & Crop Science (HCS) / Food Science & Technology (FST)
- _Stephen Opiyo_ -- Molecular & Cellular Imaging Center (MCIC) - Columbus
- _Jelmer Poelstra_ -- Molecular & Cellular Imaging Center (MCIC) - Wooster
- _Mike Sovic_ -- Infectious Diseases Institute AMSL - Genomics Lab

### Code Club practicalities

- In-person (Columbus & Wooster) and Zoom hybrid

- Mix of instruction/discussion with the entire group,
  and doing exercises in breakout groups of up to 4-5 people.

- When doing **exercises in breakout groups**, we encourage you:
  - To briefly introduce yourselves and to do the exercises as a group
  - On Zoom, to turn your cameras on and to have someone share their screen
    (use the `Ask for help` button in Zoom to get help from an organizer)
  - To let a _less_ experienced person do the screen sharing and coding
  
- You can **ask a question** at any time, by speaking or typing in the Zoom chat.

- You can generally come early or stay late for
  **troubleshooting** but also for questions related to your research.

More general notes:

- We recommend that you **read** the relevant (part of the) chapter before each session,
  especially if the material in the chapter is new to you.

- We try to make each session **as stand-alone as possible**.
  Still, if you missed one or more sessions, you would ideally catch up on reading
  those parts of the book, especially when we split a chapter across multiple sessions. 
  
- We **record** the whole-group parts of the Zoom call,
  and share the recordings only with Code Club participants.
  
### New to Code Club or R?

Take a look at these pages on our website:

- [Computer setup for Code Club](/codeclub-setup/)
- [Resources and tips to get started with R](/tutorials/r-resources-tips/)
- [List of all previous Code Club session topics](/codeclub-schedule/#previous-semesters)

<br>

------------------------------------------------------------------------

## II -- The R for Data Science book (R4DS)

This excellent book by Hadley Wickham
(also author of many of the R packages used in the book!)
and Garret Grolemund, has a freely available
**[online version](https://r4ds.had.co.nz/) that is regularly updated**
and contains exercises. It was originally published in 2016.

The book focuses on the so-called **"_tidyverse_" ecosystem** in R.
The _tidyverse_ can be seen as a **modern dialect of R**.
In previous Code Clubs,
we have often --but not always!-- been doing things "the _tidyverse_ way"
as well.

For today's chapters, The R4DS exercises I think are not so good, 
so I've replaced some and added some of my own.

<br>

--------

## III -- R4DS Chapter 4

In the R4DS exercises for this chapter,
the message is that R does not handle typos so make sure you spell things correctly.

Take a look at the RStudio keyboard shortcuts by clicking
`Tools` > `Keyboard Shortcut Help`,
or you can press <kbd>Alt</kbd> + <kbd>Shift</kbd> + <kbd>K</kbd> on a PC.

<br>

--------

## IV -- R4DS Chapter 6


<div class="puzzle">
<div>

Run the following code:

```{r, eval = FALSE}
glimpse(cars)
```

What went wrong?

<details><summary><b>Solution</b> (click here)</summary>
<br>

`glimpse()` is a function from the _dplyr_ package, one of the core _tidyverse_
packages that are loaded as part of the tidyverse.

However, in every R session in which you want to use tidyverse function,
you always need call `library(tidyverse)`.

```{r}
library(tidyverse)
```

Now you can use `glimpse()`:

```{r}
glimpse(cars)
```

<br>

Note, if you got an error like this when running `library(tidyverse)`: 

```{r}
#> Error in library(tidyverse) : there is no package called ‘tidyverse’
```

...that means you still need to _install_ it:

```{r, eval = FALSE}
install.packages("tidyverse")
library(tidyverse)
```

</details>

</div></div>

<br>

--------

## IV -- R4DS Chapter 8

<div class="puzzle">
<div>

Create an RStudio Project for Code Club.

</div></div>

<div class="puzzle">
<div>

Run the code below in your new Project:

```{r, eval=FALSE}
library(tidyverse)

ggplot(diamonds, aes(carat, price)) + 
  geom_hex()

ggsave("diamonds.pdf")

write_csv(diamonds, "diamonds.csv")
```

- What does the code above do?

- Find the files `diamonds.pdf` and `diamonds.csv` on your computer,
  without using a search function.
  How did you know where to look for them?

- Where is the R working directory on your computer?
 
</div></div>

<br> <br>
