---
output: hugodown::md_document
title: "S04E08: R for Data Science - Chapter 5.1 - 5.3"
subtitle: "Data Transformation part 1: introduction, `filter()`, and `arrange()`"
summary: "This chapter covers the manipulation of rectangular data (data frames, think data from spreadsheets) with the dplyr package that is part of the tidyverse. We'll learn about data frames & tibbles, R variable types, comparison and logical operators, and missing values (NAs) in addition to the first of the core dplyr functions: filter() and arrange()"
authors: [admin]
tags: [codeclub, r4ds]
date: "`r Sys.Date()`"
lastmod: "`r Sys.Date()`"
toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE, message = FALSE, cache = FALSE)
```

<figure><p align="center">
<img src=img/dplyr_filter.jpeg width="70%">
<figcaption>Artwork by Allison Horst</figcaption>
</p></figure>

------------------------------------------------------------------------

## I -- Chapter 5.1: Introduction

### Key points

- Function name conflicts:
  The function `filter()` in the _stats_ package (which is loaded by default in R)
  will be "masked" / "overwritten" by _dplyr_'s `filter()` function when you
  load the _tidyverse_.
  To still use a masked function
  (or a function from an installed-but-not-loaded package!),
  use the "full" notation, e.g. `stats::filter()`.

- A _data frame_ is rectangular data structure (with rows and columns),
  while a "tibble" is a tidyverse-style data frame.
  Tibbles mainly differ from regular data frames in how they are printed to screen
  by default.
  Ssee the two examples below: `cars` is a regular data frame
  and `flights` is a tibble.
  
- The most common R data types are _integers_ (tibble abbreviation: `int`),
  _doubles_ (`dbl`), ) (`chr`), _logicals_ (`lgl`), and _factors_ (`fctr`).

- The _dplyr_ package is designed to work with dataframes:
  both the input and the output is a dataframe.


```{r, eval=FALSE}
# 'mtcars' is a regular dataframe
head(mtcars)
```

```{r, echo=FALSE}
# 'mtcars' is a regular dataframe
head(mtcars)
```

```{r, eval=FALSE}
# 'flights' is a tibble, which affects its printing behavior
library(nycflights13)
head(flights)
```

```{r, echo=FALSE}
# 'flights' is a tibble, which affects its printing behavior
library(nycflights13)
head(flights)
```

<br>

--------

## II -- Chapter 5.2: `filter()`

### Key points

- The `filter()` function removes rows (observations) from a dataframe based
  on conditions. You specify those conditions for one or more columns.

- When you refer to a column, don't quote its name!
  (e.g. in `filter(flights, month == 1)`, where `month` is the column name.)

- Use "comparison operators" like `>` (greater than) to specify conditions.
  Note that _two_ equals signs `==` (and not a single, `=`) signifies equality,
  and that `!=` means "does not equal".

- To combine multiple conditions, use logical (Boolean) operators:
  `&` (and), `|` (or), and `!` (not).
  Separating conditions by a comma also means "and" in _dplyr_,
  e.g. in `filter(flights, month == 1, day == 1)`.

- The `%in%` operator tests if the value(s) on the left-hand side are contained
  in the values on the right hand side,
  e.g. `4 %in% 1:5` asks whether 4 is contained in the sequence of numbers from
  1 to 5, which will return `TRUE`.
  
- Missing values are denoted by `NA`, and almost any operation with an `NA` will
  return another `NA`.
  To test if `x` is or contains `NA`s, don't use `x == NA` but `is.na(x)`.
  When you filter based on a column,
  rows with `NA`s in that column will by default be removed by `filter()`.

<div class="puzzle">
<div>

### Exercise 1

Find all flights that...

1. Had an arrival delay of two or more hours
2. Flew to Houston (`IAH` or `HOU`)
3. Were operated by United (`UA`), American (`AA`), or Delta (`DL`)
4. Departed in summer (July, August, and September)
5. Arrived more than two hours late, but didn’t leave late
6. Were delayed by at least an hour, but made up over 30 minutes in flight
7. Departed between midnight and 6am (inclusive)

Before you start, load the necessary packages:

```{r}
library(nycflights13)
library(tidyverse)
```

<details><summary><b>Hints</b> (click here)</summary>

- Delays are given in minutes.

- Times of day are numbered from `0001` (1 minute past midnight) to
  `2400` (midnight).

</details>

<br>

<details><summary><b>Solution</b> (click here)</summary>
<br>

_In the solutions below, I am piping the output to `nrow()`,_
_so you can check if you got the same number of rows._

\1. Had an arrival delay (=> `arr_delay`) of two or more hours (=> `>= 120`):

```{r, eval=FALSE}
filter(flights,
       arr_delay >= 120) %>%
  nrow()
```

```{r, echo=FALSE}
filter(flights,
       arr_delay >= 120) %>%
  nrow()
```

\2. Flew to Houston (`IAH` or `HOU`) -- destination is the `dest` column:

```{r, eval=FALSE}
filter(flights,
       dest %in% c("IAH", "HOU")) %>%
  nrow()
```

```{r, echo=FALSE}
filter(flights,
       dest %in% c("IAH", "HOU")) %>%
  nrow()
```

\3. Were operated by United (`UA`), American (`AA`), or Delta (`DL`) &mdash;
   this information is in the `carrier` column:

```{r, eval=FALSE}
filter(flights,
       carrier %in% c("UA", "AA", "DL")) %>%
  nrow()
```

```{r, echo=FALSE}
filter(flights,
       carrier %in% c("UA", "AA", "DL")) %>%
  nrow()
```

\4. Departed in summer (July, August, and September) &mdash;
   use the `month` column:

```{r, eval=FALSE}
filter(flights,
       month %in% 7:9) %>%
  nrow()
```

```{r, echo=FALSE}
filter(flights,
       month %in% 7:9) %>%
  nrow()
```

This would also work:

```{r, eval=FALSE}
filter(flights,
       month >= 7, month <= 9) %>%
  nrow()
```

\5. Arrived more than two hours late, but didn’t leave late &mdash;
   use the `arr_delay` (arrival delay) and `dep_delay` (departure delay)
   columns:

```{r, eval=FALSE}
filter(flights,
       arr_delay > 120, dep_delay <= 0) %>%
  nrow()
```

```{r, echo=FALSE}
filter(flights,
       arr_delay > 120, dep_delay <= 0) %>%
  nrow()
```

\6. Were delayed by at least an hour, but made up over 30 minutes in flight &mdash;
   use the `dep_delay` and `arr_delay` columns,
   and note that "making up over 30 miniutes" implies that the arrival delay
   was more than 30 minutes smaller than the departure delay:

```{r, eval=FALSE}
filter(flights,
       dep_delay >= 60, dep_delay - arr_delay > 30) %>%
  nrow()
```

```{r, echo=FALSE}
filter(flights,
       dep_delay >= 60, dep_delay - arr_delay > 30) %>%
  nrow()
```

\7. Departed between midnight and 6am (inclusive) &mdash;
   use the `dep_time` column and note that `2400` is midnight:

```{r, eval=FALSE}
filter(flights,
       dep_time <= 600 | dep_time == 2400) %>%
  nrow()
```

```{r, echo=FALSE}
filter(flights,
       dep_time <= 600 | dep_time == 2400) %>%
  nrow()
```

</details>

</div></div>

<br>

<div class="puzzle">
<div>

### Exercise 3

How many flights have a missing `dep_time`?
What other variables are missing for these flights?
What might these rows represent?

<details><summary><b>Hints</b> (click here)</summary>

- A "missing" `dep_time` means that this cell contains the value `NA`.
- Recall that you can test if something is `NA` with the `is.na()` function!
- To count the number of flights, you can look at the information printed along
  with the dataframe (`... with X more rows`),
  or pipe (`%>%`) the dataframe into the `nrow()` function,
  which counts the number of rows.
  
</details>

<br>

<details><summary><b>Solution</b> (click here)</summary>
<br>

- _How many flights have a missing `dep_time`?_

```{r, eval=FALSE}
filter(flights,
       is.na(dep_time)) %>%
  nrow()
```

```{r, echo=FALSE}
filter(flights,
       is.na(dep_time)) %>%
  nrow()
```

- _What other variables are missing for these flights?_

  For example, arrival times.

- _What might these rows represent?_

  These are cancelled flights.

</details>

</div></div>

<br>

--------

## III -- Chapter 5.3: `arrange()`

Key points:

- The `arrange()` function sorts the rows in a dataframe.

- Sorting is in ascending order (smallest values first) by default &mdash;
  use `desc(column_name)` to sort in descending order.

- `NA` are always sorted at the end.  

<div class="puzzle">
<div>

### Exercise 2

- Sort flights to find the most delayed flights.
- Find the flights that left earliest (relative to the scheduled time).

<details><summary><b>Solution</b> (click here)</summary>
<br>

- If you sort in _descending_ order by `dep_delay`,
  you'll get the most delayed flights at the top:

```{r, eval=FALSE}
arrange(flights, desc(dep_delay))
```

```{r, echo=FALSE}
arrange(flights, desc(dep_delay))
```

- If you sort in _ascending_ order by `dep_delay`,
  you'll get the flights that left earliest relative to the scheduled time
  at the top:

```{r, eval=FALSE}
arrange(flights, desc(dep_delay))
```

```{r, echo=FALSE}
arrange(flights, desc(dep_delay))
```

</details>

</div></div>

<br>

<div class="puzzle">
<div>

### Exercise 4

- Which flights travelled the farthest?
- Which travelled the shortest?

<details><summary><b>Solution</b> (click here)</summary>
<br>

- _Which flights travelled the farthest?_

```{r, eval=FALSE}
arrange(flights, desc(distance))
```

```{r, echo=FALSE}
arrange(flights, desc(distance))
```

- _Which travelled the shortest?_

```{r, eval=FALSE}
arrange(flights, distance)
```

```{r, echo=FALSE}
arrange(flights, distance)
```

</details>

</div></div>

<br> <br>
