---
title: "Session 10: Faceting, multi-plots, and animating"
subtitle: "Getting up close and personal with our data"
summary: "During this tenth session of Code Club, we will be continuing to work on making visualizations using ggplot2, including faceting plots, making multi-panel figures, and animating charts."  
authors: [jessica-cooperstone]
date: "`r Sys.Date()`"
lastmod: "`r Sys.Date()`"
output: hugodown::md_document
toc: true

image: 
  caption: "Artwork by @allison_horst"
  focal_point: ""
  preview_only: false

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

<br> <br> <br>

-----

## Prep homework 

### Basic computer setup

- If you didn't already do this, please follow the [Code Club Computer Setup](/codeclub-setup/04_ggplot2/) instructions, which also has pointers for if you're new to R or RStudio.

- If you're able to do so, please open RStudio a bit before Code Club starts -- and in case you run into issues, please join the Zoom call early and we'll help you troubleshoot.

### New to ggplot?

Check out the two Code Club pages for [Session 4](/codeclub/04_ggplot2/) and [Session 5](/codeclub/05_ggplot-round-2/).

If you've never used `ggplot2` before (or even if you have), you may find [this cheat sheet](https://github.com/rstudio/cheatsheets/blob/master/data-visualization-2.1.pdf) useful.

<br>

----


## Getting Started

### RMarkdown for today's session
```{r, eval = FALSE}
# directory for Code Club Session 2:
dir.create("S10")

# directory for our RMarkdown
# ("recursive" to create two levels at once.)
dir.create("S10/Rmd/")

# save the url location for today's script
todays_R_script <- 
  'https://raw.githubusercontent.com/biodash/biodash.github.io/master/content/codeclub/10_faceting-animating/Session10_faceting_animating_multifigs.Rmd'

# indicate the name of the new script file
Session10_Rmd <- "S10/Rmd/Session10_faceting_animating_multifigs.Rmd"

# go get that file! 
download.file(url = todays_R_script,
              destfile = Session10_Rmd)
```

<br>

----


## 1 - How can I do more with my figures?

<p align="center"> 
<img src=ggplot2_exploratory.png width="95%" alt="ggplot2 exploratory cartoon">
</p>
Artwork by [Allison Horst](https://github.com/allisonhorst/stats-illustrations)

Sometimes we have so much data that it is difficult to make sense of it if we look at it all at once.  One way to get around this is to create [facets](https://r4ds.had.co.nz/data-visualisation.html?q=facet#facets) in your data -- small subplots that help you to see different relationships among different variables in your dataset.  

Today we will be using [`ggplot2`](https://ggplot2.tidyverse.org/) to make a series of plots that help us better understand the underlying structure in our dataset.

<br>
<div class="alert alert-note">
<div>

**What will we go over today**

These functions or packages will help you to get better visualize with your data.

- [`facet_wrap()`](https://ggplot2.tidyverse.org/reference/facet_wrap.html) and [`facet_grid()`](https://ggplot2.tidyverse.org/reference/facet_grid.html)- makes small multiple plots based on some variable.
- set `scales`  to indicate the linked or not-linked nature of your axes in a faceted plot
- [`gghighlight()`](https://www.rdocumentation.org/packages/gghighlight/versions/0.1.0) - allows you to direct focus on a particular portion of your data.
- [`patchwork`](https://patchwork.data-imaginist.com/) - to compose super easy multi-plot figures
- [`gganimate`](https://gganimate.com/articles/gganimate.html) - to make your plots gif!

I will also go over a few tricks along the way.

</div>
</div>

<br>

----

## 2 - Accessing our data

**Let's get set up and grab some data so that we can learn more about this world (and ggplot2)**

* You can do this locally, or at OSC.  You can find instructions if you are having trouble [here](/codeclub-setup/).

First load your libraries.  We are using a lot of new packages today.

If you've never downloaded these packages before, use the chunk below.
```{r, eval = FALSE}
install.packages(c("gghighlight",
                   "gganimate",
                   "patchwork",
                   "ggrepel",
                   "gapminder"))
```

Once you have the packages above downloaded, load your libraries.
```{r, message = FALSE}
library(tidyverse)
library(gghighlight) # for bringing attention to certain parts of your plot
library(gganimate) # for animating
library(patchwork) # for making multi-panel plots
library(ggrepel) # for getting labels to not be on top of your points

# data for today
library(gapminder)
```

Then let's access the dataset `gapminder`, which is both the name of the package, and the name of the dataset.  It contains a subset of data from [Gapminder.org](https://www.gapminder.org/), an educational non-profit aimed to fight global misconceptions about statistics of our world.

<p align="center"> 
<img src=mega_gapminder.png width="95%" alt="a big bubble plot showing the relationship between income and health of nations, with bubble size indicated by population size, and colored by region of the world">
</p>
From [Gapminder.org](www.gapminder.org)


Let's look at the data in `gapminder`.
```{r}
# look at the first 6 rows, all columns
head(gapminder)

# check the structure
# this tell us what is contained within our df
glimpse(gapminder)
```

This dataset contains the following measurements about the life expectancy, population, and GDP per capita for 142 countries from 1952 to 2007.  It includes the following variables:

* `country`
* `continent`
* `year`
* `lifeExp`
* `pop`
* `gdpPercap`

*Note, this data is already in tidy-style format* meaning:

* Each variable must have its own column.
* Each observation must have its own row.
* Each value must have its own cell. 

Learn more about tidy data [here](https://r4ds.had.co.nz/tidy-data.html).

To make things a bit less complex, let's look at data just from the Americas (i.e., North and South America).  To do that, we can use `filter()` like we learned using `dplyr` back in [Code Club Session 2](/codeclub/02_dplyr-core-verbs/)

```{r}
# make a df with data only from the Americas
gapminder_americas <- gapminder %>%
  filter(continent == "Americas")

# what countries do we have?
unique(gapminder_americas$country)
```


<br>

----

## 3 - Life expectancy vs. time

We will plot the relationship between `lifeExp` and `year` with the goal of understanding how life expectancy has changed in the second half of the 20th century.  We will use [`geom_line()`](https://ggplot2.tidyverse.org/reference/geom_path.html#:~:text=geom_line()%20connects%20them%20in,which%20cases%20are%20connected%20together.) to make a line plot.
```{r}
gapminder_americas %>%
  ggplot(aes(x = year, y = lifeExp, group = country, color = country)) +
  geom_line() 
```

This plot has so many countries, and we can only visually distinguish colors so well, that it makes this a bit of a mess.  We can do better! 

<br>

----

## 4 - Highlighting 

What if we want to highlight one country of interest, with the backdrop of all the data in the Americas?  We can do this using [`gghighlight()`](https://www.rdocumentation.org/packages/gghighlight/versions/0.1.0), which will distinguish our country of interest, from the rest of the countries which will be indicated in gray.

Let's highlight the United States, and since we are at it, let's also add x and y axis labels, a title, subtitle, and caption with `labs()`.

```{r}
gapminder_americas %>%
  ggplot(aes(x = year, y = lifeExp, group = country, color = country)) +
  geom_line() +
  gghighlight(country == "United States") +
  labs(x = "Year",
       y = "Life Expectancy (years)",
       title = "Life Expectancy in Countries in the Americas",
       subtitle = "From 1952 to 2007",
       caption = "Data from gapminder.org")
```

<br>

----

## 5 - Faceting

What if we want to see all the data at once, but just be able to better attribute each line to the correct country?  We can use the [principle of small multiples](https://en.wikipedia.org/wiki/Small_multiple#), popularized by Edward Tufte, to make a series of charts all on the same scale to allow comparison between them easily. 

We can facet using [`facet_wrap`](https://ggplot2.tidyverse.org/reference/facet_wrap.html) to create small plots for each country.  If you want a certain number of rows or columns you can indicate them by including `ncol` and `nrow` in the `facet_wrap()` statement.

I have also made the strip text, or the label on top of each of the facets bigger using `theme`.

```{r, fig.width = 14, fig.height = 8}
gapminder_americas %>%
  ggplot(aes(x = year, y = lifeExp, color = country)) +
  geom_line() +
  theme(strip.text.x = element_text(size = 14)) +
  facet_wrap(vars(country)) + # facet_wrap(~country) also works
  labs(x = "Year",
       y = "Life Expectancy (years)",
       title = "Life Expectancy in Countries in the Americas",
       subtitle = "From 1952 to 2007",
       caption = "Data from gapminder.org")
```

Now our legend is not necessary, so let's remove it.  Let's also remove the gray background since its not really doing much for us.  We will also change to `theme_minimal()` to get rid of the grey background which I don't think we need.
```{r, fig.width = 14, fig.height = 8}
gapminder_americas %>%
  ggplot(aes(x = year, y = lifeExp)) +
  geom_line(aes(color = country)) +
  theme_minimal() +
  theme(legend.position = "none",
        strip.text.x = element_text(size = 14)) +
  facet_wrap(~country) +
  labs(x = "Year",
       y = "Life Expectancy (years)",
       title = "Life Expectancy in Countries in the Americas",
       subtitle = "From 1952 to 2007",
       caption = "Data from gapminder.org")
```

Wow better!  But now its a bit hard to contextualize the line for each country to the whole dataset.  We can fix this too.

<br>

----

## 6 - Highlighting plus faceting

Let's bring the rest of data back in, and highlight in each facet the country of interest.  We can do this by just adding `gghighlight()` to our `ggplot` call.

Note: if you want to assign something in R to an object, and then view it, you can put the whole thing in parentheses, without having to call that object back at the end.
```{r, fig.width = 14, fig.height = 8}
(americas_lifeexp <- gapminder_americas %>%
  ggplot(aes(x = year, y = lifeExp)) +
  geom_line(aes(color = country)) +
  gghighlight() +
  theme_minimal() +
  theme(legend.position = "none",
        strip.text.x = element_text(size = 14)) +
  facet_wrap(~country) +
  labs(x = "Year",
       y = "Life Expectancy (years)",
       title = "Life Expectancy in Countries in the Americas",
       subtitle = "From 1952 to 2007",
       caption = "Data from gapminder.org"))
```

Wow, we now have so much more information about our data! 

<br>

----

## 7 - Adjusting scales while faceting
The default in faceting is that the x and y-axes for each plot are all the same.  This aids in the interpretation of each small plot in relation to the others, but sometimes you may want freedom to adjust your axes.

For example, if we wanted to plot population over time, if we used the same scale, it would be really hard to see trends within a country.

```{r, fig.width = 14, fig.height = 8}
(americas_pop <- gapminder_americas %>%
  ggplot(aes(x = year, y = pop)) +
  geom_line(aes(color = country)) +
  theme_minimal() +
  theme(legend.position = "none",
        strip.text.x = element_text(size = 14)) +
  facet_wrap(~country) +
  labs(x = "Year",
       y = "Population",
       title = "Population in Countries in the Americas",
       subtitle = "From 1952 to 2007",
       caption = "Data from gapminder.org"))
```

Let's change the scales so that the y-axis is "free" - i.e., each plot will have an independent y-axis.  Note, when you do this, you aren't really using the principle of small multiples anymore, since the data isn't all on comparable scales.
```{r, fig.width = 14, fig.height = 8}
gapminder_americas %>%
  ggplot(aes(x = year, y = pop)) +
  geom_line(aes(color = country)) +
  theme_minimal() +
  theme(legend.position = "none",
        strip.text.x = element_text(size = 14)) +
  facet_wrap(~country,
             scales = "free_y") +
  labs(x = "Year",
       y = "Population",
       title = "Population of Countries in the Americas",
       subtitle = "From 1952 to 2007",
       caption = "Data from gapminder.org")
```

The default for `scales` is `"fixed"`, but you can also set to be `"free_x"`, `"free_y"`, or `"free"`, which means both x and y are free.


<br>

----

## 8 - Multi-panel plots
What if I take plots I've already made and assemble them together?  You can do that simply with the package [`patchwork()`](https://patchwork.data-imaginist.com/).

<p align="center"> 
<img src=patchwork_1.jpg width="95%" alt="an illustration of how to use patchwork with cute little monsters">
</p>
Artwork by [Allison Horst](https://github.com/allisonhorst/stats-illustrations)

You can use the syntax:

* `plot1 + plot2` to get two plots next to each other
* `plot1 / plot2` to get two plots stacked vertically
* `plot1 | (plot2 + plot3)` to get plot1 in the first row, and plots 2 and 3 in a second row

You can use [`plot_annotation()`](https://patchwork.data-imaginist.com/reference/plot_annotation.html) to indicate your plots with letters or numbers.

I am going to make some quick plots so we can see how it works.  Let's look at some plots of the United States.

```{r}
# make df with just United States data
gapminder_usa <- gapminder %>%
  filter(country == "United States")

# make some plots
(usa_lifeexp <- gapminder_usa %>%
  ggplot(aes(x = year, y = lifeExp)) +
  geom_point())
```

```{r}
(usa_gdppercap <- gapminder_usa %>%
  ggplot(aes(x = year, y = gdpPercap)) +
  geom_line())
```

```{r}
(usa_pop <- gapminder_usa %>%
  ggplot(aes(x = year, y = pop)) +
  geom_col())
```

Let's make multi-panel plots.  If you need to wrap around a line, make sure you don't start your line with the +, it won't work (this is true for all `ggplot2` syntax).
```{r, cache = TRUE}
(usa_lifeexp + usa_gdppercap) / usa_pop +
plot_annotation(title = "Some plots about the United States",
                  tag_levels = "A")
```

You can see how this would be really useful for publications!

<br>

----
## 9 - Animating

<p align="center"> 
<img src=gganimate_fireworks.png width="95%" alt="an illustration including fireworks showing an application og gganimate to make action plots, also including cute monsters">
</p>
Artwork by [Allison Horst](https://github.com/allisonhorst/stats-illustrations)


Since we have time-scale data here, we could also build an animation that would help us look at our data.  What if we wanted to look at how life expectancy (`lifeExp`) and population (`pop`) change over time?  We could animate over the variable `year`, and do this by using the function [`animate()`](https://gganimate.com/reference/animate.html), and set [`transition_states()`](https://gganimate.com/reference/transition_states.html) to the variable we are giffing over.  

Note, I have included `closest_state` in the subtitle so the viewer can see what is the year at any stage of the animation.

To be able to tell which dot belongs to which country, I added a [`geom_text_repel()`](https://www.rdocumentation.org/packages/ggrepel/versions/0.9.1/topics/geom_label_repel) statement, which labels each point but is smart enough to not let the labels overlap.

I have also set `pop` to be on a log10 scale.

If you want to increase the resolution of your gif, and set the code chunk to `cache = TRUE` if the chunk runs slowly, so that it doesn't re-run when knitting if nothing has been edited, you can do this in the curly brackets at the top of your chunk, like this:

`{r, cache = TRUE, dpi = 600}`
```{r, cache = TRUE}
# install.packages("transformr") 
# if you are having problems with gganimate you may need to install transformr

p <- ggplot(gapminder_americas, aes(x = lifeExp, y = pop, fill = country, label = country)) +
  geom_point(shape = 21, color = "black") +
  geom_text_repel() +
  scale_y_log10() +
  theme_classic() +
  theme(legend.position = 'none') +
  labs(title = "Population and Life Expectancy in the Americas",
       subtitle = 'Year: {closest_state}', 
       x = "Life Expectancy", 
       y = "Log10 Population") +
  transition_states(year) 

animate(p)
```


There are many different ways to transition your data in `gganimate` - and you can learn more about them [here](https://gganimate.com/reference/index.html).

### 10 - Saving my gif
Now I want to save my gif.  We can do that simply with the function [`anim_save()`](https://gganimate.com/reference/anim_save.html) which works a lot like `ggsave()`.  

```{r, eval = FALSE}
anim_save(filename = "YOUR FILE PATH HERE",
          animation = p)
```

----

# 10 - Breakout rooms

## Loading data and get set up
Load the `palmerpenguins` dataset, look at its structure, and view the beginning of the df.
```{r}
library(palmerpenguins)
str(penguins)
head(penguins)
```

<br>

----
### Main exercises

#### Exercise 1

<div class="puzzle">
<div>
Like we did in [Code Club 7](https://biodash.github.io/codeclub/08_pivoting/), convert the two columns about penguin bill dimensions `bill_length_mm` and `bill_depth_mm` to two columns called `bill_dimension` and `value`.  Drop your NAs also.  Save this as a new df called `penguins_long`.

<details>
  <summary>
  Hints (click here)
  </summary>


  Use a combination of `drop_na()` and `pivot_longer()`, and it's helpful if you also set `names_prefix` in your `pivot_longer()` statement but not totally necessary.
<br>
</details>

<br>

<details>
  <summary>
  Solutions (click here)
  </summary>

```{r}
penguins_long <- penguins %>%
  drop_na() %>%
  pivot_longer(cols = bill_length_mm:bill_depth_mm,
               names_to = "bill_dimension",
               values_to = "value_mm",
               names_prefix = "bill_")

head(penguins_long)
```

</details>
<br>

</div>
</div>


----

#### Exercise 2

<div class="puzzle">
<div>
Plot body mass (`body_mass_g`) as related to bill length and depth.

<details>
  <summary>
  Hints (click here)
  </summary>


  Faceting will be useful here.
<br>
</details>

<br>

<details>
  <summary>
  Solutions (click here)
  </summary>
  
```{r}
penguins_long %>%
  ggplot(aes(x = body_mass_g, y = value_mm)) +
  geom_point() +
  facet_wrap(vars(bill_dimension))
```

</details>
<br>

</div>
</div>


----

#### Exercise 3

<div class="puzzle">
<div>
Take your plot from Exercise 2 and make it prettier.  You can do things like change your axis labels, add title, change themes as you see fit.  Color your points by sex.

<details>
  <summary>
  Hints (click here)
  </summary>


  Pick a theme you like.  `theme_classic()` is a good place to start, and if you want to download the package `hrbrthemes`, I really like the `theme_ipsum_rc()`.  
<br>
</details>

<br>

<details>
  <summary>
  Solutions (click here)
  </summary>

I've included some code that let's you re-name the strip text, or the text that is above each of your facets.  You do this with the [`labeller()`](https://ggplot2.tidyverse.org/reference/labeller.html) function.
```{r}
library(hrbrthemes) # for pretty & easy themes

# formatting facet strip text labels
dim_mm <- c("Culman Bill Depth", "Culman Bill Length")
names(dim_mm) <- c("depth_mm", "length_mm")

# this is just one example
penguins_long %>%
  ggplot(aes(x = body_mass_g, y = value_mm, color = sex)) +
  geom_point() +
  theme_ipsum_rc() +
  theme(axis.title.x = element_text(hjust = 0.5),
        axis.title.y = element_text(hjust = 0.5),
        strip.text = element_text(hjust = 0.5)) +
  labs(x = "Body Mass (g)",
       y = "mm",
       title = "Bill length and depth vs. body mass in penguins",
       color = "Sex",
       caption = "Data from https://allisonhorst.github.io/palmerpenguins/") +
  facet_wrap(vars(bill_dimension),
             labeller = labeller(bill_dimension = dim_mm))
```

</details>
<br>

</div>
</div>


----

#### Exercise 4

<div class="puzzle">
<div>
Add a second dimension of faceting by species.  

<details>
  <summary>
  Hints (click here)
  </summary>


  You do this within your `facet_wrap()` call.  You might want to try the formula syntax, which works like this: `var1~var2`.
<br>
</details>

<br>

<details>
  <summary>
  Solutions (click here)
  </summary>
  
```{r}
penguins_long %>%
  ggplot(aes(x = body_mass_g, y = value_mm, color = sex)) +
  geom_point() +
  theme_ipsum_rc() +
  theme(axis.title.x = element_text(hjust = 0.5),
        axis.title.y = element_text(hjust = 0.5),
        strip.text = element_text(hjust = 0.5)) +
  labs(x = "Body Mass (g)",
       y = "mm",
       title = "Bill length and depth vs. body mass in penguins",
       color = "Sex",
       caption = "Data from https://allisonhorst.github.io/palmerpenguins/") +
  facet_wrap(bill_dimension~species,
             labeller = labeller(bill_dimension = dim_mm))
```

</details>
<br>

</div>
</div>


----

#### Exercise 5

<div class="puzzle">
<div>
Using your plot from Exercise 3, highlight the datapoints coming from Dream Island in purple.

<details>
  <summary>
  Hints (click here)
  </summary>


  You can use syntax inside `gghighlight()` just like you do in `filter()`.
<br>
</details>

<br>

<details>
  <summary>
  Solutions (click here)
  </summary>

```{r}
# what are our islands?
unique(penguins_long$island)

penguins_long %>%
  ggplot(aes(x = body_mass_g, y = value_mm)) +
  geom_point(color = "purple") +
  gghighlight(island == "Dream") +
  facet_wrap(vars(bill_dimension))
```

</details>
<br>

</div>
</div>


----

#### Exercise 6

<div class="puzzle">
<div>
Using your sample plot for Exercise 3, highlight penguins that have a `body_mass_g` less than 3500 g, in blue. 

<details>
  <summary>
  Hints (click here)
  </summary>


  You can use syntax inside `gghighlight()` just like you do in `filter()`, and you can also use these filter functions like `<`, `>`, `<=`, `!` and `AND` inside your call.
<br>
</details>

<br>

<details>
  <summary>
  Solutions (click here)
  </summary>

```{r}
# what are our islands?
unique(penguins_long$island)

penguins_long %>%
  ggplot(aes(x = body_mass_g, y = value_mm)) +
  geom_point(color = "blue") +
  gghighlight(island == "Dream") +
  facet_wrap(vars(bill_dimension))
```

</details>
<br>

</div>
</div>


----

### Bonus exercises
#### Bonus 1

<div class="puzzle">
<div>
Plot `flipper_length_mm` vs. `body_mass_g` and animate the plot to show only one `species` at a time.

<details>
  <summary>
  Hints (click here)
  </summary>


 Try animating over `species`, using `transition_states()` and set `{closest_state}` in your title or subtitle so you can tell what you're looking at.
<br>
</details>

<br>

<details>
  <summary>
  Solutions (click here)
  </summary>

```{r, cache = TRUE, message = FALSE, warning = FALSE, fig.width = 4, fig.height = 4}
flipper_by_BW <- penguins %>%
  ggplot(aes(x = body_mass_g, y = flipper_length_mm, fill = species)) +
  geom_point(shape = 21, color = "black") +
  theme_classic() +
  theme(legend.position = 'none') +
  labs(title = "Population and Life Expectancy in the Americas",
       subtitle = 'Penguin Species: {closest_state}', 
       x = "Body Mass (g)", 
       y = "Flipper Length (mm)") +
  transition_states(species) 

animate(flipper_by_BW)
```

</details>
<br>

</div>
</div>


----

#### Bonus 2

<div class="puzzle">
<div>
You have now made an excellent gif, so save it! 

<details>
  <summary>
  Hints (click here)
  </summary>

Use [`anim_save()`](https://gganimate.com/reference/anim_save.html) to save your animation, which works in a similar way to [`ggsave()`](https://ggplot2.tidyverse.org/reference/ggsave.html), which is another very useful function.
<br>
</details>

<br>

<details>
  <summary>
  Solutions (click here)
  </summary>

```{r, eval = FALSE}
anim_save(filename = "YOUR FILE PATH HERE",
          animation = flipper_by_BW)
```

</details>
<br>

</div>
</div>


----

#### Bonus 3

<div class="puzzle">
<div>
Let's practice making multi-panel plots.  Plot:

Boxplot of `body_mass_g` by `sex`

Histogram of number of observations per `island`

Distribution of `flipper_length_mm` by `species`.

Tag your plots so each has a lowercase letter associated with it.

<details>
  <summary>
  Hints (click here)
  </summary>
  
  Use the syntax from the package `patchwork`.  You can learn more [here](https://patchwork.data-imaginist.com/).  Also use [`plot_annotation()`](https://patchwork.data-imaginist.com/reference/plot_annotation.html).
<br>
</details>

<br>

<details>
  <summary>
  Solutions (click here)
  </summary>
  
* `title =` allows you to set a title
* `tag_levels =` allows you to determine how you want your panels to be tagged.

Boxplot of `body_mass_g` by `sex`.
```{r}
penguins_mass_by_sex <- penguins_long %>%
  ggplot(aes(x = sex, y = body_mass_g)) +
  geom_boxplot()

penguins_mass_by_sex
```

Histogram of number of observations per `island`.
```{r}
penguins_by_island <- penguins_long %>%
  ggplot(aes(y = island, fill = island)) +
  geom_histogram(stat = "count")

penguins_by_island
```

Distribution of `flipper_length_mm` by `species`.
```{r}
penguins_flipper_species <- penguins_long %>%
  ggplot(aes(x = flipper_length_mm, group = species, fill = species)) +
  geom_density(alpha = 0.5) +
  scale_fill_viridis_d()

penguins_flipper_species
```

```{r}
penguins_flipper_species / (penguins_mass_by_sex + penguins_by_island) +
  plot_annotation(title = "Looking at penguins...",
                  tag_levels = "a")
```



</details>
<br>

</div>
</div>


----

