---
output: hugodown::md_document
title: "Session 6: Factors"
subtitle: "Learn how to use them to your advantage!"
summary: "In this sixth session of Code Club, we'll learn how to use factors to our advantage"
authors: [stephen-opiyo]
tags: [codeclub, factors]
date: 2021-01-18
lastmod: 2021-01-18
toc: true

# Featured image
# To use, add an image named `featured.jpg/png` to your page's folder.
# Focal points: Smart, Center, TopLeft, Top, TopRight, Left, Right, BottomLeft, Bottom, BottomRight.
image:
  caption: "Artwork by @allison_horst"
  focal_point: ""
  preview_only: false

# Projects (optional).
#   Associate this post with one or more of your projects.
#   Simply enter your project's folder or file name without extension.
#   E.g. `projects = ["internal-project"]` references `content/project/deep-learning/index.md`.
#   Otherwise, set `projects = []`.
projects: []
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval=TRUE)
```

<br>
<br>

-----

Factors form the basis for many powerful operations in R,
including many performed on tabular data.
The motivation for factors comes from the notion of categorical variables.
These variables are non-numeric in nature corresponding to categories such as
male and female, or Democrat, Republican and Independent. 

**A factor might be viewed simply as a vector with a bit of more information added.**
The extra information consists of a record of distinct values in that vector,
which are called: ***levels***.

Let us look at some examples of factors. We will make use of the package *forcats*,
which is one of the 8 core *tidyverse* packages.
Therefore, we start by loading the *tidyverse*:

```{r include=TRUE}
library(tidyverse)

## Check whether "forcats" is listed among the loaded packages.
## Alternatively, you could load "forcats" (and "ggplot2") separately:
# install.packages("forcats")
# library(forcats)
# library(ggplot2)
```

<br>

----

## Example 1: From a numeric vector to a factor

Let us create a factor `xf` from a vector `x` with the numbers `5`, `12`, `13`,
and `12`:

```{r include=TRUE}
x <- c(5,12, 13,12)
x

# Convert the vector to a factor:
xf <- factor(x)
xf
```

The *distinct values* in `xf` are `5`, `12` and `13`,
and are listed as *levels*.

Let us look in a bit more details at our factor using the R functions
`str` and `unclass`:

```{r include=TRUE}
str(xf)
unclass(xf)
```

Notice that the values in the factor are not stored as `(5, 12, 13, 12)`,
but rather as `(1, 2, 3, 2)`!

This means that our data consists first of a level-1 value,
then level-2 and level 3 values, and finally another level-2 value.
So, the data has been recorded by level.

The values attached to each level are recorded too,
but as *characters* such as `"5"` rather than as numbers such as `5`.

<br>

----

## Example 2: From a character vector to a factor

We will use the levels Democrat (`D`), Republican (`R`), and Independent (`I`).
First, we save a *vector*:

```{r include=TRUE}
y <- c("D", "R", "R", "I", "R", "I", "D", "I")
y
str(y)
```

Then, we again convert the *vector* to a *factor*, and look at the levels:

```{r inclue=TRUE}
fy <- factor(y)
fy
unclass(fy)
```

<br>

----

## Example 3: Ordering factor levels

Some variables can be challenging to sort automatically,
because the desired sorting order is not alphabetical or numeric.

For instance, months that are listed using characters:

```{r include=TRUE}
months_vector <- c("Dec", "Apr", "Jan", "Mar")

# Try to sort using the `sort` function
sort(months_vector)
```

That didn't sort in a useful way.
**But, the problem can be fixed by using a factor.**

First, we create a list of the valid levels, which are all 12 months in a year:

```{r include=TRUE}
month_levels <- c("Jan", "Feb", "Mar", "Apr", "May", "Jun",
                  "Jul", "Aug", "Sep", "Oct", "Nov", "Dec")
```

Then we convert the vector into a factor, like before,
but now we additionally specify the desired levels of the factor, **in order**,
using the `levels` argument:

```{r include=TRUE}
months_factor <- factor(months_vector, levels = month_levels)
```

Now it sorts the way we want to!

```{r include=TRUE}
sort(months_factor)
```

<br>

----

## Example 4: Use of factors in plots with *forcats*

### 4A: Plot after reordering manually with `fct_relevel()`

We will use the *mtcars* data,
which was extracted from the 1974 Motor Trend US magazine,
and comprises fuel consumption and 10 aspects of automobile design and
performance for 32 automobiles (1973–74 models) --
a data frame with 32 observations for  11 (numeric) variables,

```{r include=TRUE}
data(mtcars)

names(mtcars)
dim(mtcars)
str(mtcars)
```

we will select six variables (`mpg`, `cyl`, `disp`, `hp`, and `wt`)
to create a dataset `Data`.

- `mpg`: Miles per (US) gallon,
- `cyl`: Number of cylinders
- `disp`:	Displacement (cu.in.)
- `hp`: Horse power
- `wt`: Weight (in 1000 lbs)

```{r include=TRUE}
Data <- mtcars %>%
  select("mpg", "cyl", "disp", "hp", "wt")
```

Now, we'll add a new column `cyl_chr` by converting `cyl`
from *numeric* to *character*:

```{r include=TRUE}
Data <- Data %>%
  mutate(cyl_chr = recode(cyl,`4` = "Four", `6` = "Six", `8` = "Eight"))

head(Data)
```

We plot a bar chart for cyl_1:
```{r include=TRUE}
Data %>%
  ggplot(aes(x = cyl_chr)) +
  geom_bar() 
```

In the plot, the levels of the factor were arranged in alphabetical order
(`Eight`, `Four`, and `Six`).

Instead, we want the bar graph arranged in the order `Four`, `Six`, and `Eight`.

An alternative to using `factor(levels = ...)` like we did above,
is to use the `fct_relevel()` function from the *forcats* package:

```{r include=TRUE}
Data %>%
  mutate(cyl_chr = fct_relevel(cyl_chr, "Four", "Six", "Eight")) %>%
  ggplot(aes(x = cyl_chr)) +
  geom_bar() +
  labs(x = "Cylinder", y = "Miles/per gallon")
```

<br>

### 4B: Plot after reordering by the value of another column (`fct_reorder`)

Create a dataset called `Data_a`:

```{r include=TRUE}
Data_a <- data.frame(name = c("North", "South", "East", "West"),
                     var = sample(seq(1, 10), 4))
```

Plot a bar chart of `Data_a`:

```{r include=TRUE}
Data_a %>%
  ggplot(aes(x = name, y = var)) +
  geom_bar(stat = "identity", fill = "#f68034", alpha = 0.6, width = 0.4) 
```

Reorder following the value of another column using the `fct_reorder()` function,
and flip the plot:

```{r include=TRUE}
Data_a %>%
  mutate(name = fct_reorder(name, var)) %>%
  ggplot(aes(x = name, y = var)) +
  geom_bar(stat = "identity", fill = "#f68034", alpha = 0.6, width = 0.4) +
  coord_flip() 
```

<br>

There are several more convenient reordering functions in the *forcats* package,
including:

- `fact_infreq()` to reorder by occurrence frequencies of each level
  (see the picture at the top of the post).

- `fct_inorder()` to reorder by order of appearance in the dataframe.
   This can be useful, for example,
   if your dataframe has already been sorted properly,
   and you just need to prevent automatic alphabetic reordering when plotting. 

<br>

----

## Breakout rooms!

For the Breakout room exercises,
we will use datasets from *mtcars* and
the `gss_cat` dataset from the *forcats* package.

### Exercise 1

<div class="puzzle">
<div>
Convert the variable `gear` from *mtcars* to a character vector with words
for each number (link in example 4A), and plot a bar chart.

Then, use a factor to reorder the bars to appear in the regular "numeric" order:
"Three" then "Four" then "Five".

<details>
<summary>
Hints (click here)
</summary>
<p>
- Use the `mutate()`, `recode()`, and `fct_relevel()` functions.

- Add a column called `gear_chr` by converting gear from *numeric* to *character*.
</p>
</details>

<br>

<details>
<summary>
Solutions (click here)
</summary>
<br>

- Start by loading the dataset:

```{r echo=TRUE}
data("mtcars")
```

<br>

- Now, create a new dataset `Gear` from *mtcars*, adding a column `gear_chr`:

```{r echo=TRUE}
gear_df <- mtcars %>%
  mutate(gear_chr = recode(gear,
                         `3`= "Three",
                         `4` ="Four",
                         `5`= "Five"))

head(gear_df)
```

<br>

- Finally, use the *forcats* function `fct_relevel()` to rearrange `gear_chr`
  in nonalphabetical order, and plot the barchart using `geom_bar()`:

```{r echo=TRUE}
gear_df %>%
  mutate(gear_fct = fct_relevel(gear_chr, "Three", "Four", "Five")) %>%
  ggplot(aes(x = gear_fct)) +
  geom_bar() +
  labs(x = "Gear", y = "Miles/per gallon")
```

<br>
</details>

</div>
</div>

<br>

----

### Exercise 2

<div class="puzzle">
<div>

Using the `gss_cat` dataset from the *forcats* package (available as `gsscat`
in your environment),
create a plot that compares the average number of hours spent watching TV
per day across religions, and where *religions are ordered by the average 
number of hours*.

(Despite what we've learned last week, start by merely plotting the mean,
and no distributions, using a barplot or with `geom_point()`.)

*Source: (R for Data Science)*

<details>
<summary>
Hints (click here)
</summary>
<br>
In order to be able to order the factor by the average number of hours
spent watching TV, first compute this average per religion, and save 
the results in a dataframe (use `mutate()` and `summarize()`).

Then, use `fct_recorder()` to reorder the factor.

</details>

<br>

<details>
<summary>
Solutions (click here)
</summary>
<br>

First, have a look at the dataset:

```{r echo=TRUE}
forcats::gss_cat
```

Then, calculate the mean number of tv-hours and create a plot:

```{r echo=TRUE}
relig <- gss_cat %>%
	group_by(relig) %>%
	summarize(tvhours = mean(tvhours, na.rm = TRUE))

ggplot(relig, aes(tvhours, relig)) +
  geom_point()
```

It is difficult to interpret this plot because there is no overall pattern. 

We can improve the plot by reordering the level of religion
using `fct_reorder()`:

```{r echo=TRUE}
relig <- gss_cat %>%
	group_by(relig) %>%
	summarize(tvhours = mean(tvhours, na.rm = TRUE))

relig %>%
  mutate(relig = fct_reorder(relig, tvhours)) %>%
  ggplot(aes(tvhours, relig)) +
  geom_point()
```

Reordering religion makes it much easier to see that people in the
"Don't know" category watch much more TV.

<br>
</details>

<br>

</div>
</div>

----

### Bonus: Exercise 3

<div class="puzzle">
<div>

In exercise 2, we saw large differences in the *average* time spent watching TV
across religions, but we should perhaps have a closer look at the data
by plotting distributions.

Go back to the [previous Code Club session](https://biodash.github.io/codeclub/05_ggplot-round-2/)
and decide which type of plot could be ideal with so many categories.

<details>
<summary>
Hints (click here)
</summary>
<br>
`geom_density_ridges()` from the *ggridges* package is very well suited
for a plot with so many categories.
</details>
<br>
<details>
<summary>
Solutions (click here)
</summary>
<br>

```{r echo=TRUE}
library(ggridges)

ggplot(gss_cat, aes(x = tvhours, y = relig, fill = relig)) +
  geom_density_ridges(alpha = 0.8) +
  labs(x = 'Number of hours spent watching TV', y = 'Religion') +
  guides(fill = FALSE) +
  theme_minimal()
```

</details>

</div>
</div>

<br>
