---
output: hugodown::md_document
title: "S04E01: R for Data Science - Chapter 1"
subtitle: "Introducing a new season of Code Club, in which we will read the book R for Data Science (R4DS)"
summary: "We will introduce a new season of Code Club, in which we'll do things a little differently than before: we are going to work our way through a book: R for Data Science. Today, we'll look at the first, introductory chapter of the book."
authors: [admin]
tags: [codeclub, r4ds]
date: "`r Sys.Date()`"
lastmod: "`r Sys.Date()`"
toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, eval=TRUE, message=FALSE, cache=FALSE)
knitr::opts_knit$set(root.dir = here::here("content/codeclub/S04E01_r4ds-ch1"))

library(tidyverse)
```

<p align="center">
<img src=img/r4ds_cover.png width="45%">
</p>

------------------------------------------------------------------------

## I -- Intro to this season of Code Club

### Organizers

- _Michael Broe_ -- Evolution, Ecology and Organismal Biology (EEOB)
- _Jessica Cooperstone_ -- Horticulture & Crop Science (HCS) / Food Science & Technology (FST)
- _Stephen Opiyo_ -- Molecular & Cellular Imaging Center (MCIC) - Columbus
- _Jelmer Poelstra_ -- Molecular & Cellular Imaging Center (MCIC) - Wooster
- _Mike Sovic_ -- Center for Applied Plant Sciences (CAPS)

### Code Club practicalities

- In-person (Columbus & Wooster) and Zoom hybrid

- Mix of instruction/discussion with the entire group,
  and doing exercises in breakout groups of up to ~4 people.

- When doing **exercises in breakout groups**, we encourage you:
  - To briefly introduce yourselves and to do the exercises as a group
  - On Zoom, to turn your cameras on and to have someone share their screen
    (use the `Ask for help` button in Zoom to get help from an organizer)
  - To let a _less_ experienced person do the screen sharing and coding
  
- You can **ask a question** at any time, by speaking or typing in the Zoom chat.

- You can generally come early or stay late for
  **troubleshooting** but also for questions related to your research.

More general notes:

- We recommend that you **read** the relevant (part of the) chapter before each session,
  especially if the material in the chapter is new to you.

- We try to make each session **as stand-alone as possible**.
  Still, if you missed one or more sessions, you would ideally catch up on reading
  those parts of the book, especially when we split a chapter across multiple sessions. 
  
- We **record** the whole-group parts of the Zoom call,
  and share the recordings only with Code Club participants.
  
- We're always hoping for someone outside of the group of organizers to
  **lead a session** --
  this might be more feasible now that we're going through a book?

### New to Code Club or R?

Take a look at these pages on our website:

- [Computer setup for Code Club](/codeclub-setup/)
- [Resources and tips to get started with R](/tutorials/r-resources-tips/)
- [List of all previous Code Club session topics](/codeclub-schedule/#previous-semesters)

<br>

------------------------------------------------------------------------

## II -- The R for Data Science book (R4DS)

This excellent book by Hadley Wickham
(also author of many of the R packages used in the book!)
and Garret Grolemund, has a freely available
**[online version](https://r4ds.had.co.nz/) that is regularly updated**
and contains exercises. It was originally published in 2016.

The book focuses on the so-called **"tidyverse" ecosystem** in R.
The _tidyverse_ can be seen as a **modern dialect of R**.
Most of its functionality is also contained in "base R" (what comes shipped with R by default),
but it has an improved and more consistent programming interface ("syntax").
In previous Code Clubs,
we have often --but not always!-- been doing things "the tidyverse way"
as well.

The book doesn't technically assume any **previous experience with R**,
but if you're completely new to R and to coding in any language,
we would recommend you take a look at some introductory R material
([see this page for some resources](/tutorials/r-resources-tips/))
before we start with Chapter 2 next week.

We will _not_ be able to finish the book by the end of the summer.
But if folks are liking the book, we may carry on with it during the fall semester.

<br>

------------------------------------------------------------------------

## III -- R4DS Chapter 1 notes

### 1.1 - What you will learn

The data science process visualized:

<p align="center">
<img src=img/data-science.png width="80%">
</p>

#### What is tidy data?

<details>
<summary>
<b>Brief explanation and examples</b> (Click here)
</summary>
<br>

First, you can think of this along the lines of the colloquial meaning of the word:
the data is well-organized.

Additionally, computer-readability should be prioritized over human-readability
(think: no color-coded cells, multiple header columns, or merged cells).

But, most of all, "tidy" in the context of the tidyverse refers to the following,
as it is phrased in the book:

> In brief, when your data is tidy, each column is a variable, and each row is an observation.

But what does this mean?
In practice, it means having your data not in a "wide format"
(all the information about each sample/individual in one row)
but in a "long format"
(variables not spread across multiple columns) -- see the examples below.

Example 1, not tidy:

```{r, echo=FALSE}
classroom <- data.frame(name = c("Billy", "Suzy", "Lionel"),
                        quiz1 = c("C", "F", "B"),
                        quiz2 = c("D", "A", "C"),
                        test1 = c("C", "A", "B"))
classroom
```

Example 1, tidy:

```{r, echo=FALSE}
classroom %>% 
  pivot_longer(quiz1:test1, names_to = "assessment", values_to = "grade") %>% 
  arrange(name, assessment)
```

Example 2, not tidy (in _matrix_ form):

```{r, echo=FALSE}
m <- matrix(rpois(n = 25, lambda = 50), nrow = 5,
            dimnames = list(paste0("sample", 1:5), paste0("gene", 1:5)))
m
```

Example 2, tidy:

```{r, echo=FALSE}
m %>%
  as.data.frame() %>%
  rownames_to_column("sample") %>%
  pivot_longer(cols = -sample, names_to = "gene", values_to = "count")
```

For more, the book has a separate chapter on [tidy data](https://r4ds.had.co.nz/tidy-data.html),
and there is also [_tidyr_ vignette on tidy data](https://cran.r-project.org/web/packages/tidyr/vignettes/tidy-data.html).

</details>

<br>

### 1.3 - What you won't learn

Some perhaps unfamiliar terms from this section:

#### [1.3.1 - Processing big data](https://r4ds.had.co.nz/introduction.html#big-data)

> Fortunately each problem is independent of the others (a setup that is sometimes called embarrassingly parallel),
> so you just need a system (like Hadoop or Spark) that allows you to send different datasets
> to different computers for processing. 

At OSU, and most other universities, we instead tend to make use of "**supercomputers**"
when we want to simultaneously run an analysis many times,
and more broadly, if we have "big data".
Specifically, we have the ["Ohio Supercomputer Center" (OSC)](https://www.osc.edu/) here.

#### [1.3.3 - Non-rectangular data](https://r4ds.had.co.nz/introduction.html#non-rectangular-data)

Rectangular data is basically data that can be effectively entered in a spreadsheet
(and in R, we tend to put this in a so-called "dataframe" or "tibble").
**The _tidyverse_ is highly dataframe-oriented**,
so it makes sense that the book focuses on rectangular data.

<br>

### 1.4.2 - RStudio interface

R itself simply provides a "*console*" (command-line interface) where you can type your commands.
RStudio, on the other hand, allows you to see the R console side-by-side with your scripts, plots, and more.

Once you have a running instance of RStudio,
**create a new R script** by clicking `File` > `New File` > `R Script`.
Now, you should see all 4 "panes" that the RStudio window is divided into:

- _Top-left_: The **Editor** for your scripts and other documents (_hidden when no file is open_).
- _Bottom-left_: The **R Console** to interactively run your code (+ other tabs).
- _Top-right_: Your **Environment** with R objects you have created (+ other tabs).
- _Bottom-right_: Tabs for **Files**, **Plots**, **Help**, and others.

<p align="center">
<img src=img/rstudio-layout-ed.png width="95%">
</p>

<br>

<div class="puzzle">
<div>

### Check that you have R and RStudio working

Take a moment to _**explore the RStudio interface**_.
Were you able to open a new file to get the Editor pane?
Can you run some simple R code in the console
(e.g., [the example in the book](https://r4ds.had.co.nz/introduction.html#running-r-code))?

Take a look at your _**version of R**_:
this was printed in the console when you started RStudio
(see the RStudio screenshot above).

The current version of R is `4.2.0`.
If your version of R is below `4.0`, it will be a good idea to update R.
To do so, you can follow
[these instructions](/tutorials/r-resources-tips/#updating-r).
But it is better to start this process at the very end of this session
or after it, since it may take a while.
  
</div>
</div>

<br>

### 1.4.3 & 1.4.4 - R packages

> An R package is a collection of functions, data, and documentation that extends the capabilities of base R.

So, you can think of packages as "add-ons" / "extensions" to base R.
(With _base R_ meaning all the default functionality of R, or
"what R comes shipped with".)

#### Installation versus loading

Packages have to be separately **installed** (usually _from within R_, using R code)
and once you have done this, you don't need to redo it unless:

- You want a different version of the package
- You have switched to a different version of R

Unlike installation, **loading** a package is necessary again and again,
in every R session that you want to use it.

#### The tidyverse

The _tidyverse_ is unusual in that it is a _collection_ of packages
that can still be installed and loaded with a single command.
The individual _tidyverse_ packages are the focus of several chapters in the book,
for instance:

| Package     | Functionality   | Main chapter
|-------------|-----------------|-------------|
| _ggplot2_   | Creating plots                | Ch. 3
| _tidyr_ & _dplyr_ | Manipulating dataframes | Ch. 5 & 7
| _readr_     | Reading in data               | Ch. 11
| _stringr_   | Working with "strings" (text)   | Ch. 14
| _forcats_   | Working with "factors"        | Ch. 15
| _purrr_     | Iteration                     | Ch. 21

#### Data packages

Additionally, the book uses a couple of "data packages"
(packages that only contain data, and no functions):
_nycflights13_, _gapminder_, and _Lahman_.

<br>

----

## Breakout rooms

<div class="puzzle">
<div>

### 1. Introduce yourselves!

Please take a moment to introduce yourself to your breakout roommates.
You may also want to mention:

- Your level of experience with R (and perhaps other coding languages)

- What you want to use R for, or what you are already using R for

- Why you think this book might be useful, if you have an idea already

</div>
</div>

<div class="puzzle">
<div>

### 2. Install/load the packages

_Usually, exercises can be done with your breakout group on one computer,
but the following should be done individually, to check that everyone has R up
and running._

Most of you should already have the **_tidyverse_** installed,
so let's start by trying to _load_ it. 
This is done with the `library()` function.
To check if you can load the _tidyverse_,
run the following and see if you get similar output as printed below:

```{r, message=TRUE, warning=TRUE}
library(tidyverse)
```

<p align="center">
<img src=img/load_tidyverse.png width="90%">
</p>

If instead, you got something like:

```{r}
#> Error in library(tidyverse) : there is no package called ‘tidyverse’
```

...that means you still need to install it:

```{r, eval = FALSE}
## Note: the package name is "quoted" in the install.packages() function:
install.packages("tidyverse")

## ... but it is not (normally) in library():
library(tidyverse)
```

<br>

Now, let's also install the **data packages**:

```{r, eval = FALSE}
install.packages(c("nycflights13", "gapminder", "Lahman"))
```

The previous installation commands should return a whole bunch of output,
but if all went well, you should not see any errors.
Instead, look for phrases like `* DONE (nycflights13)`,
which indicate successful installation of a package.

You can also _load_ the data packages:

```{r, eval = FALSE}
library(nycflights13)
library(gapminder)
library(Lahman)
```

You won't see any output when loading most packages,
like the three above (but unlike the _tidyverse_).

**Bonus question:**
What are these "conflicts" in the _tidyverse_ startup messages referring to?

</div>
</div>

<div class="puzzle">
<div>

### 3. Bonus: Questions or remarks about the chapter?

Discuss or ask about whatever you thought was interesting/confusing/etc about
the chapter!

If nothing else comes up, you could think about and discuss the following:

- [1.3.2](https://r4ds.had.co.nz/introduction.html#python-julia-and-friends):
  _**"Data science teams"**_ -- Are grad students in a lab "Data science teams",
  or are they talking about something else?
  What does this, perhaps, say about the expected primary audience for the book?

- [1.3.4](https://r4ds.had.co.nz/introduction.html#hypothesis-confirmation):
  "**Hypothesis generation**" vs. "**hypothesis confirmation**" --
  are you familiar with this distinction and do you used it in practice?

- [1.3.2](https://r4ds.had.co.nz/introduction.html#python-julia-and-friends):
  Other languages commonly used for data analysis: **Python** and **Julia**.
  Are you familiar at all with these languages and their pros and cons
  with respect to R?
  Or, more simply: why are you trying to learn R and not those languages?
  
</div>
</div>

<br> <br>
