(despite some
disadvantages\^[As it is in some ways a dialect of R, the \*tidyverse\* can
cause confusion ("tidy evaluation" in particular) and can sometimes make it seem like
there is just more to learn -- because base R can't be fully ignored.
Its rapid development has also meant that in some cases,
new functions and approaches are being retired/soft-deprecated just
a few years later.])




<div class="alert alert-note">
<div>

### Types of data frames

When you use base R functions to read in data, you will get a `data.frame` object.
However, when you use *readr*, you will get the *tidyverse*'s **subclass** of `data.frame`:
colloquially called a `tibble` and formally a `tbl_df`.

`tibbles` and "regular" `data.frames` can generally be used interchangeably,
but some default behavior does differ.
Most strikingly, they are printed to screen differently.

To tell what kind of data frame your are dealing with,
and to tell the class of any R object,
use the `class()` function, e.g. `class(birds)`.

</div>
</div>

:::puzzle

### Exercise 2 (optional/homework)

**Comparing _tibbles_ and regular _data.frames_**

Read in the bird data as a *data.frame* and compare overviews of the
*tibble* and the *data.frame*.

- What are some differences in how they are displayed and behave?
  In particular, compare what happens when you use `print()`
  (or simply type the object name).

- Were all the columns parsed in the same way -- i.e., did the *tibble* and the
  *data.frame* assign the same class to each column?

- How could you switch (convert) directly between *tibble* and *data.frame* objects? 

<details>
  <summary>
  Hints (click here)
  </summary>
<br>
<ul>
<li>Use the <code>read.table()</code> function,
    and note that you will need to specify non-default options for the arguments <code>sep</code> and      <code>header</code>. Type <code>?read.table</code> for details.</li>
<li>Have a close look at the class for the "eventDate" column.
    Note that <code>POSIXct</code> is the formal date-time class,
    while tibbles/tidyverse will refer to these as <code>ddtm</code>.</li>
<li>Use the <code>as.data.frame()</code> and <code>as_tibble()</code> functions to change the object type.
</li>
</ul>

</details>

<br>

<details>
  <summary>
  Solutions (click here)
  </summary>
<br>

```{r}
# Read the dataset as data.frame using read.table():
birds_df <- read.table(infile, sep = '\t', header = TRUE)

# Class of the eventDate column -- read.table() did not parse the date as a date!
class(birds$eventDate)
# [1] "POSIXct" "POSIXt" 
class(birds_df$eventDate)
# [1] "character"

# Convert between a data.frame and a tibble:
birds_df <- as.data.frame(birds)
birds_tibble <- as_tibble(birds_df)
```

</details>

:::



<br>

<div class="alert alert-note">
<div>

### Pro-tip for data frame summaries

Use the `skim()` function from the *skimr* package:

```{r}
install.packages(skimr)
skim(birds)
```

</div>
</div>
