---
title: "S03E05: Introduction to correlation"
subtitle: "Using the _corrr_ package"
summary: "...."
authors: [stephen-opiyo]
date: "`r Sys.Date()`"
output: hugodown::md_document
toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

-----

## Housekeeping

#### New to Code Club?

Check out the [Code Club Computer Setup](/codeclub-setup/) instructions,
which also has pointers for if you're new to R or RStudio.
A few related Code Club sessions include:

- [S03E01](/codeclub/s03e01_ttests/): T-tests
- [S03E02](/codeclub/s03e02_anova/): ANOVA
- [S03E03](/codeclub/s03e03_pca/): PCA
- [S03E04](/codeclub/s03e04_anova2/): ANOVA part II

#### What will we go over today

-   We will introduce correlation.
-   Learn using the [`corrr()`](https://cran.r-project.org/web/packages/corrr/index.html)
    package.

#### R packages we will use

- _corrr_ -- for correlation analysis
- _palmerpenguins_ -- for the penguins data
- _tidyverse_ -- for data wrangling

<br>

----

## 1 - What is correlation?

Correlation is a statistical method used to assess a linear association between two continuous variables.
It is measured by a statistic called the correlation coefficient,
which represents the strength of the linear association between the variables in question.
It is a dimensionless quantity that takes a value in the range −1 to +1.
A correlation coefficient of zero indicates that no linear relationship exists
between two continuous variables,
and a correlation coefficient of −1 or +1 indicates a perfect linear relationship. 
(The stronger the correlation, the closer the correlation coefficient comes to −1 and +1.)

**Positive coefficient**: If the coefficient is a positive number,
the variables are positively related (i.e., as the value of one variable goes up,
the value of the other also tends to do so). 
 
**Negative coefficient**: If the coefficient is a negative number,
the variables are inversely related (i.e., as the value of one variable goes up,
the value of the other tends to go down).  

**Types of correlation coefficients**:
There are two main types of correlation coefficients,
_Pearson's correlation coefficient_ and _Spearman's rank correlation coefficient_.
The correct usage of correlation coefficient type depends on the types of variables being studied.   

- *Pearson's correlation coefficient*:
  Pearson's correlation coefficient is denoted as **ϱ** for a population parameter
  and as **r** for a sample statistic.
  It is used when both variables being studied are normally distributed.

- *Spearman's rank correlation coefficient*:
  Spearman's rank correlation coefficient is denoted as **ϱs** for a population
  parameter and as **rs** for a sample statistic.
  It is appropriate when one or both variables are skewed or ordinal.

**Rule of thumb for interpreting the size of a correlation coefficient**

- .90 to 1.00 (−.90 to −1.00) -- Very high positive (negative) correlation
- .70 to .90 (−.70 to −.90) -- High positive (negative) correlation
- .50 to .70 (−.50 to −.70) -- Moderate positive (negative) correlation
- .30 to .50 (−.30 to −.50) -- Low positive (negative) correlation
- .00 to .30 (.00 to −.30) -- Negligible correlation

<br>

------

## 2 - The _corrr_ package and our data

The _corrr_ package is a tool for exploring correlations.
It makes it easy to perform routine tasks when exploring correlation matrices
such as ignoring the diagonal,
focusing on the correlations of certain variables against others,
or rearranging and visualizing the matrix in terms of the strength of the correlations.
The _corrr_ package exists within the Comprehensive R Archive Network,
or [CRAN](https://cran.r-project.org/).

Let's install it -- we only need to do this once:

```{r, eval = FALSE}
install.packages("corrr")
```

To use the _corrr_ package, we need to load it up using `library()`.
We also need to load the _tidyverse_ since we will be using it later:

```{r}
library(corrr)
library(tidyverse)
```

-----

**Let's get set up and grab some data to work with.**

We will use the same dataset [`palmerpenguins`](https://allisonhorst.github.io/palmerpenguins/)
used in the previous weeks.

If you didn't install this package previously, please do so now:

```{r, eval = FALSE}
install.packages("palmerpenguins")
```

Then, to use the package, we need to use the function `library()` to load it:

```{r}
library(palmerpenguins)
```

The data we will use today is a dataframe called `penguins`,
which we reference after loading the package.
We will look at the structure of the data:

```{r}
# look at the first 10 rows, all columns
head(penguins, 10)

# check the structure of penguins_data
# glimpse() which is a part of dplyr functions 
# similarly to str() and can be used interchangeably
glimpse(penguins)
```

Okay, now we have a sense of what the `penguins` dataset is.

<br>

------

## 2 - Basic usage of the _corrr_ package

Now we want to use the _corrr_ package to correlate variables in the `penguins`
dataframe.
The _corrr_ package uses the `correlate()` function and returns correlation
results in a tibble output.

Now let's select variables from the `penguins` dataframe using the pipe operator
`%>%` and the `select()` function.
We then specify the name of the variables that we want to select.
In this example, we are selecting variables `bill_length_mm`, `bill_depth_mm`,
`flipper_length_mm`, and `body_mass_g`.

Then, we will compute correlations among the variables:

```{r}
penguins_cor <- penguins %>%   # Take penguins_data
  select(bill_length_mm, bill_depth_mm, flipper_length_mm, body_mass_g) %>%
  correlate()       # Select variables and calculate their correlations

penguins_cor        # Correlation results in tibble
```

The output of the `correlate()` function can be piped (`%>%`) to:

- The `shave()` and `rearrange()` functions for internal changes
- The `focus()`, and `stretch()` functions to reshape the structure
- And the `rplot()`, `fashion()`, and `network_plot()` for visualizations.

*Let us start with internal changes*.

- The `shave()` function removes values of the upper or lower triangle and sets them to `NA`.
- The `rearrange()` function arranges the columns and rows based on correlation strengths.

We can apply the `shave()` function to remove the top triangle:

```{r}
penguins_cor %>%   # Take penguins_cor results
  shave()          # Remove the upper triangle
```

You can see that the values in the upper triangle have been replaced by `NA`s. 
Now let us remove `NA`s by using the `fashion()` function:

```{r}
penguins_cor %>%   # Take penguins_cor results
  shave %>%        # Remove the upper triangle
  fashion()        # Remove NAs
```

You can see that the `NA`s have been removed, and we have a clean tibble (dataframe).

Now, let's rearrange columns based on correlation strengths:

```{r}
penguins_cor %>%   # Take penguins_cor results
  rearrange %>%    # Rearrange based on correlation strengths
  shave %>%        # Remove the upper triangle
  fashion()        # Remove NAs
```

You can see that the values are arranged based on correlation strengths.

<br>

-----

<div class="puzzle"><div>

## Breakout session 1

- From the `penguins` dataframe,
  create a new dataset called `penguins_biscoe` by selecting only the penguins
  from the island of Biscoe.  

- In the `penguins_biscoe` dataframe,
  what are the correlations between `bill_length_mm`, `bill_depth_mm`,
  `flipper_length_mm`, and `year`? 

- Remove the upper triangle and `NA`s from the results,
  and arrange based on correlation strengths.

<details><summary><b>Hints</b> (click here)</summary>
<br>

Use the `filter()` function to only select penguins from the island of Biscoe.

</details>

<details><summary><b>Solution</b> (click here)</summary>
<br>

Select the penguins from the island of Biscoe:

```{r, eval = TRUE}
penguins_biscoe <- penguins %>%  # Save results in new object
  filter(island == "Biscoe")     # Select data from only Biscoe island  
```

Calculate correlation of the variables in `penguins_biscoe`:

```{r, eval = TRUE}
biscoe_cor <- penguins_biscoe %>%
  # select bill_length_mm, bill_depth_mm, flipper_length_mm, and year
  select(bill_length_mm, bill_depth_mm, flipper_length_mm, year) %>%
  correlate()        # Calculate correlation

biscoe_cor
```

To remove the upper triangle and rearrange the results:

```{r, eval = TRUE}
biscoe_cor  %>%      # Take biscoe_cor results
  rearrange() %>%    # Rearrange based on correlation strengths
  shave() %>%        # Remove the upper triangle
  fashion()          # Remove NAs
  
```

</details>

</div></div>

<br>

----

## 3 - Reshaping and visualizations

The output of the `correlate()` function can also be piped to:

- The `focus()` and `stretch()` functions to reshape the structure
- And the `rplot()`, and `network_plot()` for visualizations.

*Reshape structure*:

- The `focus()` function select columns or rows based on the variable specified.
- The `stretch()` converts correction results from a tibble into a long format. 

*Visualizations*:

- The `rplot()` function plots correlation results
- The `network_plot()` function plots a point for each variable,
  joined by paths for correlations

Let use the `stretch()` function to convert our correlation results
`penguins_cor` into a long format:

```{r}
penguins_cor %>%
  stretch()
```

You can see that the results are coverted into a long format. 

We can also select a column we are interested in.
Let us select only correlation between "bill_depth_mm" and the rest of the variables.

```{r}
penguins_cor %>%
  focus("bill_depth_mm")
```

You can see that we selected only the correlations between `bill_depth_mm`
and other variables.

We can now visualize the results:

```{r}
penguins_cor %>% 
  rearrange() %>%   # Rearrange based on correlation strengths
  rplot()           # Plot
```

We plotted based on correlation strengths.

<br>

-----

<div class="puzzle"><div>

## Breakout session 2 

Use the dataframe (`penguins_biscoe`) you created in Breakout session 1 to:

- Calculate correlations among the variables,
  and present the results in a long format

- Select the correlations that include `year`

- Plot a correlation graph based on correlation strengths

<details><summary><b>Solution</b> (click here)</summary>
<br>

```{r}
biscoe_cor <- penguins_biscoe %>%
  select(bill_length_mm, bill_depth_mm, flipper_length_mm, year) %>%
  correlate()

biscoe_cor %>%
  stretch()
```

You can see that results are in a long format.

```{r}
biscoe_cor %>%
   focus("year")
```

Correlations with `year`:

```{r}
biscoe_cor %>%
  rearrange() %>%
  rplot()
```

</details>

</div></div>

<br>

