---
title: "Session 16: lubridate"
subtitle: "Working with dates and times in R."
summary: "Today, we will learn how to work effectively with dates and times in R using the lubridate package."  
authors: [stephen-opiyo]
date: "2021-03-31"
output: hugodown::md_document
toc: true
image:
  caption: "Artwork by @allison_horst"
  focal_point: ""
  preview_only: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

<br> <br> <br>

-----

## New to Code Club?

- If you didn't already do this,
  please follow the [Code Club Computer Setup](/codeclub-setup/04_ggplot2/) instructions,
  which also has pointers for if you're new to R or RStudio.

- If you're able to do so, please open RStudio a bit before Code Club starts --
  and in case you run into issues,
  please join the Zoom call early and we'll help you troubleshoot.

<br>

-----

## 1. Getting set up

R has a range of functions that allow you to work with dates and times.
However, today we will discuss how to work with dates and times in R using
the package “*lubridate*”.

While *lubridate* is *tidyverse*-style, it is not part of the core *tidyverse*,
so we need to install it.

We are also going to use the bird data that was first discussed in Code Club session 1,
and we will need to download that.

```{r, include=TRUE}
# If needed, install the lubridate package:
# install.packages("lubridate")

# Load the tidyverse and lubridate:
library(tidyverse)
library(lubridate)
```

```{r, eval=FALSE}
# Create a dir for our bird data ("recursive" to create two levels at once):
dir.create("data/birds/", recursive = TRUE)

# The URL to our file:
birds_file_url <- "https://raw.githubusercontent.com/biodash/biodash.github.io/master/assets/data/birds/backyard-birds_Ohio.tsv"

# The path to the file that we want to download the data to:
birds_file <- "data/birds/backyard-birds_Ohio.tsv"

# Download:
download.file(url = birds_file_url, destfile = birds_file)

# Read the data:
birds <- read_tsv(file = birds_file)
```

```{r, echo=FALSE, eval=TRUE}
birds_file <- "data/birds/backyard-birds_Ohio.tsv"
birds <- read_tsv(file = birds_file)
```

<br>

----

## 2. What is *lubridate*?

*lubridate* provides tools that make it easier to parse and manipulate dates.

We will discuss the common use of lubridate under the following themes:  

1)	Parsing dates  

2)  Manipulating dates  
        a) *Instants*: Rounding dates, time zones  
        b) *Time spans*: Durations, periods, intervals

<br>

----

## 3. Parsing dates

*lubridate*’s parsing functions read strings into R as “date-time” objects.
Year is represented by `y`, month by `m`, and day by `d`.

Strings can be parsed using the following functions:
`dmy()`, `myd()`, `ymd()`, `ydm()`, `dym()`, `mdy()`, `ymd_hms()`.  

*Let us look at  some  examples*

```{r, include=TRUE}
# parsing by year, month, day
ymd(20170131)

# parsing by month, day, year
mdy("December 1st, 2020")

# parsing by day, month, year
dmy("01-Dec-2020")

dmy("01/Dec/2020")

dmy("01Dec2020")

# parsing by year, month, day, hour, minutes, and seconds
ymd_hms("2020-01-31 20:11:59")
```

<br>

----

## 4. Manipulating dates: Instants

*lubridate* distinguishes between *moments* in time (instants)
and *spans* of time (time spans).  

Instants are specific moments of time.
They are a combination of measurements on different units
(i.e, years, months, days, etc.).  

- `now()` returns the current system time.  
- `today()` returns the current system date.  

The individual values for `now()` and `today()` units can be extracted from an
instant and set with the accessor functions `second()`, `minute()`, `hour()`,
`day()`, `yday()`, `mday()`, `wday()`, `week()`, `month()`, and `year()`.

*Let us look at  some  examples*

```{r, include=TRUE}
# Find the current system date using function today ()
today()

# Find the current system time using function now ()
Now <- now()
Now

# Extract the day of the month from an object Now using function mday ()
mday(Now)

# Extract the day of the week from an object Now using function wday()
wday(Now)

# Extract the week of the year from an object Now using function week ()
week(Now)

# Extract the month from an object Now using function month ()
month(Now)

# Extract the year from an object Now using function year ()
year(Now)
```

### Rounding instants

Instants can be *rounded* to a convenient unit using the functions
`ceiling_date()`, `floor_date()`, and `round_date()`.  

- `ceiling_date()` takes a date-time object and rounds it up to the nearest
  boundary of the specified time unit.  

- `round_date()` takes a date-time object and time unit,
  and rounds it to the nearest value of the specified time unit.  

- `floor_date()` takes a date-time object and rounds it down to the nearest
  boundary of the specified time unit.  
  
*Let us look at  some  examples*

```{r, include=TRUE}
ceiling_date(Now, unit = "minute")

round_date(Now, unit = "minute")

floor_date(Now, unit = "minute")
```

### Time zones

Naming time zones is challenging because everyday names of time zones tend to be ambiguous. For example, USA has EST, or Eastern Standard Time.
However, both Australia and Canada also have EST!

To avoid confusion, R uses the international standard IANA time zones.
These use a consistent naming scheme “/”, typically in the form `<continent>/<city>` (there are a few exceptions because not every country lies on a continent).
Examples include `America/New_York`, `Europe/Paris`, and `Pacific/Auckland`.  

Unless otherwise specified, *lubridate* always uses UTC.
UTC (Coordinated Universal Time) is the standard time zone used by the scientific community and roughly equivalent to its predecessor GMT (Greenwich Mean Time).

Example: `ymd_hms("2021-03-27 11:54:54 EDT", tz="America/New_York")`

To find your current time zone, use the `Sys.timezone()` function:

```{r}
Sys.timezone()
```

To see the complete list of all time zone names, use `OlsonNames()`:

```{r}
# See the first four time zones in the list of the time zone
head(OlsonNames(), 4)
```

*lubridate* has two functions for working with time zones:

- `with_tz()`: Changes the time zone in which an instant is displayed.
  The clock time displayed for the instant changes,
  but the moment of time described remains the same. 

- `force_tz()`: Changes only the time zone element of an instant.
  The clock time displayed remains the same,
  but the resulting instant describes a new moment of time.

```{r}
x1 <- now()

# An example using with_tz()
x1a <- with_tz(x1, tzone = "Australia/Lord_Howe")
x1 - x1a

# Now use force_tz()
x1b <- force_tz(x1, tzone = "Australia/Lord_Howe")
x1 - x1b
```

<br>

----

## 4. Manipulating dates: Time spans

A timespan is a length of time that may or may not be connected to a particular instant.
For example, two months is a timespan.
*lubridate* has three timespan classes: Durations, Periods and Intervals.  

### Durations

Durations measure the exact amount of time that occurs between two instants.
  
Functions for working with durations include `is.duration()`,
`as.duration()` and `duration()`.
For specific lengths, `dseconds()`, `dminutes()`, `dhours()`,  `ddays()`, `dweeks()`
and `dyears()` convenient lengths.   

### Periods

Periods measure the change in clock time that occurs between two instants. 
  
Functions for working with periods include `is.period()`, `as.period()` and `period()`.
`seconds()`, `minutes()`, `hours()`, `days()`, `weeks()`, `months()` and `years()`
quickly create periods of convenient lengths.  

### Intervals

Intervals are timespans that begin at a specific instant and end at a specific instant.
Intervals retain complete information about a timespan.
They provide the only reliable way to convert between periods and durations.

Functions for working with intervals include `is.interval()`, `as.interval()`,
`interval()`, `int_shift()`, `int_flip()`, `int_aligns()`, `int_overlaps()`.

*Let us look at an example*

```{r, include=TRUE}
# John was born on 19841014. How old is John
h_age <- today() - ymd(19841014)
h_age

# Time difference in years
as.duration(h_age)
```

<br>

----

## 5. Plotting the bird data

We will plot the bird data using *ggplot2*.

First, we plot a bar graph of days of the week:

```{r, include=TRUE}
birds %>%
  mutate(Wday = wday(eventDate, label = TRUE)) %>% 
  ggplot(aes(x = Wday)) +
  geom_bar()
```

Second, we'll plot the relative abundance of different bird orders
by day of the week:

```{r, include=TRUE}
birds %>% 
  mutate(Wday = wday(eventDate, label = TRUE)) %>% 
  ggplot(aes(x = Wday, fill = order)) +
  geom_bar()
```

<br>

----

## 6. Breakout rooms!

<div class="puzzle">
<div>

### Exercise 1

Jane was born on January 31st, 1992. How old is Jane today?  

<details>
<summary>
Hints (click here)
</summary>
<br>
Use the functions `today()`, `mdy()`, and `as.duration()`.

</details>

<br>

<details>
<summary>
Solution (click here)
</summary>
<br>
```{r, include=TRUE}
Jane_age <- today() - mdy("January 31st, 1992")
as.duration(Jane_age)
```

</details>

</div> </div>

<div class="puzzle">
<div>

### Exercise 2   

Calculate the time differences between the last four time zones with the current
time.

<details>
<summary>
Hints (click here)
</summary>
<br>

- You can get the last four time zones using the `tail()` function in
  combination with the `OlsonNames()` functions.

- Then, you can compare the current time (`now()`) with times in different
  time zones using the `force_tz()` function.

</details>

<br>

<details><summary>Solution (click here)</summary>
<br>
```{r, include=TRUE}
# Current time 
C_time <- now()

# Last four time zones
tail(OlsonNames(), 4)

# Calculate time for UTC time zone 
UTC_time <- force_tz(C_time, tzone = "UTC")

# Calculate time difference
C_time - UTC_time

# Calculate time for W-SU time zone 
WSU_time <- force_tz(C_time, tzone = "W-SU")
WSU_time

# Calculate time difference
C_time - WSU_time
C_time

# Calculate time for WET time zone 
WET_time <- force_tz(C_time, tzone = "WET")

# Calculate time difference
C_time - WET_time

# Calculate time for Zulu time zone 
Zulu_time <- force_tz(C_time, tzone = "Zulu")

# Calculate time difference
C_time - Zulu_time
```

</details>

</div> </div>

<div class="puzzle">
<div>

### Bonus exercise

Remove the order "Passeriformes" from the bird data,
and plot relative abundance of order based on days of the  week.   

<details>
<summary>
Hints (click here)
</summary>
<br>
Use the functions `filter()` and `mutate()`.

</details>

<br>

<details>
<summary>
Solution (click here)
</summary>
<br>
```{r, include=TRUE}
# Remove Passeriformes:
birds_a <- filter(birds, order != "Passeriformes")

# Create the plot:
birds_a %>% 
  mutate(Wday = wday(eventDate, label = TRUE)) %>% 
  ggplot(aes(x = Wday, fill = order)) +
  geom_bar()
```

</details>

</div> </div>

