---
output: hugodown::md_document
title: "S03E07: Avoid Copy-pasting Code - Intro and Vectorization"
subtitle: "How to repeat operations without repeating your code"
summary: "In this first session on strategies for repeating operations without copy-pasting your code, we will focus on vectorization."
authors: [admin]
tags: [codeclub, iteration]
date: "`r Sys.Date()`"
lastmod: "`r Sys.Date()`"
toc: true

# Featured image
# To use, add an image named `featured.jpg/png` to your page's folder.
# Focal points: Smart, Center, TopLeft, Top, TopRight, Left, Right, BottomLeft, Bottom, BottomRight.
image:
  caption: "Artwork by @allison_horst"
  focal_point: ""
  preview_only: false

# Projects (optional).
#   Associate this post with one or more of your projects.
#   Simply enter your project's folder or file name without extension.
#   E.g. `projects = ["internal-project"]` references `content/project/deep-learning/index.md`.
#   Otherwise, set `projects = []`.
projects: []
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, eval=TRUE, message=FALSE, cache=FALSE)
```

<br> <br> <br>

------------------------------------------------------------------------

## Housekeeping

#### New to Code Club?

Check out the [Code Club Computer Setup](/codeclub-setup/) instructions,
which also has pointers for if you're new to R or RStudio.

#### Session goals

Today, you will learn:

- That you should avoid copy-pasting your code
- Which alternatives to repeating yourself exist in R
- What *vectorization* is and how to make use of it

#### R packages we will use

```{r}
## If necessary, install the packages we will use
if (!require(palmerpenguins)) install.packages("palmerpenguins")
if (!require(tidyverse)) install.packages("tidyverse")
if (!require(glue)) install.packages("glue")

## Load the packages we will use
library(palmerpenguins)
library(tidyverse)
library(glue)
```

<br>

------------------------------------------------------------------------

## I -- Avoid copy-pasting code

#### Don't Repeat Yourself (DRY)

Sometimes, you have a bit of code, and you need to repeat the operations in
that code *almost* exactly.

This can apply to anywhere from a single line to dozens of lines of code.
For instance, you may want to rerun a statistical model with different parameter
values, or repeat an analysis for different batches or subsets of samples.
In the context of our trusty penguins dataset,
we may want to repeat an analysis for each of the 4 morphological measurements.

Your first instinct is perhaps to copy-paste your code several times,
and make the necessary slight adjustments in each instance.
There are problems with this approach, including:

- You will end up with a lot of code,
  reducing clarity and making it more error-prone

- Making changes to the parts of the code shared by all blocks becomes
  challenging.

Avoiding such code repetition is where the programming mantra
"Don't Repeat Yourself" ("DRY") comes from.

<br>

#### Alternatives to repeating yourself

So what are the alternatives?

In R, two key approaches to avoiding copy-pasting your code are to use
_iteration_ to repeat a procedure, and you can do so either:

- Using a _**loop**_

- Using _**"functional programming"**_:
  apply a function multiple times with special functions ("functionals")
  from the base R `apply`-family or _purrr_'s `map`-family.

Loops are especially useful if you have a whole block of code that needs
to be rerun,
while functionals are easier to apply when you need to rerun a single function
call.

Furthermore, you can avoid code repetition by:

- _**Writing your own functions**_ (using _arguments_ to make them flexible)

- In simple cases, making use of R's _**vectorization**_ capabilities.

These approaches are clearer, less error-prone, and more flexible than
copy-pasting code. They can also be combined.

**We will tackle all of these approaches in Code Club in the upcoming weeks,
starting with vectorization this week.**

<br>

#### But first, an iteration example

Say that we wanted to compute the mean for each of the 4 measurements
for each penguin: bill length, bill depth, flipper length, and body mass.

To do this, we could write:

```{r}
mean(penguins$bill_length_mm, na.rm = TRUE)
mean(penguins$bill_depth_mm, na.rm = TRUE)
mean(penguins$flipper_length_mm, na.rm = TRUE)
mean(penguins$body_mass_g, na.rm = TRUE)
```

But that is a bit repetitive.
And it would get especially so if we had 20 different measurements.
Or if, instead of just computing the mean,
we wanted to perform an analysis consisting of multiple steps.

What would it look like to use iteration in a case like this?

- With a loop:
```{r}
## The columns we are interested in are columns 3 through 6
for (column_index in 3:6) {
    column_mean <- mean(penguins[[column_index]], na.rm = TRUE)
    print(column_mean)
} 
```

- With _purrr_'s `map()` function:
```{r}
map(penguins[, 3:6], mean, na.rm = TRUE)
```

<div class="alert alert-note"><div>

In this simple example where we are working with a dataframe,
a specialized _dplyr_ approach with `across()` also works:

```{r}
penguins %>% summarise(across(3:6, mean, na.rm = TRUE))
```

</div></div>

<br>

#### What about vectorization?

While iteration using loops or functionals is very useful,
in R, we don't need to use these strategies as much as in other languages.
The main reason for this is that R often makes use of _vectorization_.

To illustrate vectorization, we'll work with a vector of bill lengths
that we extract from the penguins dataframe
(though as we'll see later, all of this works in dataframes, too):

```{r}
penguins_noNA <- drop_na(penguins)              # Remove rows with NAs
bill_len <- penguins_noNA$bill_length_mm[1:10]  # Take first 10 values
bill_len
```

Say that we wanted to convert each value in the `bill_len` vector from
millimeters to inches.
Would we need to multiply each individual value by 0.0393701?

```{r}
bill_len[1] * 0.0393701
bill_len[2] * 0.0393701
bill_len[3] * 0.0393701

# And so on...
```

Or should we resort to a loop or a `map()`-type function here?
Fortunately, none of this is necessary!
You may already know that in R, you can simply do:

```{r}
bill_len * 0.0393701
```

Similarly, say that we wanted to log-transform every value in the vector,
then we can just use the `log()` function once for the entire vector:

```{r}
log(bill_len)
```

If you knew about this, perhaps you didn't even think of much of it?
Actually, it is worth dwelling on this capability,
which is called _vectorization_ and is a pretty unique feature of the R language.
In many other languages, you would in fact write a loop to transform each
individual value.

So let's learn a bit more about vectorization.

<br>

------------------------------------------------------------------------

## II -- Vectorization patterns

### A vector and a "scalar"

When we multiplied the value `0.0393701` with the vector `bill_len`,
`0.0393701` was automatically recycled as many times as needed to be multiplied
with *each individual value* in the `bill_len` vector.

```{r}
bill_len

bill_len * 0.0393701
```

(A single value like `0.0393701` is often called a "scalar" or a variable,
but in R it is really a vector of length 1.)

As mentioned, you would write a loop to do this in many other languages,
and in fact, under the hood, R *also* uses a loop to do this!

Vectorization is very useful for two reasons:

- _You_ don't have to write the loop (or another iteration construct),
  which saves you a fair bit of typing and makes the code clearer.

- The under-the-hood-loop is being executed *much* faster than a loop that you
  would write with R code, because it is written in `C`/`C++`.

### Vectors of equal length

We can also use vectorized operations when both vectors contain multiple items.
For instance, say we want to get the ratio of bill length to bill depth
for each penguin:

```{r}
## Like above with bill length, we create a vector with 10 bill depths:
bill_dp <- penguins_noNA$bill_depth_mm[1:10]

## We compute the ratio:
bill_len / bill_dp

bill_len
bill_dp
```

What happened here is that the first value is `bill_len` was divided by the first
value in `bill_dp`, the second value in `bill_len` by the second value in `bill_dp`,
and so forth.

This also works directly for the columns of a data frame:

```{r}
bill_ratio <- penguins$bill_length_mm / penguins$bill_depth_mm

head(bill_ratio)
```

In the above examples, both vectors had the same length.
In Exercise 1, you'll see that vectorization also works with two vectors with 
multiple values that _differ in length_.

### Vectorized functions

Above, we already briefly saw that we can simply pass a vector to the `log()`
function and it will compute the log for each of them and return a vector
of the same length.

So, the `log()` function works the same regardless of whether you pass a single
value or a vector with multiple values:

```{r}
log(21)

log(bill_len)
```

Because in R, a single value like `21` is really a vector of length 1,
this behavior makes sense. 

Just remember that for most functions,
you do really need to pass a vector and not just a sequence of numbers:

```{r, error=TRUE}
log(10, 15, 20)

log(c(10, 15, 20))
```

There are many other vectorized functions that will transform each value in a
vector, such as `round()` (rounding numbers) and `abs()` (taking
absolute numbers).

Other vectorized functions summarize a vector into a single value,
such as `sum()` and `mean()`.

<br>

----------------------------------------------------------

## Breakout Rooms I

<details><summary><b>Code to get set up</b> (click here)</summary>
<br>

```{r}
## If necessary, install the packages we will use
if (!require(palmerpenguins)) install.packages("palmerpenguins")
if (!require(tidyverse)) install.packages("tidyverse")
if (!require(glue)) install.packages("glue")

## Load the packages we will use
library(palmerpenguins)
library(tidyverse)
library(glue)

## Create a vector of bill lengths
penguins_noNA <- drop_na(penguins)
bill_len <- head(penguins_noNA$bill_length_mm, 10)
```

</details>

<div class="puzzle"><div>

### Exercise 1: Unequal length

Vectorization also works when two vectors with multiple elements do not have the
same length.
For instance, in the example below, we divide the first value by 10,
the second by 100, the third again by 10, and so on:

```{r}
bill_len / c(10, 100)
```

- **Given the length of `bill_len` (10), do you see any issues if you would**
  **divide by a vector of length 3? Try it out and see what happens.**

<details><summary><b>Hints</b> (click here)</summary>
<br>

While 10 is a multiple of 2, it is not a multiple of 3.
This means that the shorter vector will not be recycled in its entirety the
last time around.

</details>

<details><summary><b>Solution</b> (click here)</summary>
<br>

R will perform the operation but issue a warning about it:

```{r}
bill_len / c(10, 100, 1000)
```

</details>

- **Negate _every other_ value in the `bill_len` vector.**

<details><summary><b>Hints</b> (click here)</summary>
<br>

- Negation means turning a positive value into a negative value and vice versa
  (e.g. `3` => `-3` and `-15` => `15`).

- You can leave the other values as-is by multiplying them by 1.

</details>

<details><summary><b>Solution</b> (click here)</summary>
<br>

```{r}
bill_len * c(1, -1)
```

</details>

</div></div>

<br>

<div class="puzzle"><div>

### Exercise 2: Strings

The `glue` function from the package of the same name allows you to combine
literal strings with the values or strings contained in R objects.
For instance:

```{r}
island <- "Biscoe"
glue("The island of {island}")
```

So, you combine both literal strings and R objects in a single quoted string,
and access the values of R objects using braces `{}`.

- **Extract the names of the three islands contained in the `penguins` dataframe,**
  **and save them in an vector called `islands`.**

<details><summary><b>Hints</b> (click here)</summary>
<br>

Use the `unique()` function to get a "deduplicated" vector of islands,
i.e. with one entry per island.
   
</details>

<details><summary><b>Solution</b> (click here)</summary>
<br>

```{r}
islands <- unique(penguins$island)
islands

## Or:
islands <- penguins %>% pull(island) %>% unique()
```

Note: it is fine that `islands` is still a _factor_,
like the `island` column in `penguins` was.

</details>

- **Make use of vectorization to print each island's name like so:**
```{r}
#> The island of Torgersen
#> The island of Biscoe
#> The island of Dream
```

<details><summary><b>Solution</b> (click here)</summary>
<br>

```{r}
glue("Island of {islands}")
```

</details>

</div></div>

<br>

----------------------------------------------------------

## IV -- Vectorization with indices

We can also use vectorized solutions when we want to operate only on elements
that satisfy a certain condition.
To do so, we make use of R's ability to index a vector with a _logical vector_.

Let's say we don't trust any bill length measurement of over 40 mm,
and we want to remove those from our vector.

First, note that the code below will return a logical vector of the same length
as `bill_len`, with `TRUE` if the value is >40, and `FALSE` if it is not:

```{r}
bill_len > 40
bill_len
```

When we index the original vector with such a logical vector
(sometimes referred to as a _mask_), we only get the values >40:

```{r}
bill_len[bill_len > 40]
```

This is very succinct and powerful!

With a similar strategy, you can also retain all elements of the vector
but manipulate some of them:

```{r}
## We create a separate vector so we don't change the original one:
bill_len_ed <- bill_len

## Only change values > 40:
bill_len_ed[bill_len > 40] <- bill_len_ed[bill_len > 40] - 100
bill_len_ed
```

But for those kinds of operations,
the vectorized `ifelse()` function is easier and clearer:

```{r}
bill_len_ed <- ifelse(bill_len > 40, bill_len - 100, bill_len)
bill_len_ed
```

<br>

<div class="alert alert-note"><div>

When creating logical vectors, the `any()` and `all()` functions are very handy.

For instance, say we had a vector of p-values:

```{r}
pvals <- c(0.06, 0.048, 0.01, 0.73)
```

To check whether any of the p-values are significant:

```{r}
any(pvals < 0.05)
```

To check whether all of the p-values are significant:

```{r}
all(pvals < 0.05)
```

Moreover, because `TRUE` corresponds to 1 and `FALSE` to 0,
you can also directly count the number of elements that satisfy a condition:

```{r}
sum(pvals < 0.05)
```

</div></div>

<br>

<br>

----------------------------------------------------------

## Breakout Rooms II

<div class="puzzle"><div>

### Exercise 3: Logical vectors

- Create a vector `bill_len_NA` where all values > 40 have been turned
  into `NA`s.

<details><summary><b>Solution</b> (click here)</summary>
<br>

```{r}
## Using logical vector subsetting:
bill_len_NA <- bill_len
bill_len_NA[bill_len_NA > 40] <- NA
bill_len_NA

## Or, using `ifelse()`:
bill_len_NA <- ifelse(bill_len > 40, NA, bill_len)
bill_len_NA
```

</details>

- Remove all `NA`s from `bill_len_NA`. 
  (If you don't know the function to identify `NA`s in a vector,
  take a look at the Hints.)

<details><summary><b>Hints</b> (click here)</summary>
<br>

- The function `is.na()` will check which values in a vector are `NA`s:
  it returns a logical vector with `TRUE`s for `NA` values and `FALSE`s for non-`NA`
  values.

- Since you want to _remove_ `NA` values, you need to negate the output of the
  `is.na()` function when subsetting.
  You can negate logical tests in R with a `!`.
  So, `!is.na()` would have `TRUE` for non-`NA` values, which would allow you to
  keep them.
  
</details>

<details><summary><b>Solution</b> (click here)</summary>
<br>

```{r}
bill_len_NA[!is.na(bill_len_NA)]
```

</details>

</div></div>

<br>

<div class="puzzle"><div>

### Exercise 4: `ifelse()` in a plot

With _ggplot_,
make a `geom_point()` plot of `bill_length_mm` versus `bill_depth_mm` in
Gentoo Penguins.
In this plot,
highlight penguins with a bill length to bill depth ratio larger than 3.5
by giving those points a different color.

Don't hesitate to look at the Hints if you're not sure how to approach this.

<details><summary><b>Hints 1 </b> (click here)</summary>
<br>

- Create a new dataframe with:
  - Just Gentoo penguins (use the function `filter()`)
  - No `NA`s (use the function `drop_na()`)
  - A new column with a logical vector indicating whether the bill length to bill
    depth ratio is >3.5
    (e.g., use the `mutate()` function with an `ifelse()` statement).
    
- When creating the plot, assign the new column to the `color` aesthetic. 

</details>

<details><summary><b>Hints 2 </b> (click here)</summary>
<br>

- Here is some example skeleton code for the data processing:

```{r, eval=FALSE}
... <- penguins %>%
  ## Only retain rows for 1 penguin species:
  filter(species == ...) %>%
  ## Remove rows with NAs:
  ... %>% 
  ## Create a new column `ratio` with a logical vector:
  mutate(ratio = ifelse(..., "> 3.5", "< 3.5"))
```

- Here is some example skeleton code for the plot:

```{r, eval=FALSE}
ggplot(...) +
  ## Use the new column with the logical vector for the `color` aesthetic
  geom_point(aes(x = ..., y = ..., color = ...))
```

</details>

<details><summary><b>Solution</b> (click here)</summary>
<br>

```{r}
## Create the new dataframe, all tidyverse
gent <- penguins %>%
  filter(species == "Gentoo") %>%
  drop_na() %>% 
  mutate(
    ratio = ifelse(bill_length_mm / bill_depth_mm > 3.5, "> 3.5", "< 3.5")
    )

## Make the plot:
ggplot(gent) +
  geom_point(aes(x = bill_length_mm, y = bill_depth_mm, color = ratio))
```

</details>

</div></div>

<br>

----------------------------------------------------------

## Bonus -- Vectorization with matrices

We can also perform vectorized operations on *entire matrices*.
With the following matrix:

```{r}
## We use the "sample" function to get 25 random values between 1 and a 100,
## and put those in a 5*5 matrix:
mat <- matrix(sample(1:100, 25), nrow = 5, ncol = 5)
mat
```

...we could multiple all values by 10 or get the square of each value simply as follows:

```{r}
mat * 10

mat * mat
```

