---
output: hugodown::md_document
title: "Session 7: R Markdown"
subtitle: "Text + Code + Output"
summary: "In this 7th session of Code Club, we will learn about Markdown syntax and some of the great functionality of R Markdown."
authors: [admin]
tags: [codeclub, markdown, rmarkdown]
date: 2021-01-24
lastmod: 2021-01-24
toc: true

# Featured image
# To use, add an image named `featured.jpg/png` to your page's folder.
# Focal points: Smart, Center, TopLeft, Top, TopRight, Left, Right, BottomLeft, Bottom, BottomRight.
image:
  caption: "Artwork by @allison_horst"
  focal_point: ""
  preview_only: false

# Projects (optional).
#   Associate this post with one or more of your projects.
#   Simply enter your project's folder or file name without extension.
#   E.g. `projects = ["internal-project"]` references `content/project/deep-learning/index.md`.
#   Otherwise, set `projects = []`.
projects: []
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval=TRUE)

knitr::knit_hooks$set(wrapper = function(before, options) {
  # the original chunk might be indented
  if (is.null(indent <- options$indent)) indent <- ''
  
  # hide the wrapper=TRUE option
  opts <- gsub(', wrapper=TRUE', '', options$params.src)
  
  if (before) {
    # add the header
    sprintf('\n\n%s````\n```{r,%s}\n````', indent, opts)
  } else {
    # add the footer
    #sprintf('\n\n%s````\n```\n````\n', indent)
    sprintf('%s```', indent)
  }
})
```

<br>
<br>
<br>

-----

## Introduction

R Markdown consists of an amazing ecosystem of R packages to produce many types
of technical content. Its signature capability is that is can **run R code
and print the code along with its results and nicely formatted prose.**

To understand R Markdown, we need to learn about three new things:

1. **Markdown**, a very lightweight text formatting language.

2. **Code chunks**, which allow us to incorporate R code that can be executed
   and whose results we can display in text, figures, and tables.

3. **The YAML header**, which encodes metadata about the output,
   such as the desired output format and specific formatting features.
  
We'll focus on HTML page output, but will glimpse at the **many possibilities for
the output format**: with R Markdown, it is possible to create not just
technical reports, but also slide decks, websites, books, scientific articles,
and so on.

<br>

#### Setup

At the core of the R Markdown ecosystem is the *rmarkdown* package.
We need to install this but don't need to load it:

```{r, echo=TRUE, eval=FALSE}
install.packages("rmarkdown")
```

Inside your directory for Code Club, create a directory for this week:

```{r, eval=FALSE}
dir.create('S07')
```

```{r, echo=FALSE, include=FALSE}
library(tidyverse)
library(palmerpenguins)
```

<br>

#### First, an example

Before we go into details, let's first see a quick demonstration of what we
are talking about:

1. Go to `File` => `New File` => `R Markdown`,
   change the *Title* to "Markdown demo", and click `OK`.

2. Take a look at the R Markdown document,
   and notice that there seems to be some sort of header (=> **_YAML_**),
   followed by R code wrapped in strange constructs with backticks
   (=> **_Code chunks_**),
   and plain written text (=> **_Markdown_**).

3. Before we can create output, we need to save the document.
   Click the `Save` button and save the files as `demo.Rmd` inside
   your newly created directory.

4. Now click the **Knit button** in one of the top bars,
   and a document should show up in a pop-up or the Viewer pane.
   This is the *rendered output* from the R Markdown document.

Notice that the YAML header is not printed, at least not verbatim,
while some of the code is printed, and we also see code output including a plot!

This is what the raw and rendered output look side-by-side:

<p align="center">
<img src=img/rmd-demo-both.png width="1000px">
</p>

We'll now talk about Markdown, code chunks, and the YAML header in turn.

<br>

----

## I: Markdown

Markdown is a very lightweight language to format plain text files, 
which evolved from simple in-line formatting applied in emails before those
started using HTML.

Need to emphasize a word without being able to make it
italic or bold? How about adding emphasis with asterisks \*like so\*? 

<br>

#### An overview of commonly used Markdown syntax

| Syntax            | Result
|-------------------|-------|
| # My Title        | Header level 1 (largest)
| ## My Section     | Header level 2
| ### My Subsection | Header level 3 &ndash; and so forth
| \*italic\* or \_italic\_       | *italic*
| \*\*bold\*\* or  \_\_bold\_\_     | **bold**
| `[Markdown Guide](markdownguide.org)`   | [Markdown Guide](https://markdownguide.org) (Link with custom text)
| \!\[\](path/to/figure.png)   | Figure
| - List item       | Unordered (bulleted) list
| 1. List item      | Ordered (numbered) list
| `` `inline code` `` | `inline code`
| ```` ``` ```` or 4 leading spaces     | Start/end of generic code block (for formatting only)
| ```` ```r ```` | Start of `r` code block (end with ```` ``` ````) (for formatting only)
| `---`             | Horizontal rule (line)

To see this formatting in action, see below an example of a raw Markdown file
on the left, and its *rendered* (formatted) output on the right:

<p align="center">
<img src=img/md-demo.png width="1000px">
</p>

<div class="alert alert-note">
<div>
"Plain" Markdown files have the extension `.md`,
whereas R Markdown files have the extension `.Rmd`.
</div>
</div>
  
<br>

----

## II: Integrating R code

As we saw above, plain Markdown has syntax for code *formatting*,
but the code is not actually being executed.
**In R Markdown, however, we are able run code!**
The syntax to do so is only slightly modified from what we saw above:

- For **inline code**, we add `r ` and a space before the R code that is to be
  executed, for example:

   | Raw    | Rendered
   |--------|---------
   | There are `` `r '\x60r 365*24\x60'` `` hours in a year | There are `r 365*24` hours in a year

- To generate *code blocks*, which we call **code chunks**
  in an R Markdown context:  
  following the three backticks, we add r *inside curly braces* (`{r}`).  
  
  We can optionally add settings that we want to apply to that chunk
  and/or chunk labels:  
  `{r, option1=value, ...}` or `{r unique-chunk-label, option1=value, ...}`.
  
  <div class="alert alert-note">
  <div>
  RStudio keyboard shortcut to insert a code chunk:
  <kbd>Cmd/Ctrl</kbd>+<kbd>Alt</kbd>+<kbd>I</kbd>.  
  </div>
  </div>
  
<br>

#### Code chunk examples

- A code chunk with **default options**...

  <p align="left">
  <img src=img/example-chunk-1.png width="100%">
  </p>
  
  ...will be executed and shown followed by the output of the code:
  
  ```{r}
  mean(penguins$bill_depth_mm, na.rm = TRUE)
  ```

- As an example of using a **code chunk option**,
  we will disable printing the code using `echo=FALSE`
  (the code will still run and the output will still be shown):
  
   <p align="left">
   <img src=img/example-chunk-2.png width="100%">
   </p>

   ```{r, echo=FALSE}
   mean(penguins$bill_depth_mm, na.rm = TRUE)
   ```

- **Figures** can, of course, also be printed:

  <p align="left">
  <img src=img/example-chunk-3.png width="100%">
  </p>
  
  ```{r, fig.cap="Fig. 1: Bill length and depth are correlated within species, <br> and differ subtly between species."}
  ggplot(penguins) +
    geom_point(aes(x = bill_length_mm, y = bill_depth_mm, color = species)) +
    theme_bw()
  ```

Above, we added a **caption** for the figure using the `fig.cap` argument
(with a little trick to force a line break, using the `<br>` HTML syntax).

<div class="alert alert-note">
<div>

#### Code chunk options

Here is an overview of some the most commonly made *changes to defaults*
for code chunk options. This quickly gets confusing, but you'll get the hang
of it after experimenting a bit.

  - `echo=FALSE` -- Don't print the code in the output file.
  - `eval=FALSE` -- Don't run (**eval**uate) the code.
  - `include=FALSE` -- Run but don't print the code, nor any of its results. 
  - `results="hide"` -- Don't print the *text output* of the code.
  - `fig.show="hide"` -- Don't print *figures* produced by the code.

Furthermore, you can use `message=FALSE` and `warning=FALSE`
to suppress any *messages* (like the output when loading packages) and
*warnings* (like the warning for the penguin figure above), respectively,
that R might produce.

For figures, the following options are especially useful:

  - `fig.cap="My caption"` -- Include a caption.
  - `fig.asp=0.6` -- Aspect ratio.
  - `fig.width=6` -- Width of in inches: same as sizing in regular R code.
  - `fig.height=9.6` -- Height in inches: same as sizing in regular R code.
  - `out.width="70%"` -- Width at which the figure is *printed in the document*.

Finally, if your document takes a long time to knit, use `cache=TRUE` to
enable caching of results.
</div>
</div>

<div class="alert alert-note">
<div>

#### Default chunk options

It is often practical to set default chunk options for the entire
document, and you can do so with the `opts_chunk$set()` function as shown below:

<p align="left">
<img src=img/global-chunk.png width="100%">
</p>

This is usually done in separate *"global setup chunk"* at the start of the
document.  
Whenever necessary, you can then override these defaults for specific chunks.

</div>
</div>

<br>

----

## III: The YAML header

YAML (*"YAML Ain't Markup Language"*) is a simple format commonly used for
*configuration files*,
which allows you to provide key-value pairs such as `author: John Doe`.

In R Markdown files, it is used as a **header** which configures certain
aspects of the output, especially the formatting.
Put another way, the YAML header *contains the metadata* for the output.

#### A basic YAML header

Here is an example of a very basic YAML header:

```
---
author: My name
title: The document's title
output: html_document
---
```

Note the lines which just contain **three dashes**,
which mark the beginning and the end of the YAML header.

#### Adding options

Often, a value (like `html_document`) can itself be given key-value pairs to
specify additional options --
see the example below where we include a Table of Contents (`toc`)
and also set it to "float":

```
---
output:
  html_document:
    toc: true
    toc_float: true
---
```

- Note the syntax changes (newlines and added indentation) between the above
  two examples, this is perhaps a little awkward and often leads to mistakes.

- Indentation in YAML is using two or four spaces (no tabs!) per indentation
  level, and it is *sensitive to indentation errors*.
  (Fortunately, RStudio inserts spaces for tabs by default -- check/set in
  `Tools` => `Global Options` => `Code` => `Editing`.)

<div class="alert alert-note">
<div>

#### Some options for `html_document` output

`html_document` is the most commonly used output format for R Markdown
documents, and here are few particularly useful options to customize the output:

  - `code_download: true` -- Include a button to download the code.
  - `code_folding: hide` -- Using `hide` or `show`
     will enable the folding of code chunks, with `hide` hiding them by default.
  - `toc: true` -- Include a table of contents (Also: `toc_depth: 3` sets depth to 3,
    `toc_float: true` lets the TOC "float" as you scroll down the document).
  - `number_sections: true` -- Number the section headings.
  - `df_print: paged` -- Get nicely formatted and paged data frame printing
    (also try: `df_print: kable`).
  - `theme: cerulean` -- Use a pre-built theme,
     controlling the overall look and feel of the document.
     See [here](https://www.datadreaming.org/post/r-markdown-theme-gallery/)
     for a visual overview.

<figure>
<p align="center">
<img src=img/rmd-3themes.png width="100%">
<figcaption> Three HTML document <code>theme</code> options:
<code>darkly</code>, <code>flatly</code>, and <code>cerulean</code>.</figcaption>
</p>
</figure>

</div>
</div>

<br>

---

## IV: R Markdown and RStudio

The RMarkdown ecosystem of packages is being developed by RStudio,
so it should come as no surprise that the RStudio IDE has some nice
RMarkdown functionality.

#### Knitting and previewing your document

The process of rendering an R Markdown file into another format,
as specified by the YAML header, is called **knitting**.
We already saw the button to knit the current document
(keyboard shortcut: <kbd>Cmd/Ctrl</kbd>+<kbd>Shift</kbd>+<kbd>K</kbd>).

<div class="alert alert-note">
<div>
If you get preview pop-up windows in RStudio, click the cog wheel icon
next to the Knit button, and then select "*Preview in Viewer Pane*".
</div>
</div>

<br>

Instead of knitting the entire document,
you can also **run individual code chunks** using the green "play button"
(or <kbd>Cmd/Ctrl</kbd>+<kbd>Shift</kbd>+<kbd>Enter</kbd>),
or all code chunks up until the current one (button to the left of the play button).

For a **live preview (!)** of R Markdown output for your active document,  
use the *infinite moon reader* from the *xaringan* package:

```{r, eval=FALSE}
install.packages("xaringan")

# Simply running the function without arguments will start the preview:
xaringan::inf_mr()

# To shut down the preview server, if needed, run `servr::daemon_stop()`
```

<br>

#### Visual Markdown Editor

If your RStudio version is at least 1.4 (Click `Help` => `About RStudio`),
which was released last fall, you can also use the *Visual Markdown Editor*.

This makes writing in R Markdown almost like using a word processor,
and also includes many other features such as better citation support
with Zotero integration. Read more about the visual editor
[here](https://rstudio.github.io/visual-markdown-editing).

To switch between the visual editor and regular ("source") editing mode,
click the A-shaped ruler button in the top-right corner or press
<kbd>Cmd/Ctrl</kbd>+<kbd>Shift</kbd>+<kbd>F4</kbd>.

This is what our document looks like in the visual editor --
kind of intermediate between the raw R Markdown and the rendered output:

<p align="center">
<img src=img/visual-editor.png width="70%">
</p>

<br>

----

## V: A single source doc, many output formats!

One of the greatest features of R Markdown is that you can output to many
formats.
So from one source document, or very similar variants,
you can create completely different output depending on what you need.

#### Built-in output formats

The [built-in output formats](https://rmarkdown.rstudio.com/docs/reference/index.html#section-output-formats), which can be used with just the *rmarkdown*
package, are listed below. These include HTML, PDF, Word, PowerPoint,
and different HTML slide show formats!

<p align="center">
<img src=img/rmarkdown-output-formats.png width="80%">
</p>

<br>

#### Extension output formats

It's worth highlighting a few of the output formats that can be used with the
following packages in the R Markdown ecosystem:

- [*distill*](https://rstudio.github.io/distill/) --
  An output format geared towards technical content,
  e.g. with extended support for equations, citations, and footnotes.
  Can also create websites.

- [*rticles*](https://github.com/rstudio/rticles) --
  R Markdown templates to format output for specific scientific journals.

- [*flexdashboard*](https://rmarkdown.rstudio.com/flexdashboard/) --
  Create interactive "dashboards" to present data.

- [*bookdown*](https://bookdown.org/) -- A book format,
  the [R Markdown book](https://bookdown.org/yihui/rmarkdown/) is an example.
 
- [*xaringan*](https://bookdown.org/yihui/rmarkdown/xaringan.html) --
  Create fancier presentation slides thanks to a JavaScript library.  

Starting to use these and other output formats is often as simple as changing
the YAML header:

```YAML
---
output: distill::distill_article
---
```

<br>

----

## Breakout rooms!

In the exercises, we will work with an `.Rmd` file that you can download as follows:

```{r, eval=FALSE}
# dir.create("S07")     # You should have already done this

# Save the URL for the Rmd file:
todays_rmd <- 'https://raw.githubusercontent.com/biodash/biodash.github.io/master/content/codeclub/07_rmarkdown/penguins.Rmd'

# Download the Rmd file:
download.file(url = todays_rmd, destfile = 'S07/penguins.Rmd')
```              

Next, open the document in RStudio, and fire up the _infinite moon reader_:

```{r, eval=FALSE}
# install.packages("xaringan")

xaringan::inf_mr()
```

This way, you will be able to nearly instantaneously see the effect of your
changes: save the document whenever you want the server to update.

You can use either the "visual editor" or the regular ("source") editor --
and you could also start by compating the two.

<div class="puzzle">
<div>

### Exercise 1: Output formatting with YAML

In this exercise, you will fiddle with the YAML header to modify
aspects of the `html_document` output format:

- Add a `theme` key to `html_output`, and try a few of the available
  value options ("*default*", "*cerulean*", "*journal*", "*flatly*", "*darkly*",
  "*readable*", "*spacelab*", "*united*", "*cosmo*", "*lumen*", "*paper*",
  "*sandstone*", "*simplex*", "*yeti*").
  
  Determine, once and for all, what the best theme is.

- Try some of the other options mentioned above (`code_download`,
  `code_folding`, `toc`, `toc_float`, `toc_depth`, `df_print`),
  and look at the effects on the rendered output.

<details>
<summary>
Hints (click here)
</summary>
<p>

- To add options to `html_document` in the YAML header,
  you'll need to go from `output: html_document` on a single line,
  to a multi-line format with indentation,
  and with a colon added after `html_document`:
  
  ```YAML
  output:
    html_document:
      <option>
  ```
</p>
</details>

<br>

<details>
<summary>
Solutions (click here)
</summary>
<br>

- An example YAML header with several options added:

```YAML
---
title: "Penguins, demystified."
author: "Jelmer Poelstra"
date: "1/29/2021"
output:
  html_document:
    theme: flatly
    toc: true
    toc_float: true
    toc_depth: 5
    number_sections: true
    code_download: true
    code_folding: hide
    df_print: kable
---
```

<br>
</details>

</div>
</div>

<br>


<div class="puzzle">
<div>

### Exercise 2: Code chunks

Our output document looks nice, but there is plenty of room for
improvement.
In this exercise, we'll refine the output using code chunk options.

Before you start, take another look at the
box [Code chunk options](#code-chunk-options)
above.

- Did you notice those messages (when the *tidyverse* is loaded) and warnings
  (for the two plots) in the output?
  Let's get rid of those all at once:
  suppress R *messages* and *warnings* for all chunks by adding arguments to the 
  `knitr::opts_chunk$set()` function in the first code chunk. 

- Currently, the code line in the `install-package` code chunk is commented out
  to avoid the code from running, while still printing it.
  Try to accomplish this using a code chunk option instead,
  so you can uncomment the line.

- We do want to print the code in some cases, but not in others.
  For the chunk labeled `print-tibble`, which prints `penguins`,
  alter the settings such that the code is no longer printed.

- Our first figure is kind of squished, and the point and font sizes
  are perhaps too large.
  Compare this with the second figure,
  which has a different setting *only* for `out.width`.
  
  Play around with the values for the three options that are
  already in the code chunks (`fig.width`, `out.width`, and `fig.asp`),
  for one or both figures, see what the effects are, and try to make some
  improvements.
  
  Do you understand the difference between the two methods to indicate
  the figure size (`fig.width` and `out.width`)?

- Insert a new code chunk that prints the `penguins_raw` tibble in some way
  (this is available in your environment).

<details>
<summary>
Hints (click here)
</summary>
<p>

- To suppress messages and warnings throughout:  
  Add `message=FALSE` and `warnings=FALSE` inside `knitr::opts_chunk$set()`
  in the `setup` chunk. 

- To avoid running the code:  
  Use `eval=FALSE` in the header of the `install-package` code chunk.
  
- To avoid printing the code:  
  Use the `echo` option in the header of the `print-tibble` code chunk.

- Figure sizing:  
  There are two types of sizes that you can set:
  the size at which R creates figures (`fig.width` and `fig.height`),
  and the size at which the figures are inserted in the document
  (`out.width` and `out.height`).
  The former will effectively only control relative font and point sizes,
  whereas the latter controls the "actual" / final size.
  For more details and advice, see
  [this section](https://r4ds.had.co.nz/graphics-for-communication.html#figure-sizing)
  in R for Data Science.

  The aspect ratio (`fig.asp`) is height/width, so a value smaller than one
  creates a wide figure and a value larger than one creates a narrow figure.
  
  Here, we've been setting width only -- you can also set `fig.height` and
  `out.height`, but these options become redundant when you set
  the width *and* the aspect ratio.
</p>
</details>

<br>

<details>
<summary>
Solutions (click here)
</summary>
<br>

- To suppress messages and warnings throughout:  
  `knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)`

- To avoid running the code: 
  `{r install-package, eval=FALSE}`

- To avoid printing the code:  
  `{r print-tibble, echo=FALSE}`

- Figure sizing:  
  Example settings for better-sized figures --  
  `{r plot-bills, out.width="80%", fig.width=6, fig.asp=0.7}` 

- A code chunk to print the `penguins_raw` tibble
  (replace single quotes by backticks):

```
'''{r}
penguins_raw
'''
```

<br>
</details>

</div>
</div>

<br>



<div class="puzzle">
<div>

### Bonus 1: Markdown and inline code

The formatting for the prose in our document could also be improved.
For instance:

- Use inline code *formatting* in a couple of cases where this is appropriate.

- Instead of simply saying "8 variables (n = 344 penguins)"
  (under the *Summary of the dataset*" heading),
  use *inline R code* that makes these calculations and print the results.
  
- Try a couple of other things: heading levels (one of them is currently not
  right!), *italic* text, **bold** text,
  and/or ordered (numbered) and unordered (bulleted) lists.
  
<details>
<summary>
Hints (click here)
</summary>
<p>

- Simply put backticks around the inline text you want have formatted as code.
  You can do this, for instance, for mentions of `palmerpenguins::penguins`.

- For inline code that runs, use `` `r '\x60r my-code\x60'` ``.
  
  The number of variables and penguins in the penguin dataset are
  the number of columns and rows, respectively, in the `penguin` tibble.

</p>
</details>

<br>

<details>
<summary>
Solutions (click here)
</summary>
<br>

Inline calculation of the number of variables and penguins:

```
[...] that contains `r '\x60r ncol(penguins)\x60'` variables
(n = `r '\x60r nrow(penguins)\x60'` penguins).
```

<br>
</details>

</div>
</div>

<br>


<div class="puzzle">
<div>

### Bonus 2: Other output formats

Try one or more output formats other than `html_document`, see
  [this website](https://rmarkdown.rstudio.com/docs/reference/index.html#section-output-formats)
  for the list of available options.
  If you want to try presentations, note that three dashes `---` are used
  to separate slides.
  
<div class="alert alert-note">
<div>
It might be confusing that on the website linked to above (see also the
screenshot in section V), the output formats are listed  *functions*
(`html_document()` rather than `html_document`) -- but this is simply
because under the hood, these functions are called via the YAML header.
</div>
</div>

</div>
</div>

<br> <br>

----

## Go further

#### Pitfalls / Tips

- **The working directory**  
  By default, the working directory for an R Markdown document is the directory
  in which the file resides.
  
  This can be a bit annoying if you're used to using your project's root directory
  as your working directory (which you should be) and the R Markdown file is
  not in the project's root directory (which it probably shouldn't be).
  Nevertheless,
  simply using `../` notation to move one or two directories up should
  generally work.
  
  If you really need to set a different working directory, you should be aware
  that surprisingly, setting the working directory with `setwd()` in a code
  chunk is **not persistent** across code chunks.
  To set a different working directory for the entire document,
  use `knitr::opts_knit$set(root.dir = '/my/working/dir')` in a setup chunk.
  
- **Chunk labels**  
  Chunk labels are optional but if you do give them, note that they have to be
  *unique*: the document will fail to render if have two chunks with the same
  label. Also, *avoid using spaces and underscores in the labels*
  (`good-chunk-label`, `bad chunk label`, `bad_chunk_label`).

<br>

#### Tables

- **Tables produced by Markdown text**  
  The syntax for basic Markdown tables is as follows:

  ```
  | Time          | Session | Topic    |
  |:--------------|:-------:|---------:|
  | _left_        | _center_| _right_  |
  | Wed 5 pm      | 1       | Getting started  |
  | Fri 3 pm      |         |          |
  | Wed 5 pm      | 2       | *dplyr*  |
  | Fri 3 pm      |         | *Break*  |
  ```
  
  | Time          | Session | Topic    |
  |:--------------|:-------:|---------:|
  | _left_        | _center_| _right_  |
  | Wed 5 pm      | 1       | Getting started  |
  | Fri 3 pm      |         |          |
  | Wed 5 pm      | 2       | *dplyr*  |
  | Fri 3 pm      |         | *Break*  |

  In the Visual Markdown editor in RStudio, you can simply insert a table
  with a little dialogue box after clicking `Table` => `Insert Table`.

- **Tables (dataframes) produced by R code**  
  Using `kable(my_df)` in a code chunk will create nicer output for individual
  dataframes (recall the `df_print: kable` YAML option for document-wide "kable"
  printing).
  
  There are many packages available for more advanced options,
  such as [*GT*](https://gt.rstudio.com/),
  [*DT*](https://rstudio.github.io/DT/), and
  [*reactable*](https://glin.github.io/reactable/).

<br>

#### Websites

Note that `rmarkdown::render_site()` can create simple websites that connects
multiple pages with a navigation bar.
All you need is a simple YAML file called `_site.yml` with some settings,
and a file for the front page which needs to be called `index.Rmd`.
See [here in the R Markdown book](https://bookdown.org/yihui/rmarkdown/rmarkdown-site.html)
for more details.

Options with more features, like a blog, are
[*distill websites*](https://rstudio.github.io/distill/website.html),
and the [*blogdown* package](https://bookdown.org/yihui/blogdown/)
for Hugo sites.

<br>

#### Further resources

- Free online books by the primary creator of R Markdown and other authors:
  - [R Markdown -- The Definitive Guide](https://bookdown.org/yihui/rmarkdown/)
  - [R Markdown Cookbook](https://bookdown.org/yihui/rmarkdown-cookbook/)
  
- [RStudio's 5-page R Markdown Reference PDF](https://rstudio.com/wp-content/uploads/2015/03/rmarkdown-reference.pdf)
- [RStudio's R Markdown Cheatsheet](https://github.com/rstudio/cheatsheets/raw/master/rmarkdown-2.0.pdf)
- [RStudio R Markdown lessons](https://rmarkdown.rstudio.com/lesson-1.html)
- [Markdown tutorial](https://commonmark.org/help/tutorial/)
