---
title: "Session 4: Visualizing Data"
subtitle: "Introduction to Data Visualization with **ggplot2**"
summary: "In this session of Code Club, we'll look at how to visualize data in R using **ggplot2**."  
authors: [michael-broe]
date: "2020-12-10"
output: hugodown::md_document
toc: true
image: 
  caption: "Image from http://r-statistics.co/ggplot2-Tutorial-With-R.html"
  focal_point: ""
  preview_only: false
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

<br> <br> <br>

------------------------------------------------------------------------

## New To Code Club?

-   First, check out the [Code Club Computer Setup](/codeclub-setup/)
    instructions, which also has some pointers that might be helpful if
    you're new to R or RStudio.

-   Please open RStudio before Code Club to test things out -- if you
    run into issues, join the Zoom call early and we'll troubleshoot.

------------------------------------------------------------------------

## Session Goals

-   Learn the philosophy of **coding** a graphic.
-   Learn the basic **template** of a **ggplot2** graphic, so you can
    reuse it for multiple chart types.
-   Learn how you can quickly add visual information to a graphic using
    **aesthetics** and **layers**.

------------------------------------------------------------------------

## Intro: The ggplot2 philosophy

We have already seen that in R, instead of manually manipulating data
frames as you might do when editing Excel sheets, we **code** the
operations we want to perform using **dplyr** verbs like `select()`,
`mutate()`, `inner_join()`, and so on.

In a similar way when performing visualization, instead of clicking on a
chart type in Excel, we **code the chart** in R.

And just as **dplyr** gives us efficient ways to manipulate data frames,
**ggplot2** (which is also part of the **tidyverse**) gives us efficient
ways to manipulate charts/plots/graphics (we use these terms
interchangeably).

The **gg** in **ggplot2** stands for *grammar of graphics*, a systematic
approach for designing statistical plots developed by Leland Wilkinson.
The idea behind this was to think about 'pulling apart' various plots
into their shared component pieces, then provide code that could put
them together again. We can then create new plots like we create new
sentences (once we understand this grammar).

There are two parts to this. First, the 'nouns and verbs' we need to
work with plots are very different than those we need to work with data
frames. **ggplot2** is like a mini-language of its own, with its own
verbs and syntax.

Second, this notion of pulling apart a graphic leads to the idea of
*layers*. You can build up a plot of any complexity by *overlaying*
different views of the same data.

There's a learning curve here for sure, but there are a couple of things
that help us.

First, every graphic shares a *common template*. This is like thinking
about the sentence "The cat sat on the mat" grammatically as the
template `NP V PP` (`N`oun `P`hrase "The cat", `V`erb "sat",
`P`repositional `P`hrase "on the mat"). Once you understand this
structure you can "say" a *lot* of different things.

(And I mean a *lot*. The [ggplot cheat
sheet](https://github.com/rstudio/cheatsheets/blob/master/data-visualization-2.1.pdf) lists over 40 geoms, but because this is a
language, users can create their own
[extensions](https://exts.ggplot2.tidyverse.org/gallery/) that you can
also utilize, adding over 80 more.)

Second, the way we put layers together is identical to the way we use
pipes. You can read `%>%` as "and then": `select()` and then `mutate()`
and then `summarize()`. In graphics, we can say "show this layer, and
then *overlay* this layer, and then *overlay* this layer", etc., using a
very similar syntax.

<br>

------------------------------------------------------------------------

## Examples

So how does this work in practice? We'll work through visualizing the
**iris** dataset that you've seen before. This is an extremely famous
[dataset](https://en.m.wikipedia.org/wiki/Iris_flower_data_set) that was
first analyzed by R. A. Fisher in 1936: *The use of multiple
measurements in taxonomic problems*. He was attempting to use the petal
and sepal measurements to discriminate one species from another.

**ggplot2** is part of the tidyverse package so we need to load that
first:

```{r, warning = FALSE}
# this assumes you've already installed tidyverse
library(tidyverse)
```

And recall that the **iris** dataset (3 species, 50 observations per
species) is automatically available to us:

```{r}
head(iris)
```

What is the correlation between petal length and width in these species?
Are longer petals also wider? We can visualize this with a scatterplot.
But first let's look a the ggplot template:

    ggplot(data = <DATA>) + 
      <GEOM_FUNCTION>(mapping = aes(<MAPPINGS>))

These are the obligatory parts of any plot. The first argument to
`ggplot` is the data frame:

```{r}
ggplot(data = iris)
```

This is not very interesting! but it's notable that it is *something*.
`ggplot()` has created a base coordinate system (a base layer) that we
can add visual layers to. The *add a layer* operator is "**+**", which
is the ggplot equivalent of the pipe symbol, and **it must occur at the
end of the line**.

The next argument specifies the kind plot we want: scatterplot, bar
chart, fitted line, boxplot, pie chart, etc. **ggplot2** refers to these as
**geoms**: the geometrical object that a plot uses to represent data.
You can see an overview of many of these geoms in the [cheat
sheet](https://github.com/rstudio/cheatsheets/blob/master/data-visualization-2.1.pdf).
The geom for a scatterplot is `geom_point()`.

But we also require a `mapping` argument, which maps the *variables* in
the dataset we want to focus on to their *visual representation* in the
plot. 

And finally we need to specify an **aesthetic** for the geometric
objects in the plot, which will control things like shape, color,
transparency, etc. Perhaps surprisingly, for a scatterplot, the x
and y coordinates are aesthetics, since these control,  not the shape or
color, but the relative position of the points in the coordinate system.

Here is our complete plot:

```{r}
ggplot(data = iris) +
  geom_point(mapping = aes(x = Petal.Length, y = Petal.Width))
```

There is clearly a positive correlation between length and width. And we
can make this even more apparent by visually fitting a line to the data, by
*overlaying* another geom in the same plot.

```{r}
ggplot(data = iris) +
  geom_point(mapping = aes(x = Petal.Length, y = Petal.Width)) +
  geom_smooth(mapping = aes(x = Petal.Length, y = Petal.Width))
```
There is clearly some code redundancy here, and we here we really don't want the mapping of these two layers to be independent. We can extract the common mapping information and move it to the top level:

```{r}
ggplot(data = iris, (mapping = aes(x = Petal.Length, y = Petal.Width))) +
  geom_point() +
  geom_smooth()
```
So we have the possibility of *local* layer specifications, and *global* specifications. Global specifications are *inherited* by all the local layers.

### The power of aesthetics

The aim of Fisher's paper was to try to discriminate different species based on their morphological measurements. It looks from this plot that there are two distinct clusters. Do these clusters correspond to different species? There are two clusters, but three species. How can we explore this further?

Our current plot uses two numeric variables: `Petal.Length` and `Petal.width`. We can add a third categorical variable, like `Species`, to a two dimensional scatterplot by mapping it to a different visual aesthetic. We've mapped length and width to x,y coordinates. Now we'll simultaneously map species to `color` by expanding our list of aesthetics:

```{r}
ggplot(data = iris) +
  (mapping = aes(x = Petal.Length, y = Petal.Width, color = Species)) +
  geom_point()
```

## Breakout Rooms

In the exercises we'll be looking a little more at the **iris** data, and in addition, the NHANES data we used last week, and the left-joined bird dataset we built last week in **Excercise 7**.

If you haven't installed the NHANES dataset do:
```{r}
install.packages("NHANES", repos = "http://cran.us.r-project.org")
```
Once installed, load it with:
```{r}
library(NHANES)
```

A prebuilt joined data set has been loaded on github.

```{r}
# create a data directory for the new file if you haven't done so yet:
dir.create('data/birds', recursive = TRUE)

# set the url
joined_data_url <- 'https://raw.githubusercontent.com/biodash/biodash.github.io/master/content/codeclub/04_ggplot2/joined_data.tsv'

# set the path for the downloaded file
joined_file <- 'data/birds/joined_data.tsv'

#download to file
download.file(url = joined_data_url, destfile = joined_file)

# read file
joined_data <- read_tsv(joined_file)

```



## Exercise 1

:::puzzle

Revisit the **iris** data set, and plot sepal width against sepal length colored by species. Which morphological character, petals or sepals, provides the greatest discrimination between species?

<details> <summary> Hints (click here) </summary>
Simply reuse the code we used for petals. You can often leverage code from an old plot for a new one.
<br> 
</details>

<details> <summary> Solution (click here) </summary>
```{r}
ggplot(data = iris) +
  (mapping = aes(x = Sepal.Length, y = Sepal.Width, color = Species)) +
  geom_point()
```

Note this solution shows yet another way to position global mapping information: as its own layer. This can help readability and avoid too many nested parentheses.
</details>
:::

------------------------------------------------------------------------

## Exercise 2

:::puzzle

Use the NHANES data set to plot body mass index against height. Color by gender. Which gender has the highest BMI?

<details> <summary> Hints (click here) </summary>
glimpse() the dataset to identify the variable names.
<br> 
</details>

<details> <summary> Solution (click here) </summary>
```{r}
ggplot(data = NHANES) +
  geom_point(mapping = (aes(x = Height, y = BMI, color = Gender)))
```
</details>
:::

------------------------------------------------------------------------

## Exercise 3

:::puzzle

Use the same plot but now color by physical activity. How active are those people with the highest BMI?

<details> <summary> Hints (click here) </summary>
Again, glimpse() the dataset to identify the variable names.
<br> 
</details>

<details> <summary> Solution (click here) </summary>
```{r}
ggplot(data = NHANES) +
  geom_point(mapping = (aes(x = Height, y = BMI, color = PhysActive)))
```
</details>
:::

------------------------------------------------------------------------

## Exercise 4

:::puzzle

Often plotting the data allows us to identify outliers, which may be data-entry errors, or genuinely extreme data. Using the `joined_data` set, plot adult body mass against longevity. Identify extreme data points at the high end of both scales. How can we identify what these points represent?

<details> <summary> Hints (click here) </summary>
Examine the plot to find an appropriate threshold value, and filter the data using that value. How many data points are there passing that threshold? What species are represented by these data points? How many weights are reported? Why is the plot misleading here?
<br> 
</details>

<details> <summary> Solution (click here) </summary>
```{r}
ggplot(data = joined_data) +
  geom_point(mapping = (aes(x = longevity_y, y = adult_body_mass_g)))
```

```{r}
joined_data %>% 
  filter(adult_body_mass_g > 10000)
```
```{r}
joined_data %>% 
  filter(adult_body_mass_g > 10000) %>% 
  select(species) %>% 
  distinct()
```
```{r}
joined_data %>% 
  filter(adult_body_mass_g > 10000) %>% 
  select(adult_body_mass_g) %>% 
  distinct()
```

</details>
:::

## Bonus, a new geom!

:::puzzle

Revisit the **iris** data and generate a density histogram for sepal length, categorized by species.

<details> <summary> Hints (click here) </summary>
Use geom_density(). Check the help to see what aesthetics it supports.
<br> 
</details>

<details> <summary> Solution (click here) </summary>

```{r}
ggplot(data = iris) +
  (mapping = (aes(x = Sepal.Length, fill = Species))) +
  geom_density(alpha = 0.5)
```
Note, what does the **alpha** aesthetic control?
<br>

</details>
:::
------------------------------------------------------------------------




<br> <br> <br> <br>
