---
title: "Session 4: Visualizing Data"
subtitle: "Introduction to Data Visualization with `ggplot2`"
summary: "In this session of Code Club, we'll look at how to visualize data in R using `ggplot2`."  
authors: [michael-broe]
date: "2020-12-10"
output: hugodown::md_document
toc: true
image: 
  caption: "Image from http://r-statistics.co/ggplot2-Tutorial-With-R.html"
  focal_point: ""
  preview_only: false
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

<br> <br> <br>

------------------------------------------------------------------------

## New To Code Club?

-   First, check out the [Code Club Computer Setup](/codeclub-setup/)
    instructions, which also has some pointers that might be helpful if
    you're new to R or RStudio.

-   Please open RStudio before Code Club to test things out -- if you
    run into issues, join the Zoom call early and we'll troubleshoot.

------------------------------------------------------------------------

## Session Goals

-   Learn the philosophy of **coding** a graphic.
-   Learn the basic **template** of a **ggplot2** graphic, so you can
    reuse it for multiple chart types.
-   Learn how you can quickly add visual information to a graphic using
    **aesthetics** and **layers**.

------------------------------------------------------------------------

## Intro: The ggplot2 philosophy

We have already seen that in R, instead of manually manipulating data
frames as you might do when editing Excel sheets, we **code** the
operations we want to perform using **dplyr** verbs like `select()`,
`mutate()`, `inner_join()`, and so on.

In a similar way when performing visualization, instead of clicking on a
chart type in Excel, we **code the chart** in R.

And just as **dplyr** gives us efficient ways to manipulate data frames,
**ggplot2** (which is also part of the **tidyverse**) gives us efficient
ways to manipulate charts/plots/graphics (we use these terms
interchangeably).

The **gg** in **ggplot2** stands for *grammar of graphics*, a systematic
approach for designing statistical plots developed by Leland Wilkinson.
The idea behind this was to think about 'pulling apart' various plots
into their shared component pieces, then provide code that could put
them together again. We can then create new plots like we create new
sentences (once we understand this grammar).

There are two parts to this. First, the 'nouns and verbs' we need to
work with plots are very different than those we need to work with data
frames. **ggplot2** is like a mini-language of its own, with its own
verbs and syntax.

Second, this notion of pulling apart a graphic leads to the idea of
*layers*. You can build up a plot of any complexity by *overlaying*
different views of the same data.

There's a learning curve here for sure, but there are a couple of things
that help us.

First, every graphic shares a *common template*. This is like thinking
about the sentence "The cat sat on the mat" grammatically as the
template `NP V PP` (`N`oun `P`hrase "The cat", `V`erb "sat",
`P`repositional `P`hrase "on the mat"). Once you understand this
structure you can "say" a *lot* of different things.

(And I mean a *lot*. The [ggplot cheat
sheet](https://github.com/rstudio/cheatsheets/blob/master/data-visualization-2.1.pdf)
gives an overview of the things you can do, but because this is a
language, users can create their own
[extensions](https://exts.ggplot2.tidyverse.org/gallery/) that you can
also utilize.)

Second, the way we put layers together is identical to the way we use
pipes. You can read `%>%` as "and then": `select()` and then `mutate()`
and then `summarize()`. In graphics, we can say "show this layer, and
then *overlay* this layer, and then *overlay* this layer", etc., using a
very similar syntax.

<br>

------------------------------------------------------------------------

## Examples

So how does this work in practice? We'll work through visualizing the
**iris** dataset that you've seen before. This is an extremely famous
[dataset](https://en.m.wikipedia.org/wiki/Iris_flower_data_set) that was
first analyzed by R. A. Fisher in 1936: *The use of multiple
measurements in taxonomic problems*. He was attempting to use the petal
and sepal measurements to discriminate one species from another.

**ggplot2** is part of the tidyverse package so we need to load that
first:

```{r, warning = FALSE}
# this assumes you've already installed tidyverse
library(tidyverse)
```

And recall that the **iris** dataset (3 species, 50 observations per
species) is automatically available to us:

```{r}
head(iris)
```

What is the correlation between petal length and width in these species?
Are longer petals also wider? We can visualize this with a scatterplot.
But first let's look a the ggplot template:

    ggplot(data = <DATA>) + 
      <GEOM_FUNCTION>(mapping = aes(<MAPPINGS>))

These are the obligatory parts of any plot. The first argument to
`ggplot` is the data frame:

```{r}
ggplot(data = iris)
```

This is not very interesting! but it's notable that it is *something*.
`ggplot()` has created a base coordinate system (a base layer) that we
can add visual layers to. The *add a layer* operator is "**+**", which
is the ggplot equivalent of the pipe symbol, and **it must occur at the
end of the line**.

The next argument specifies the kind plot we want: scatterplot, bar
chart, fitted line, boxplot, pie chart, etc. **ggplot2** refers to these as
**geoms**: the geometrical object that a plot uses to represent data.
You can see an overview of many of these geoms in the [cheat
sheet](https://github.com/rstudio/cheatsheets/blob/master/data-visualization-2.1.pdf).
The geom for a scatterplot is `geom_point()`.

But we also require a `mapping` argument, which maps the *variables* in
the dataset we want to focus on to their *visual representation* in the
plot. 

And finally we need to specify an **aesthetic** for the geometric
objects in the plot, which will control things like shape, color,
transparency, etc. Perhaps surprisingly, for a scatterplot, the x
and y coordinates are aesthetics, since these control,  not the shape or
color, but the relative position of the points in the coordinate system.

Here is our complete plot:

```{r}
ggplot(data = iris) +
  geom_point(mapping = aes(x = Petal.Length, y = Petal.Width))
```

There is clearly a positive correlation between length and width. And we
can make this even more apparent by visually fitting a line to the data, by
*overlaying* another geom in the same plot.

```{r}
ggplot(data = iris) +
  geom_point(mapping = aes(x = Petal.Length, y = Petal.Width)) +
  geom_smooth(mapping = aes(x = Petal.Length, y = Petal.Width))
```
There is clearly some code redundancy here, and we here we really don't want the mapping of these two layers to be independent. We can extract the common mapping information and move it to the top level.

```{r}
ggplot(data = iris) +
  (mapping = aes(x = Petal.Length, y = Petal.Width)) +
  geom_point() +
  geom_smooth()
```
So we have the possibility of *local* layer specifications, and *global* specifications. Global specifications are *inherited* by all the local layers.

### The power of aesthetics

The aim of Fisher's paper was to try to discriminate different species based on their morphological measurements. It looks from this plot that there are two distinct clusters. Do these clusters correspond to different species? There are two clusters, but three species. How can we explore this further?

Our current plot uses two numeric variables: `Petal.Length` and `Petal.width`. We can add a third variable, like `Species`, to a two dimensional scatterplot by mapping it to a different visual aesthetic. We've mapped length and width to x,y coordinates. Now we'll simultaneously map species to `color` by expanding our list of aesthetics:

```{r}
ggplot(data = iris) +
  (mapping = aes(x = Petal.Length, y = Petal.Width, color = Species)) +
  geom_point()
```


The National Health and Nutrition Examination Survey [(NHANES)
dataset](https://www.rdocumentation.org/packages/NHANES/versions/2.1.0/topics/NHANES)
contains survey data obtained annually from \~5,000 individuals on a
variety of health and lifestyle-related metrics. A subset of the data
are available as an R package - install and load it...

```{r}
install.packages("NHANES", repos = "http://cran.us.r-project.org")
library(NHANES)
```

Now preview the dataset...

```{r}
glimpse(NHANES)
```

To try out merging/joining, we'll create two separate data frames by
pulling out some variables from this NHANES dataset. One will contain
demographic variables, and the other with have some physical
measurements. Then we'll join them back together. Let's create the two
sub-datasets first...

```{r}
#Filter out rows with data from 2009-2010 and Age > 5, 
#select a subset (4) of the variables, then get rid of 
#all duplicate rows. Assign the output to object 'dem_data'.
dem_data <- NHANES %>% 
            filter(SurveyYr == "2009_10") %>% 
            filter(Age > 5) %>% 
            select(ID, Gender, Age, Education) %>% 
            distinct()

#similar as above, but with a different filter and 
#selecting different variables. Save as 'phys_data'
phys_data <- NHANES %>% 
             filter(SurveyYr == "2009_10") %>% 
             filter(Height < 180)  %>%
             select(ID, Height, BMI, Pulse) %>% 
             distinct()

```

Now explore them a bit...

```{r}
#view the first 6 rows of each - note the shared ID column
head(dem_data)
head(phys_data)

#preview in another way - note the different numbers of observations (rows)
glimpse(dem_data)
glimpse(phys_data)
```

```{r}
head(phys_data)
```

Let's use the shared ID column to join the two datasets together. We'll
do this in 4 different ways to compare different types of joins:
`inner_join()`, `left_join()`, `right_join()`, and `full_join()`. Pay
attention to the number of rows in the joined dataset each time and how
it relates to the number of rows in each of the two individual datasets.

The basic structure of the dplyr `*_join()` functions is...

`*_join(dataframe 'x', dataframe 'y', by = shared column name)`

<br>

------------------------------------------------------------------------

## 1 - `inner_join()`

```{r}

#preview the new object
head(join_inner)

```
```{r}
nhanes <- NHANES
```

```{r}
ggplot(data = NHANES) +
  geom_point(mapping = (aes(x = Height, y = BMI, color = PhysActive)))
```

```{r}
ggplot(data = NHANES) +
  geom_point(mapping = (aes(x = Height, y = BMI, color = Gender)))
```



## 2 - `left_join()`

```{r}
#perform an left join
join_left <- left_join(dem_data, phys_data, by = "ID")

#preview the new object
head(join_left)

#get dimensions
dim(join_left)
```

## 3 - `right_join()`

```{r}
#perform an right join
join_right <- right_join(dem_data, phys_data, by = "ID")

#preview the new object
head(join_right)

#get dimensions
dim(join_right)
```

## 4 - `full_join()`

```{r}
#perform an full join
join_full <- full_join(dem_data, phys_data, by = "ID")

#preview the new object
head(join_full)

#get dimensions
dim(join_full)
```

<br>

------------------------------------------------------------------------

# Breakout rooms

We're going to add to our backyard birds dataset. I found a dataset that
has life history data for a large number of species (birds and others).
We'll use species names to merge some of these life history variables in
to the occurrence data we already have.

If you're new and haven't yet gotten the backyard bird dataset, get it
first by running the code below. Otherwise, you can skip this step...

```{r}
# create a directory called data that contains a subdirectory called birds
dir.create('data/birds/', recursive = TRUE)

# set the location of the file
birds_file_url <-
'https://raw.githubusercontent.com/biodash/biodash.github.io/master/assets/data/birds/backyard-birds_Ohio.tsv'

# set the path for the downloaded file
birds_file <- 'data/birds/backyard-birds_Ohio.tsv'
#download
download.file(url = birds_file_url, destfile = birds_file)

```

Now **(everybody)**, read in the bird data for this session...

```{r}
birds_file <- 'data/birds/backyard-birds_Ohio.tsv'
birds <- read_tsv(birds_file)

```

## Exercise 1

<div class="alert puzzle">
<div>

Reduce the backyard bird dataset and keep just the following columns:
**species, locality, stateProvince, eventDate, species\_en**

<details> <summary> Hints (click here) </summary>

<br> Use `select()` to pull out the columns you want. <br>

</details>

<details> <summary> Solution (click here) </summary>

<br>

    birds <- birds %>% select(species, locality, stateProvince, eventDate, species_en)

</details>

</div>
</div>

------------------------------------------------------------------------

## Exercise 2

<div class="alert puzzle">
<div>

Check to make sure things look right - how many columns does the birds
dataset now have?

<details> <summary> Hints (click here) </summary>

<br> Use the `dim()` function. Or the `ncol()` function. Or `glimpse()`. Or
`head()`. Or `str()`. Or even `summary()`. There's lots of ways to do this.
<br> </details>

<details> <summary> Solution (click here) </summary>

<br>

```{r}
dim(birds)
```

</details>

</div>
</div>

------------------------------------------------------------------------

## Exercise 3

<div class="alert puzzle">
<div>

Now download and read in the new life history dataset (tab separated)
available at
<https://github.com/biodash/biodash.github.io/raw/master/assets/data/birds/esa_life_history_data_cc.tsv>.
Then explore it a bit - how many rows and columns are there?

<details> <summary> Hints (click here) </summary>

<br> Use the `download.file()` function like we did previously for the
bird dataset. You'll need to define the arguments 'url' and 'destfile'
inside the parentheses. You can put the file anywhere you want, but I'd
suggest in the same directory as the bird file we got, so, for example, the
destination file could be "data/birds/life_history_data.tsv". <br>
</details>

<details> <summary> Solution (click here) </summary>

<br>

```{r}
#download the file from online and save it as a '.tsv' file (since it's tab delimited)
download.file(url = "https://github.com/biodash/biodash.github.io/raw/master/assets/data/birds/esa_life_history_data_cc.tsv",
destfile = "data/birds/life_history_data.tsv")

#read the data in to R as an object named 'life_hist'
life_hist <- read_tsv(file = "data/birds/life_history_data.tsv")

#preview the data
glimpse(life_hist)
```

</details>

</div>
</div>

------------------------------------------------------------------------

## Exercise 4

<div class="alert puzzle">
<div>

This new dataset contains life history data for more than just birds.
What Classes of organisms are represented in the 'Class' variable?

<details> <summary> Hints (click here) </summary>

<br> Try using a combination of the `select()` and `distinct()` functions to pull out the column you're interested in, and then to get the distinct values, respectively.
<br> </details>

<details> <summary> Solutions (click here) </summary>

```{r}
life_hist %>% select(class) %>% distinct()
```

</details> <br>

</div>
</div>

------------------------------------------------------------------------

## Exercise 5

<div class="alert puzzle">
<div>

Reduce the life history dataset down to keep just the rows for Class
**Aves** and the columns **species, adult_body_mass_g,
adult_svl_cm, longevity_y, litter_or_clutch_size_n**. What are the dimensions now?

<details> <summary> Hints (click here) </summary>

Use `filter()` along with an appropriate logical expression to keep the
rows we want. Use `select()` to get the desired columns. <br> </details>

<details> <summary> Solutions (click here) </summary>

```{r}
# pull out target rows and columns
life_hist_aves <- life_hist %>% 
    				      filter(class == "Aves") %>% 
    				      select(species, adult_body_mass_g, adult_svl_cm, longevity_y, litter_or_clutch_size_n)

dim(life_hist_aves)
```

</details> <br>

</div>
</div>

------------------------------------------------------------------------

## Exercise 6

<div class="alert puzzle">
<div>

Preview each dataset again, just to make sure you're clear about what's
in each one. Are there any columns that are shared between the two?

<details> <summary> Hints (click here) </summary>

Consider `glimpse()` or `head()` to preview the datasets (tibbles/data frames).
If you want to use a function to find shared columns, try a combination of `intersect()` and `names()`

</details>

<details> <summary> Solutions (click here) </summary>

```{r}
glimpse(birds)

glimpse(life_hist_aves)

intersect(names(birds), names(life_hist_aves))
```

</details> <br>

</div>
</div>

------------------------------------------------------------------------

## Exercise 7

<div class="alert puzzle">
<div>

Now lets join them together based on their shared variable. Not all
species in the backyard bird (Ohio) dataset are included in the life
history dataset. Likewise, there are life history data for many species
that aren't in the Ohio dataset. We want to keep all the Ohio
observations, and merge in life history data for species where it's
availble, but we also don't want to add in life history data for species
that aren't in the Ohio dataset. Choose an appropriate join function
with those things in mind.

<details> <summary> Hints (click here) </summary>

Try a `left_join()`, defining the Ohio backyard bird dataset as the 'x'
dataset in the join and the life history data as the 'y' dataset. Get
details on that function with `?left_join`. <br>

</details>

<details> <summary> Solutions (click here) </summary>

```{r}
joined_data <- left_join(x = birds, y = life_hist_aves, by = "species")
```

</details> <br>

</div>
</div>

------------------------------------------------------------------------

## Exercise 8

<div class="alert puzzle">
<div>

What are the longest- and shortest-living bird species in Ohio based on
the data in the longevity\_y column?

<details> <summary> Hints (click here) </summary> Try using `select()` to pull out just
the columns species and longevity_y, then use `distinct()` to get the unique
rows, then `arrange()` based on the longevity_y column. You might also
find the dplyr function `desc()` helpful.

Alternatively, you could try grouping by species, then use summarise() to get
either the max, min, or mean value for longevity_y for each species
(there's just one value for each species, so all of those statistics
give the same value in this case). Then sort (arrange) the resulting
summarized data frame on the longevity value.

<br> </details>

<details> <summary> Solutions (click here) </summary>

```{r}
#option 1 - shortest-lived birds
joined_data %>% select(species, longevity_y) %>% 
                distinct() %>% 
                arrange(longevity_y)

#option 1 - longest-lived birds
joined_data %>% select(species, longevity_y) %>% 
    				    distinct() %>% 
    				    arrange(desc(longevity_y))

#option 2 - shortest-lived birds
joined_data %>% group_by(species) %>% 
                summarise(longevity = max(longevity_y)) %>% 
                arrange(longevity)
    
#option 2 - longest-lived birds
joined_data %>% group_by(species) %>% 
                summarise(longevity = max(longevity_y)) %>% 
                arrange(desc(longevity))
    
```

</details> <br>

</div>
</div>

------------------------------------------------------------------------

## Bonus time!

## Bonus 1

<div class="alert puzzle">
<div>

What species in Ohio has the largest ratio of adult body mass to
length (measured as snout vent length, or 'adult_svl_cm')?

<details> <summary> Hints (click here) </summary>

Use mutate() to create a new variable containing the body mass divided by svl, then
arrange the dataset using that new variable to get the species with the highest value.

</details>

<details> <summary> Solutions (click here) </summary>

```{r}
joined_data %>% mutate(ratio = adult_body_mass_g/adult_svl_cm) %>% 
                select(species_en, ratio) %>% distinct() %>% 
                arrange(desc(ratio))
```

</details> <br>

</div>
</div>

------------------------------------------------------------------------

### Bonus 2

<div class="alert puzzle">
<div>
There are 2 additional joins we haven't talked about - semi_join() and anti_join(). Take a look at the documentation to see what these do. Use one of them to find what species in the backyard birds dataset are not in the life history dataset.

<details>
  <summary>
  Hints (click here)
  </summary>
  
  Use anti_join() and distinct().

</details>

<details>
  <summary>
  Solutions (click here)
  </summary>
  
```{r}
anti_join(birds, life_hist_aves, by = "species") %>% select(species, species_en) %>%
                                                     distinct()

```

</details>

</div>
</div>

------------------------------------------------------------------------

## Bonus 3

<div class="alert puzzle">
<div>

The life history dataset we downloaded above is actually a modified
version of the original file, which is located at
'<http://www.esapubs.org/archive/ecol/E096/269/Data_Files/Amniote_Database_Aug_2015.csv>'

Try starting with the original file and repeating what we did above -
merging the variables **species, adult_body_mass\_g, adult_svl_cm, longevity_y, litter_or_clutch_size_n** in to the original birds
dataset. First, make sure to get it read in correctly. Then pay
attention to the species column in the life history dataset - what needs
to be done before a join/merge can be performed?

<details> <summary> Hints (click here) </summary>

Pay attention to how missing data are coded in this dataset (it's -999).
Also, data are very sparse for some of the variables - in other words,
they have lots of missing data. This seems to cause a problem with the
read\_csv function, as it only considers the first 1000 rows for the
purpose of defining the class of each column. This can be a problem if
all of the first 1000 rows are missing. Finally, it appears that even
though this is a comma separated file (commas define the column breaks),
there are a few instances where commas are used within a field. This
happens in the 'common name' column in a few cases where multiple common
names are listed for a specific observation. This is one example of
something that can become quite frustrating when trying to get data
loaded in, and is worth keeping an eye out for. Fortunately, in our case, it only seems to happen for non-bird species in this dataset, which we filter out anyway, so it can be dealt with. However, if it had impacted any of the bird observations, I think fixing this might require a solution outside of R - possibly a command line approach.

<br>

</details>

<details> <summary> Solutions (click here) </summary>

```{r}
#download
download.file(url = "http://www.esapubs.org/archive/ecol/E096/269/Data_Files/Amniote_Database_Aug_2015.csv", 
destfile = "data/birds/orig_life_history.csv")

#read the data in to R as an object named 'full_life_hist'

full_life_hist <- read_csv("data/birds/orig_life_history.csv", 
                  na = "-999", 
                  col_types = cols(birth_or_hatching_svl_cm = col_double(),
                  weaning_d = col_double(),gestation_d = col_double(), 
                  weaning_weight_g = col_double(), 
                  male_svl_cm = col_double(), 
                  female_svl_cm = col_double(),
                  no_sex_svl_cm = col_double(), 
                  female_body_mass_at_maturity_g = col_double(),
                  female_svl_at_maturity_cm = col_double()))

#get the original version of the birds dataset
birds <- read_tsv('data/birds/backyard-birds_Ohio.tsv')

#subset each for the columns and rows we want
life_hist_aves <- full_life_hist %>% filter(class == "Aves") %>% 
                  select(species, adult_body_mass_g, adult_svl_cm, longevity_y, litter_or_clutch_size_n)

birds <- birds %>% select(species, locality, stateProvince, eventDate, species_en)

glimpse(birds)

glimpse(life_hist_aves)

#notice the species column in the life history data doesn't include the genus name. Since the names don't match in the species column from each dataset, a join won't work. Add the genus variable in from the original life history data...

life_hist_aves <- full_life_hist %>% filter(class == "Aves") %>% 
                  select(genus, species, adult_body_mass_g, adult_svl_cm, longevity_y, litter_or_clutch_size_n)

#now use mutate to replace the species column so it includes both the genus and species...

life_hist_aves <- life_hist_aves %>% mutate(species = paste0(genus, " ", species)) %>% select(-genus)

#preview again
glimpse(birds)
glimpse(life_hist_aves)

#now we can join
joined_data <- left_join(birds, life_hist_aves, by = "species")
```

</details> <br>

</div>
</div>

------------------------------------------------------------------------

<br> <br> <br> <br>
