---
output: hugodown::md_document
title: "S04E15: R for Data Science - Exploratory Data Analysis"
subtitle: "Chapter 7.3 - 7.4: Plotting the distributions of categorical and continuous variables"
summary: "This chapter covers so-called Exploratory Data Analysis (EDA): computing summary stats and especially making quick plots to explore the variation in and distributions of single variables (this session), and looking at covariation among variables (next session)."
authors: [admin]
tags: [codeclub, r4ds]
date: 2022-11-15
lastmod: 2022-11-15
toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE, message = FALSE, cache = FALSE)
```

--------------------------------------------------------------------------------

## Setting up

To start with, we'll only need to load the _tidyverse_,
as we'll explore a dataset that is automatically loaded along with it.

```{r}
## You only need to install if you haven't previously done so
# install.packages("tidyverse")

library(tidyverse)
```

We'll be working with the `diamonds` dataset today,
so let's take a quick look at it before we begin:

```{r}
diamonds
```

On each row, we have information about one individual diamond,
such as its `carat` and `price`. 
(`x`, `y`, and `z` represent the diamond's length, width, and depth,
respectively.)

Since we'll be making a bunch of plots with _ggplot2_,
let's use the following trick to set an overarching "theme" for all plots
that is a little better-looking than the default one:

```{r}
# This changes two things:
# - theme_minimal() gives an overall different look, with a white background
# - base_size = 14 will make the text relatively bigger
theme_set(theme_minimal(base_size = 14))
```

<br>

--------------------------------------------------------------------------------

## Chapter 7.3: Variation

### Exploring variation in a categorical variable

Let's say we want to see how many diamonds there are for each value of `cut`.
When we printed the first lines of the dataframe above,
we could see that `cut` has values like `Ideal`, `Premium`, and `Good`:
this is therefore a "categorical" and not a "continuous" variable.

We could also see that the data type indication for `cut` was `<ord>`,
which is short for _ordered factor_.
In R, categorical variables can be represented not just as character strings
or integers, but also as **factors**.
Factors have a defined set of _levels_ which can be assigned a custom order.
That is handy when plotting or when you need to set a reference level
in a statistical model.
(For more,
see [the page for this previous Code Club session on factors](/codeclub/06_factors/).)

To quickly see which values the variable `cut` contains,
and what their frequencies are,
we can use `count()`:

```{r}
diamonds %>% count(cut)
```

To get a feel for the distribution of a categorical variable,
making a barplot can also be useful.
Recall that when making a plot with _ggplot2_,
we at least need the following components:

- The `ggplot()` function, in which we supply the dataframe that we
  want to use.
  
- A _geom_ function, which is basically the type of plot we want to make,
  such as `geom_point()` for a scatterplot and `geom_bar()` for a barplot.

- An "aesthetic mapping" that defines which variables to plot along the axes.

For a barplot showing `cut`, our _ggplot2_ code would look as follows:

```{r}
ggplot(data = diamonds) +
  geom_bar(mapping = aes(x = cut))
```

When making plots, we typically specify which variable should go along the y-axis,
too.
But that is not the case for barplots,
where the default is to automatically 
plot a _count which is **computed** from the data_.

<br>

### Exploring variation in a continuous variable

We'll take another look at the `diamonds` dataframe and pick a _continuous_
variable:

```{r}
diamonds
```

Let's explore the variation in the continuous variable `carat`,
and do so by making a histogram using `geom_histogram()`:

```{r}
ggplot(data = diamonds) +
  geom_histogram(mapping = aes(x = carat))
```

Under the hood, a histogram discretizes the continuous data into bins,
and then shows the counts (here: number of diamonds) in each bin.
We may want to play around with the width of the bins
to see more fine-grained or coarse-grained patterns,
and can do so using the `binwidth` argument:

```{r}
ggplot(data = diamonds) +
  geom_histogram(mapping = aes(x = carat), binwidth = 0.5)
```

If we wanted to see this kind of representation in table-form,
using `count()` directly wouldn't work:
we don't have a column with bins for `carat`, only the raw, numeric values.

To create bins, we can use the _ggplot2_ function `cut_width`,
whose `width` argument is equivalent to `geom_histogram`'s `binwidth`:

```{r}
diamonds %>%
  mutate(carat_discrete = cut_width(carat, width = 0.5)) %>%
  count(carat_discrete)
```

If we want to show the variation in `carat` _separately for each level of `cut`_,
we can map `carate` _also_ to `fill`, which is the fill color of the bars:

```{r}
## First, let's subset the data to only keep relatively small diamonds: 
smaller <- diamonds %>% filter(carat < 3)

## Then, we make the plot:
ggplot(data = smaller,
       # The mapping is now inside ggplot, and we use `cut` twice:
       mapping = aes(x = carat, fill = cut)) +
  geom_histogram(binwidth = 0.1, color = "grey20")
```

Though in a case like this,
a linegraph with `geom_freqpoly()` might be easier to interpret:

```{r}
ggplot(data = smaller,
       mapping = aes(x = carat, colour = cut)) +
  geom_freqpoly(binwidth = 0.1, size = 1.5) # (Making thicker lines with 'size')
```

### Unusual values

Sometimes, plots like histograms have very wide axis limits
yet no visible bars on the sides:

```{r}
ggplot(diamonds) + 
  geom_histogram(mapping = aes(x = y), binwidth = 0.5)
```

The x-axis limits are automatically picked based on the data,
so there really _should_ be some values all the way up to about 60.
We just can't see them,
since the y-axis scale goes all the way up to 12 thousand.

If we want to see these counts in the graph,
we can _zoom in_ on the y-axis with `coord_cartesian()`:

```{r}
ggplot(diamonds) + 
  geom_histogram(mapping = aes(x = y), binwidth = 0.5) +
  coord_cartesian(ylim = c(0, 50)) # c(<lower-limit>, <upper-limit>)
```

Note that in _ggplot2_,
zooming in on a graph and setting axis limits isn't the same thing:
you'll learn more about that in the exercises.

Of course we could also try to find these values in the dataframe itself,
which might be more useful than a graph in cases like this.
To do so, we can use the `filter()` function we learned about in the previous
chapter:

```{r}
diamonds %>% filter(y < 3 | y > 20)
```

<br>

--------------------------------------------------------------------------------

## Breakout Rooms

These exercises will continue to use the `diamonds` data,
which is automatically loaded when you load the _tidyverse_.

<div class="puzzle">
<div>

### Exercise 1

In the `diamonds` data, explore the distribution of `price`,
which is the price of a diamond in USD.
Do you discover anything unusual or surprising?

Make sure to try different values for the `binwidth` argument!

<details><summary><b>Hints</b> (click here)</summary>
<br>

- This is a continuous variable, so use `geom_histogram()`.

- A more fine-grained plot (smaller bins with `binwidth`)
  than the default should reveal something odd.
  
- You might want to use `coord_cartesian()` to see the area with the odd
  pattern in more detail.
  (Alternatively, you could try `filter()`ing the data before plotting.)

</details>

<br>

<details><summary><b>Solution</b> (click here)</summary>
<br>

- `geom_histogram()` with default settings doesn't reveal anything too weird:

```{r}
ggplot(data = diamonds,
       mapping = aes(x = price)) +
  geom_histogram()
```

- But with a binwidth of e.g. 100, we start to see something odd:
  a gap in the distribution.

```{r}
ggplot(data = diamonds,
       mapping = aes(x = price)) +
  geom_histogram(binwidth = 100)
```

- Let's take a closer look by zooming in on prices of $2,500 or less:

```{r}
diamonds %>%
  ggplot(mapping = aes(x = price)) +
  geom_histogram(binwidth = 25) +
  coord_cartesian(xlim = c(0, 2500))
```

(An alternative approach would be to filter the data before plotting:)

```{r, eval=FALSE}
diamonds %>%
  filter(price < 2500) %>% 
  ggplot(mapping = aes(x = price)) +
  geom_histogram(binwidth = 25)
```

I have no idea why there are no diamonds with a price of around $1,500 --
anybody?

</details>
</div></div>

<br>

<div class="puzzle">
<div>

### Exercise 2

Compare `coord_cartesian()` with the similar function `lims()` to 
see a narrower range along the y-axis for a histogram.
Specifically, make two histograms of `price` with a y-axis that only goes
up to 3,000:
one with `coord_cartesian(ylim = ...)` and one with `lims(y = ...)`.

What is happening in the graph made with `lims()`?

(See the hint for example usage of `lims()`, a function we haven't seen yet.

<details><summary><b>Hints</b> (click here)</summary>
<br>

You can use `lims()` to set arbitrary axis limits:

```{r}
ggplot(diamonds) +
  geom_point(mapping = aes(x = x, y = y)) +
  lims(x = c(5, 10),   # c(<lower-limit>, <upper-limit>)
       y = c(0, 20))   # c(<lower-limit>, <upper-limit>)
```

You could also use the very similar `ylim()` / `xlim()` pair of functions,
though note the slightly simplified syntax:

```{r}
ggplot(diamonds) +
  geom_point(mapping = aes(x = x, y = y)) +
  xlim(5, 10) +   # Note: you don't pass a vector inside 'c()' here
  ylim(0, 20)
```

</details>

<br>

<details><summary><b>Solution</b> (click here)</summary>
<br>

Whereas the graph produced with `coord_cartesian()` is simply "cut off" at
the specified limit,
the graph produced with `lims()` is missing bars!

It turns out that _ggplot2_ removes the bars that can't be shown given our y-limit.
Notice that it warns us about doing so:
`#> Warning: Removed 5 rows containing missing values (geom_bar).`

```{r}
ggplot(diamonds) +
  geom_histogram(mapping = aes(x = price)) +
  coord_cartesian(ylim = c(0, 3000))
```

```{r}
ggplot(diamonds) +
  geom_histogram(mapping = aes(x = price)) +
  lims(y = c(0, 3000))
```

</details>
</div></div>

<br>

<div class="puzzle">
<div>

### Exercise 3

Using scatterplots,
explore the relationship between the depth `y` and the width `z` of the diamonds.

What do you think about the outliers?
Are they more likely to be unusual diamonds or data entry errors?

<details><summary><b>Hints</b> (click here)</summary>
<br>

- Make a scatterplot with `geom_point()`.

- Zoom in on the area with most points,
  to get a better feel for the overall relationship between `y` and `z`.

- Could a diamond with a value of `y` larger than 20 just be a very large diamond?
  Or does the corresponding value for `z`,
  and the overall relationship between `y` and `z` make it more likely that
  they are outliers?

</details>

<br>

<details><summary><b>Solution</b> (click here)</summary>
<br>

Let's start with a simple scatterplot with all data and default axis limits:

```{r}
ggplot(data = diamonds,
       mapping = aes(x = z, y = y)) +
  geom_point()
```

Phew! There are definitely some striking outliers.
Let's zoom in on the main cloud of points:

```{r}
ggplot(data = diamonds,
       mapping = aes(x = z, y = y)) +
  geom_point() +
  coord_cartesian(xlim = c(0, 10), ylim = c(0, 15))
```

That looks like an overall very tight correlation between
width (`y`) and depth (`z`).

Therefore,
the outliers of `y` and `z` don't just seem to represent very large or very small 
diamonds, and are likely data entry errors or something along those lines.

</details>
</div></div>

<br>

--------------------------------------------------------------------------------

## Chapter 7.4: Missing values

### Removing outliers

If you've established that certain outliers are untrustworthy and
want to get rid of them,
you have two main options.

First, you could remove the rows (diamonds) with outliers:

```{r}
diamonds2 <- diamonds %>% filter(y > 3, y < 20)
```

But you may not want throw out entire rows,
because the values for the other variables might be valid and still valuable.
Alternatively, you can convert outliers to `NA`s (missing values),
and a convenient way to do that is with the `ifelse()` function.
To understand `ifelse()`, a simple example may help:

```{r}
# Create a vector with integers from 1 to 10:
x <- 1:10
x
```

```{r}
# This will return a logical vector, indicating, for each value,
# whether it is smaller than 5:
x < 5
```

```{r}
# We can turn the small values into NAs, and leave big values unchanged, as follows:
ifelse(test = x < 5, yes = NA, no = x)
```

In case of the diamonds,
we can use this function as follows to turn `y` outliers into `NA`s:

```{r}
diamonds2 <- diamonds %>%
  mutate(y = ifelse(test = y < 3 | y > 20, yes = NA, no = y))
```

### Comparing observations with and without missing data 

Sometimes you may want to compare distributions among observations with and
without missing values.
To do that, we can create a new variable that indicates whether a value is
missing or not, and map (for example) `color` to this variable: 

```{r}
library(nycflights13)

flights %>% 
  mutate(cancelled = is.na(dep_time)) %>% 
  ggplot(mapping = aes(x = sched_dep_time, color = cancelled)) +
  geom_freqpoly(binwidth = 100) # (100 = 1 hour, so we plot by hour)
```

However, the large difference in the absolute counts of cancelled versus
not-cancelled flights makes it hard to see relative differences along the
x-axis.

We can use `geom_density()` to produce a density plot,
where the height of the lines is only determined by the relative counts,
and where we can see if cancelled flights have a different distribution:

```{r}
flights %>% 
  mutate(cancelled = is.na(dep_time)) %>% 
  ggplot(mapping = aes(x = sched_dep_time, color = cancelled)) + 
  geom_density()
```

It looks like flights at the end of the day are much more commonly cancelled
than those early on, which is what we might have expected!

<br>

<div class="puzzle">
<div>

### Exercise 4 (bonus)

Explore the distribution of `carat`.
Specifically, compare the number of diamonds of 0.99 (and a little less) carat and
those of 1 (and a little more) carat?
What do you think is the cause of the difference?

<details><summary><b>Hints</b> (click here)</summary>
<br>

- Make a histogram (`geom_histogram()`) for `carat`,
  and optionally zoom in to the area around 1.

- Use `filter()` and `count()` to specifically check out the diamond counts
  with a carat of around 1.

</details>

<br>

<details><summary><b>Solution</b> (click here)</summary>
<br>

We can start by simply making a histogram for `carat`:

```{r}
ggplot(data = diamonds,
       mapping = aes(x = carat)) +
  geom_histogram(binwidth = 0.01)
```

That's a weird pattern, with a bunch of peaks and valleys!
Let's just show the area around a carat of `1`:

```{r}
diamonds %>%
  filter(carat > 0.9, carat < 1.1) %>%
  ggplot(mapping = aes(x = carat)) +
  geom_histogram(binwidth = 0.01)
```

There's clearly a big uptick around `1`,
but checking out the raw counts would make it easier to answer the original question:

```{r}
diamonds %>%
  filter(carat > 0.9, carat < 1.1) %>%
  count(carat)
```

There are suspiciously few diamonds with a carat of 0.99
(and, to a lesser extent, with a carat anywhere above 0.9):
could there be some rounding-up going on? 

</details>
</div></div>

