---
title: "Session S03E06: Data structures and subsetting"
subtitle: "Overview of data structures and how to access them"
summary: "In this session of Code Club, we'll move below and beyond the tidyverse to get an overview of accessing various kinds of data structure in R."  
authors: [michael-broe]
date: "2022-02-15"
output: hugodown::md_document
toc: true
image: 
  caption: "Image from https://r4ds.had.co.nz"
  focal_point: ""
  preview_only: false
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

------------------------------------------------------------------------

## New To Code Club?

-   First, check out the [Code Club Computer Setup](/codeclub-setup/)
    instructions, which also has some pointers that might be helpful if
    you're new to R or RStudio.

-   Please open RStudio before Code Club to test things out -- if you
    run into issues, join the Zoom call early and we'll troubleshoot.

------------------------------------------------------------------------

## Session Goals

-   Learn the uses of R's three basic subsetting operators: `[ ]`,
    `[[ ]]`, and `$`.
-   Learn how the behavior of these operators varies depending on the
    **data structure** you are subsetting (vector, list, or data frame).
-   Learn the value of the `str()` command.
-   Learn how these base-R operators relate to tidyverse commands.

------------------------------------------------------------------------

## Intro: What is 'subsetting' anyway?

Subsetting (also known as indexing) is simply a way of using base-R syntax to extract specific pieces of a data structure. 

We've already seen two **dplyr**
verbs that perform this kind of operation: `filter` (to extract
specific rows) and `select` (to extract specific columns). 

But these are tidyverse commands, and only work with data frames.

R has two more basic data structures, from which everything else is built, **vectors** and **lists**, and for
these we need different subsetting operators (dplyr functions won't work). We'll also see that data
frames are just a special kind of list, and that base-R subsetting
operators also work for these, which can often be useful and efficient.

In your R experience you will almost
certainly come across both code and output which does not adhere to tidyverse
conventions. We have already come across outputs which are *not* data frames in 
[SO3E01: t-tests](https://biodash.github.io/codeclub/s03e01_ttests/)
and [S03E02: ANOVA](https://biodash.github.io/codeclub/s03e02_anova/)
which we will revisit below.

Since the behavior of these operators depends on the actual data
structure you are working with, it's useful when experimenting to use
them in conjunction with the `str()` function, which compactly displays
the internal structure of an any R object. A knowledge of the make-up of
these data structures is also important when you come to write your own
loops, iterations, and functions.

The most important distinction between vectors and lists is within
vectors *every value must be of the same type*: for example, all
characters, or all integers, etc. Inside lists, you can *mix values of
any type*.

In addition, a list is best thought of as a general purpose container,
which can contain not just mixed values, but entire vectors of any type. 

Since we'll be comparing base-R with tidyverse functions, we need to
load the tidyverse, and we'll also be using the `palmerpenguins` package to reproduce our previous ANOVA results:

```{r, warning = FALSE, message = FALSE}
library(tidyverse)
library(palmerpenguins)
```

------------------------------------------------------------------------

## Vectors

A vector is absolutely the most basic data structure in R. Every value
in a vector must be of the **same type**.

There are four basic **types** of vector: integer, double, character,
and logical. Vectors are created by hand with the `c()` (combine,
concatenate) function. We can check the type with the `typeof()`
operator. This is totally redundant if you just created the vector
yourself, but when you are debugging code, or creating a vector using an
expression, you might want to check exactly what type of vector is being
used:

```{r}
vec_dbl <- c(1, 2, 3, 4, 5)
vec_dbl
typeof(vec_dbl)
```

```{r}
vec_seq <- seq(1, 5)
vec_seq
typeof(vec_seq)
```

```{r}
vec_which <- 1:5
vec_which
typeof(vec_which)
```

```{r}
vec_chr <- c("a", "b", "c", "d", "e")
vec_chr
typeof(vec_chr)
```

Vectors have an insanely simple **str**ucture:

```{r}
str(vec_dbl)
```

`str()` also displays the type, and RStudio displays the result of
`str()` in the Values pane. (Note that 'double' and 'num(eric)' mean exactly the same thing in R.)

For such a simple structure, there are a surprisingly large number of
ways to subset a vector. We'll just look a a small sample here, and use the following example:

```{r}
x <- c(2.1, 4.2, 3.3, 5.4)
x
str(x)
```

(Notice for this example the *number after the decimal point* indicates
the position (index) of the element of the vector.)

**Positive integers** return elements at the specified positions. Any
expression that evaluates to a vector of positions can be used as the
index. The index operator is `[ ]`:

```{r}
x[3]
x[c(3, 1)]
x[2:4]
```

**Negative integers** *exclude* elements at the specified positions:

```{r}
x[-3]
x[c(-3, -1)]
```

The bottom line here is that each value in a vector has an implicit index (position), and we can use that index to pull out values of interest. This can be extremely useful when writing for-loops that move through a vector accessing one value at a time.

**Attributes**. One of the unusual features of R as opposed to other
programming languages is that you can assign metadata of various kinds
to the elements of vectors (and lists). For example, we can assign a
**name** to each element, and then use a **character vector** as the
index expression:

```{r}
y <- c("a" = 2.1, "b" = 4.2, "c" = 3.3, "d" = 5.4)
str(y)
y[c("d", "c", "a")]
```

The `str()` command now shows us that we now have a 'Named'
numeric vector, and that we have a "names" attribute, which is itself a
character vector.

### Exercise 1

:::puzzle

Consider the words "yellow", "red", and "green".

Create a numeric vector called "lengths" which simply shows the *length of the words* (in that order).

Look at its structure using `str()`.

Extract the first and last elements of this vector, indexing by position.

Then, create a second vector called "named_lengths", with the same word-lengths, but now also using a corresponding names attribute: "yellow", "red" and "green".

Look at its structure using `str()`.

Again, extract the first and last elements, but now using a character vector as the index.

<details> <summary> Solution (click here) </summary>

```{r}
lengths <- c(6, 3, 5)
str(lengths)
lengths[c(1, 3)]
```

```{r}
named_lengths <- c("yellow" = 6, "red" = 3, "green" = 5)
str(named_lengths)
named_lengths[c("yellow", "green")]
```

<br>


</details>

:::

------------------------------------------------------------------------

## Lists

There are two main differences between vectors and lists: (i) lists can
contain elements of **different types**; and (ii) lists can contain **entire vectors** as elements (and even 
other lists: which is why lists are sometimes referred to as
**recursive**: it can be lists of lists of lists, 'all the way down'.
This is a topic for another day!). 

Let's directly compare the structure
of a *list* of numbers to a *vector* of numbers. Just as we create
vectors by hand with the `c()` function, we create lists with the
`list()` function.

```{r}
l <- list(2.1, 4.2, 3.3, 5.4)
l
str(l)
```

Notice the difference between *printing* a list (all those brackets!!)
and using the `str()` command, which is much more compact and readable.

What if we mix values of different types?

```{r}
l_mixed <- list(2.1, 2L, T, "a")
l_mixed
str(l_mixed)
```


Things get more interesting when we create a **list of vectors**:

```{r}
mixed_vectors <- list(c("kim", "sandy", "lee"), c(23, 21, 26))
mixed_vectors
str(mixed_vectors)
```

In these examples we  see the appearance of a new subsetting operator
`[[ ]]`, in addition to `[ ]`. How do they differ? Let's experiment,
focussing on the second element of the list `c(23, 21, 26)`. Let's try to pull out that vector using the `[2]` notation (it is the second element after all).

```{r}
mixed_vectors_subset_a <- mixed_vectors[2]
mixed_vectors_subset_a
str(mixed_vectors_subset_a)
```

**This does not pull out the vector!!** Instead, it returns a
**sublist** which *contains* that vector as the only element (we'll see
why R does this below...).

So how to we get our hands on the actual vector? This is where the
`[[ ]]` operator comes in:

```{r}
mixed_vectors_subset_b <- mixed_vectors[[2]]
mixed_vectors_subset_b
str(mixed_vectors_subset_b)
```

We see that the behavior of the `[ ]` operator is very different for lists: it
selects the element(s) you request, but *always still wrapped inside a list*. It 'shrinks' the original list. The
`[[ ]]` operator on the other hand 'drills-down' and just returns the
'un-listed' vector in that position.

## Data frames

The reason R does things this way is because data frames are so central
to the language. 
Let's build a data frame from the ground up to see how
it works.

Basically the "inside" of a data frame is just a list with name
attributes:

```{r}
attr_list <- list(name = c("kim", "sandy", "lee"), age = c(23, 21, 26))
attr_list
str(attr_list)
```

Notice that instead of `str()` displaying `$ : ...` for each entry, we
now see attributes `$ name: ...` , `$ age: ...` for each entry. Also note that all
those double `[[ ]]` notations have disappeared when we print. This
should give you a clue that `$age`, for example, is a kind of alias for
`[[2]]`.

Finally, we can 'wrap' this into an official data frame structure:

```{r}
my_df <- as.data.frame(attr_list)
my_df
str(my_df)
```

Or wrap it into a tidyverse tibble:

```{r}
my_tibble <- as_tibble(attr_list)
my_tibble
str(my_tibble)
```

So a data frame is basically a list (intepreted as columns), with a names attribute
for the columns (interpreted as headers). And with the extra condition that all the columns
are of the same length, so it's rectangular. So we should be able to
use our standard list subsetting operators on it:

```{r}
col_2 <- my_df[2]
str(col_2)
```

Since a data frame is a list, subsetting using `[ ]` returns the
specified column *still inside a data frame*. What about `[[ ]]`?

```{r}
vec_2 <- my_df[[2]]
str(vec_2)
```

Using `[[ ]]` pulls out the data vector from the column.

Just like vectors, we can also subset a data frame by the name
attribute, instead of by position:

```{r}
col_2_by_name <- my_df["age"]
str(col_2_by_name)
```

```{r}
vec_2_by_name <- my_df[["age"]]
str(vec_2_by_name)
```

Finally `my_df$age` is simply a shorthand for `my_df[["age"]]` without
the `[[ ]]` and the `" "`:

```{r}
vec_2_by_dollar_name <- my_df$age
str(vec_2_by_dollar_name)
```

## Direct comparison with tidyverse functions

The dplyr command `select()` over a data frame is exactly analogous to
the single bracket operator `my_df["age"]`. It returns a data frame with a
single column:

```{r}
col_2_select <- select(my_df, "age")
col_2_select
str(col_2_select)
```

The dplyr command `pull()` over a data frame is exactly analogous to the
double bracket operator `my_df[["age"]]`. It returns the data vector inside
that column:

```{r}
col_2_pull <- pull(my_df, "age")
col_2_pull
str(col_2_pull)
```

The 'problem' with these dplyr functions is that they *require* a data frame as 
input, and we recently saw in [S03E01:
t-tests](https://biodash.github.io/codeclub/s03e01_ttests/) that the
statistical t-test output was *not* a data frame:

```{r}
pop1 <- rnorm(n = 20, mean = 10, sd = 3)
pop2 <- rnorm(n = 20, mean = 10, sd = 3)

tresult <- t.test(x = pop1, y = pop2)
str(tresult)
```

This is not a data frame, but an 'htest' class object. Further, it
cannot be converted to a data frame in the usual way:

``` r
as.data.frame(tresult)
Error in as.data.frame.default(tresult): cannot coerce class ‘"htest"’ to a data.frame
```

This is precisely why the tidyverse developed the `broom::tidy()` function,
which works with legacy base-R outputs, and converts them to data frames. But if you have lots of t-tests, the overhead
of converting all the outputs using broom, then using dplyr functions to
access data, can be inefficient and overkill.

The t-test output is not a data frame, but it **is** a named list, so we can subset it directly. For example, to pull out the `p.value` we can do either:

```{r}
tresult[[3]]
```

or

```{r}
tresult$p.value
```

which is really much simpler than going through broom. In addition, we can get extra granularity
very quickly using this notation. Say we want the lower bound of the
confidence interval. We can 'stack' indexes:

```{r}
tresult[[4]][[1]]
```

or

```{r}
tresult$conf.int[[1]]
```

This is saying 'give me the 4th element (or the conf.int element), and then give me the 1st element of that'.

------------------------------------------------------------------------

### Exercise 2

:::puzzle

Reuse the t.test() code above, run `str` on the output, and extract the `stderr` value using the $ indexing approach.

<br>

<details> <summary> Solution (click here) </summary>

```{r}
tresult$stderr
```

<br>

</details>

:::

------------------------------------------------------------------------


### Exercise 3

:::puzzle

When we ran our first ANOVA, we never actually looked at the data structure that was produced.

Run the following code and inspect the output.

```r
bill_length_anova <- 
  aov(data = penguins %>% drop_na(),
      bill_length_mm ~ species + sex + species*sex)

str(bill_length_anova)
```

Aieee! 

What happens when you try to turn this into a data frame, using `as.data.frame(bill_length_anova)`?

Now you can see why `broom:tidy()` is so useful! But we can still extract values from this directly.

See if you can extract the residual df from this data structure.

<br>

<details> <summary> Solution (click here) </summary>

```{r}
bill_length_anova <- 
  aov(data = penguins %>% drop_na(),
      bill_length_mm ~ species + sex + species*sex)

str(bill_length_anova)
```


```r
as.data.frame(bill_length_anova)

Error in as.data.frame.default(bill_length_anova) : 
  cannot coerce class ‘c("aov", "lm")’ to a data.frame
```

```{r}
bill_length_anova$df.residual
```

<br>

</details>

:::


------------------------------------------------------------------------

<br> <br> <br> <br>
