---
title: "R4DS 3.5-3.6"
author: "Jessica Cooperstone"
date: "6/28/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load libraries
```{r}
library(tidyverse)
```
# Data
This chapter is using the `mpg` dataset. Let's see what's in there.
```{r}
glimpse(mpg)
```


# Facets

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_wrap(~ class, nrow = 2)
```

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_grid(drv ~ cyl)
```

## Exercises

### 3.5.1 continuous variable
What happens if you facet on a continuous variable?

Let's look back at `mpg` to see which are continuous variables (i.e. doubles/numeric) and which are factors, integers, characters, etc.

```{r}
glimpse(mpg)
```


```{r}

```

### 3.5.2 empty cells
What do the empty cells in plot with `facet_grid(drv ~ cyl)` mean? How do they relate to this plot?

```{r}

```

```{r}

```

You could also look at it like this using dplyr and tidyverse wrangling tools which to me is more intuitive:
```{r}

```


### 3.5.3 `.`
What plots does the following code make? What does . do?

```{r}

```

```{r}

```

### 3.5.4 Facets vs. other aesthetics
Take the first faceted plot in this section:

```{r}

```

What are the advantages to using faceting instead of the colour aesthetic? What are the disadvantages? How might the balance change if you had a larger dataset?
```{r}

```

```{r}

```

### 3.5.5 `ncol` and `nrow` in `facet_wrap`
Read `?facet_wrap`. What does nrow do? What does ncol do? What other options control the layout of the individual panels? Why doesn’t facet_grid() have nrow and ncol arguments?

```{r}

```



### 3.5.6 Variable with more levels as columns
When using facet_grid() you should usually put the variable with more unique levels in the columns. Why?

```{r}

```

```{r}

```

# Geometric objects (i.e. geoms)

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy))
```

```{r}
ggplot(data = mpg) + 
  geom_smooth(mapping = aes(x = displ, y = hwy))
```

```{r}
ggplot(data = mpg) + 
  geom_smooth(mapping = aes(x = displ, y = hwy, linetype = drv))
```

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, color = drv)) +
  geom_smooth(mapping = aes(x = displ, y = hwy, color = drv, linetype = drv))
```

```{r}
ggplot(data = mpg) +
  geom_smooth(mapping = aes(x = displ, y = hwy))
              
ggplot(data = mpg) +
  geom_smooth(mapping = aes(x = displ, y = hwy, group = drv))
    
ggplot(data = mpg) +
  geom_smooth(
    mapping = aes(x = displ, y = hwy, color = drv),
    show.legend = FALSE
  )
```


```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) +
  geom_smooth(mapping = aes(x = displ, y = hwy))
```

With less redundancy
```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point() + 
  geom_smooth()
```

Adding different aesthetics per geom
```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point(mapping = aes(color = class)) + 
  geom_smooth()
```

Having different data in each layer
```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point(mapping = aes(color = class)) + 
  geom_smooth(data = filter(mpg, class == "subcompact"), se = FALSE)
```

## Exercises

### 3.6.1 Investigating geoms
What geom would you use to draw a line chart? A boxplot? A histogram? An area chart?

Line chart:
Boxplot:
Histogram:
Area chart:

### 3.6.2 Mind coding
Run this code in your head and predict what the output will look like. Then, run the code in R and check your predictions.

```{r, eval = FALSE}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy, color = drv)) + 
  geom_point() + 
  geom_smooth(se = FALSE)
```


### 3.6.3 show.legend = FALSE
What does show.legend = FALSE do? What happens if you remove it?
Why do you think I used it earlier in the chapter?

With legend
```{r}

```
Without legend
```{r}

```

### 3.6.4 `se`
What does the se argument to geom_smooth() do?

```{r}

```

Is `se = TRUE` the default? How would you know?

### 3.6.5 Will graphs look different?
Will these two graphs look different? Why/why not?
```{r, eval = FALSE}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point() + 
  geom_smooth()

ggplot() + 
  geom_point(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_smooth(data = mpg, mapping = aes(x = displ, y = hwy))
```
### 3.6.7 Code from plots
Recreate the R code necessary to generate the following graphs.

Graph 1
```{r}

```

Graph 2
```{r}

```

Graph 3
```{r}

```

Graph 4
```{r}

```

Graph 5
```{r}

```

Graph 6
```{r}

```
