---
output: hugodown::md_document
title: "S05E02: R4DS (2e) - Ch. 6 - Tidy Data"
subtitle: "Today, we'll start with the R4DS chapter on \"tidy\" data, learning what it means for a dataframe to be in a tidy format, and how to reshape untidy into tidy data."
summary: "Today, we'll start with the R4DS chapter on \"tidy\" data, learning what it means for a dataframe to be in a tidy format, and how to reshape untidy into tidy data."
authors: [stephen-opiyo]
tags: [codeclub, r4ds]
date: "`r Sys.Date()`"
lastmod: "`r Sys.Date()`"
toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE, message = FALSE, cache = FALSE)
knitr::opts_knit$set(root.dir = here::here("content/codeclub/S05E02"))
```

------------------------------------------------------------------------

## Introduction

Today we will learn about a consistent way to organize your data in R,
using a system called **"tidy" data**. 

Then we will cover the primary tool use for tidying data, **pivoting**.
Pivoting allows you to change the form of your data without changing any of the values.

We will again be using _tidyverse_ packages,
so we'll need to load it packages for the current R session
using the `library()` function:

```{r}
library(tidyverse)

# You should have the tidyverse installed, but if not, do so using:
# install.packages("tidyverse")
```

<br>

------------------------------------------------------------------------

## What is tidy data?

We can represent any dataset in multiple ways.
Let us look at the following dataframes,
each of which comes is available once you've loaded the _tidyverse_:
`table1`, `table2`, `table3`, `table4a`, and `table4b`.

Each dataframe (or _tibble_, as _tidyverse_-style dataframes are called)
shows the same values for four variables:
`country`, `year`, `population`, and cases of TB (tuberculosis),
but each dataset organizes the values in a different way:

```{r}
table1
```

```{r}
table2
```

```{r}
table3
```

```{r}
table4a

table4b
```

Among these four formats,
the one in `table1` is easier to work with, certainly inside the _tidyverse_,
because it is "tidy".

Three rules that make a dataset tidy are:

1. Each variable is a column; each column is a variable.

2. Each observation is a row; each row is an observation.

3. Each value is a cell; each cell is a single value.

### An example of a computation with a tidy dataframe

The key advantage of having your data in a tidy format is that
it makes it easier to work with:
compute summaries, new variables, make plots, etc.

For instance, `table1` makes it straightforward to calculate a rate of cases
(cases per 10,000 people) using the `mutate()` function:

```{r}
# Compute rate per 10,000
table1 |>
  mutate(
    rate = cases / population * 10000
  )
```

Note that `mutate()` always adds new columns at the end of your dataset --
in this case, `rate`.
This function is discussed [here in R4DS Chapter 4](https://r4ds.hadley.nz/data-transform.html#sec-mutate).

The book also gives two other examples of working with a tidy dataframe,
including how to make a plot with _ggplot2_.

<br>

------------------------------------------------------------------------

## Breakout Rooms 1

(These are the same as Exercises 1 and 2 in
[R4DS 6.2.1](https://r4ds.hadley.nz/data-tidy.html#exercises).)

1. Using prose,
   describe how the variables and observations are organised in each of the
   sample tables (`table1`, `table2`, etc.).

2. Sketch out the process you’d use to calculate the `rate` for `table2`,
   and for `table4a` + `table4b`.
   You will need to perform four operations:

  - Extract the number of TB cases per country per year.
  - Extract the matching population per country per year.
  - Divide cases by population, and multiply by 10000.
  - Store back in the appropriate place.
  
  You haven’t yet learned all the functions you’d need to actually perform these      
  operations,
  but you should still be able to think through the transformations you’d need.

<br>

------------------------------------------------------------------------

## Pivoting

In the real world, data is often untidy because of two reasons:

1) Data is often organised to facilitate some goal other than analysis.

2) Most people aren’t familiar with the principles of tidy data.

Therefore, untidy data has to be pivot to pivot your data into a tidy form,
with variables in the columns and observations in the rows. 

### Two functions for pivoting data

- **`pivot_longer()`** makes datasets longer by increasing the number of rows,
  and reducing the number of columns.
  _Typically_, we use `pivot_longer()` to make an untidy dataset tidy,
  since untidy datasets often have multiple columns for a single variable,
  and multiple observations in a single row.

- **`pivot_wider()`** makes datasets wider by increasing the number of columns,
  and reducing the number of rows.
  _Typically_, `pivot_wider()` will make a dataset untidy --
  but that can still be useful, as we'll see next week.

### How does pivoting work?

Let’s start with a very simple dataset to make it easier to see what’s happening:

```{r}
# Row-wise tibble creation with the 'tribble()' function
df <- tribble(
  ~var, ~col1, ~col2,
   "A",     1,     2,
   "B",     3,     4,
   "C",     5,     6
)

df
```

We can tidy `df` using `pivot_longer()`:

```{r}
df |> 
  pivot_longer(
    cols = col1:col2,
    names_to = "names",
    values_to = "values"
  )
```

It’s easier to see how pivot_longer works if we take it component by component.
Let us look at how it works visually using Figure 6.3, Figure 6.4, and Figure 6.5
in the book.

### Example with data in column names

This example will work with the `billboard` dataframe that is loaded along
with the _tidyverse_.
This dataset records the billboard rank of songs in the year 2000:

```{r}
billboard 
```

We will use `pivot_longer()` to tidy this dataframe:

```{r}
billboard |> 
  pivot_longer(
    cols = starts_with("wk"), 
    names_to = "week", 
    values_to = "rank"
  )
```

Pivot longer has the following key arguments:

- `cols` -- these are the columns that should be combined into 2 columns,
  namely:
- `names_to` -- the name you would like to give to the new column whose values
  will hold what were the _column names_ in the original dataframe
- `values_to` -- the name you would like to give to the new column whose values
  will hold what were the _column valies_ in the original dataframe.

You can see that we get some `NA`s after pivoting.
In this case, `NA`s are forced to exist because of the structure of the dataset:
that is, they aren't actual missing values.
It would therefore make sense to get rid of them,
which we can do with `values_drop_na = TRUE`:

```{r}
billboard |> 
  pivot_longer(
    cols = starts_with("wk"), 
    names_to = "week", 
    values_to = "rank",
    values_drop_na = TRUE
  )
```

This data is now tidy,
but we could make future computations a bit easier by converting the values
in the column `week` (`wk1`, `wk2`, etc.) into a number
using `mutate()` and `readr::parse_number()`.
`parse_number()` is a handy function that will extract the numbers from a string,
removing all other characters:

```{r}
billboard_tidy <- billboard |> 
  pivot_longer(
    cols = starts_with("wk"), 
    names_to = "week", 
    values_to = "rank",
    values_drop_na = TRUE
  ) |> 
  mutate(
    week = parse_number(week)
  )

billboard_tidy
```

Now we’re in a good position to look at how song ranks vary over time by drawing a plot. The code is shown below and the result is Figure 6.2.

```{r}
billboard_tidy |> 
  ggplot(aes(x = week, y = rank, group = track)) + 
  geom_line(alpha = 1/3) + 
  scale_y_reverse()
```

<br>

------------------------------------------------------------------------

## Breakout Rooms 2

### Exercise 1

A) Create a dataset `table_pivot_long1` from `table1` with the following variables:
   `country`, `year`, `population`, `type`, and `count`.

Hint use starts_with("case), values_to = "count"

B) What are the number of rows and columns of `table_pivot_long1`?

Solution question 1

```{r}
#a)
table_pivot_long1 <- table1 |>
  pivot_longer(cols = starts_with("cases"), 
                            names_to = "type", 
                            values_to = "count")

# b)The dimension of table_pivot_long1 is 6 observations and 5 variables
```

### Exercise 2

Compute the rate per 10,000 for `table_pivot_long1`,
and name the new computed variable `rate`.

Hints use mutate, count, population

Solution question 2

```{r}
table_pivot_long1 |>
  mutate(rate = count / population * 1000)
```
  
<br>

----

<br>
