### The `next` statement

In the above examples, we have used `if` statements to perform some operations
only if the test evaluated to `TRUE`.
So our operations were done _inside_ the `if` statement.

This can also be "turned around" -- inside the `if` statement, we can
**instruct the loop to skip the rest of the current iteration using the `next` statement**.
Then, any operations that we code after that will only be executed when
the `if` statement test was `FALSE`:

```{r}
for (one_number in 1:10) {
  if (one_number <= 7) next
  print(one_number)
}
```

This can especially be useful if you want to do a whole bunch of tests,
because it avoids burying your code in multiple nested `if` statements.

<br>

------

As a side note, the variable as it was last assigned in the loop does
**persist in your environment**:

```{r}
i
```


### Looping over all indices

Sometimes you want to loop over indices rather than e.g. the values of a vector itself,
like we did in exercise 2 with `1:length(first_names)`.

To loop over all column indices of a dataframe, you can similarly use `ncol()`
(which returns the number of columns in a dataframe) --
for example, `1:ncol(penguins)`.

An alternative is using `seq_along()`, which will create an index for you:

```{r}
birds <- c("titmouse", "chickadee", "cardinal")
seq_along(birds)
```

The advantage of `seq_along()` is that it will behave better when your vector
accidentally has length 0.

Here is an example of using `seq_along()` in a loop:

```{r}
for (index in seq_along(birds)) {
  print(glue("Bird number {index} is {birds[index]}"))
}
```


-----


In case we wanted to calculate a statistic for each vector which would require
multiple steps, it might be efficient to write our own function.
If we imagine that no function exists to compute the mean:

```{r}
my_mean <- function(x) {
  sum(x) / length(x)
}
map(measurements, my_mean)
```

```{r}
penguins_noNA <- drop_na(penguins)

bill_length_mm <- head(penguins_noNA$bill_length_mm, 20)
bill_depth_mm <- head(penguins_noNA$bill_depth_mm, 20)
flipper_length_mm <- head(penguins_noNA$flipper_length_mm, 20)
body_mass_g <- head(penguins_noNA$body_mass_g, 20)
```

To compute the mean for each measurement, we could do:

```{r}
mean(bill_length_mm)
mean(bill_depth_mm)
mean(flipper_length_mm)
mean(body_mass_g)
```

Let's start by putting our four vectors in a list:

```{r}
measurements <- list(bill_length_mm, bill_depth_mm,
                     flipper_length_mm, body_mass_g)
```

    ```{r}
    for (measurement in measurements) {
      print(mean(measurement))
    }
    ```

```{r}
## Alternatively, create the new column base-R-style:
gent <- penguins %>%
  filter(species == "Gentoo") %>%
  drop_na()
gent$ratio <- ifelse(gent$bill_length_mm / gent$bill_depth_mm > 3.5,
                     "> 3.5", "< 3.5")
```

