---
output: hugodown::md_document
title: "S04E13: R for Data Science - Chapter 5.6: summarize"
subtitle: "The first of two parts on _dplyr_'s `summarize()` function"
summary: "Today we will introduce summarize() function. Together with group_by(), this function is extremely useful to produce summary statistics of your data by group."
authors: [mike-sovic]
tags: [codeclub, r4ds]
date: "`r Sys.Date()`"
lastmod: "`r Sys.Date()`"
toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE, message = FALSE, cache = FALSE)
```

------------------------------------------------------------------------

## Get the Dataset and Packages

```{r, eval=FALSE}
# Only if you haven't done so before, install the packages
install.packages("tidyverse")
install.packages("nycflights13")
```

```{r}
# Load the flights dataset and tidyverse
library(nycflights13)
library(tidyverse)
```

### Preview the Dataset

```{r}
glimpse(flights)
```

<br>

------------------------------------------------------------------------

## Review of `select()`, `filter()`, `mutate()`

### `select()`

The flights tibble has 19 variables --
to keep things simple, we'll focus on just a few of these for now.
Let's choose the variables (columns) `carrier`, `flight`, `air_time`, and `dep_delay`:

```{r}
flights_exp <- select(flights, carrier, flight, air_time, dep_delay)
flights_exp
```

### `filter() `

There are >336,000 observations (flights) in this dataset.
Let's reduce it to just American Airlines flights:

```{r}
flights_exp <- filter(flights_exp, carrier == "AA")
flights_exp
```

### `mutate()`

The column `air_time` is measured in minutes.
What if we wanted a new column `air_time_hrs` that reports the air time in hours?

```{r}
flights_exp <- mutate(flights_exp, air_time_hrs = air_time/60)
flights_exp
```

<br>

------------------------------------------------------------

## Section 5.6: `summarise()`

Here's an example of how to use `summarise()` in the simplest way.
Notice the similarity in syntax with `mutate()`:

```{r}
summarise(flights, delay = mean(dep_delay, na.rm = TRUE))
```

`summarise()` is typically combined with `group_by()`,
calculating the new summarized variable separately for each set of
grouped observations in the dataset:

```{r}
by_day <- group_by(flights, year, month, day)

summarise(by_day, delay = mean(dep_delay, na.rm = TRUE))
```

And more than one new summarized variable can be calculated - here, there are 3:

```{r}
by_dest <- group_by(flights, dest)

delay <- summarise(by_dest,
                   count = n(),
                   dist = mean(distance, na.rm = TRUE),
                   delay = mean(arr_delay, na.rm = TRUE))

delay
```

Let's look a bit closer at what this is doing (note, this is not in the book):

```{r}
flights_sub <- select(flights, dest, distance, arr_delay)
flights_sub
```

```{r}
albuquerque_data <- filter(flights_sub, dest == "ABQ")
albuquerque_data
```

```{r}
mean_albuquerque_delay <- mean(albuquerque_data$arr_delay, na.rm = TRUE)
mean_albuquerque_delay
```

Compare the value `mean_albuquerque_delay` to that in the summarized `delay`
tibble created above:

```{r}
mean_albuquerque_delay

delay
```

<br>

----------------------------------------------------------------

## Breakout Exercises 1

Start with the original `flights` tibble for each of the following exercises.

### Exercise 1

<div class="alert puzzle">
<div>

Overall, which carrier had the longest mean arrival delay (*arr_delay*) in 2013?
(Note, all these data are from 2013.)

<details> <summary> Hints (click here) </summary>
<br>

Group the tibble by the *carrier* variable,
then summarise to calculate the mean *arr_delay* for each group.
Remember to ignore missing values by setting `na.rm = TRUE` within the `mean()` function.

</details>

<details> <summary> Solution (click here) </summary>
<br>

```{r}
carrier_grp <- group_by(flights, carrier)

summarise(carrier_grp,
          mean_delay = mean(arr_delay, na.rm = TRUE)) 
```    

</details>

</div></div>

------------------------------------------------------------------------

### Exercise 2

<div class="alert puzzle">
<div>

Evaluate arrival delay by carrier again, but this time,
evaluate only for carriers who made at least 10,000 flights.

<details> <summary> Hints (click here) </summary>
<br> 

Include a second variable in the summarized data from above to reflect the
number of observations that went in to calculating the mean value for each group,
then filter on this sample size variable. Consider using the function `n()`.

</details>

<details> <summary> Solution (click here) </summary>
<br>

```{r}
carrier_delays <- summarise(carrier_grp,
                            mean_delay = mean(arr_delay, na.rm = TRUE),
                            n = n())

filter(carrier_delays, n >= 10000)
```

</details>

</div></div>

------------------------------------------------------------------------

### Exercise 3

<div class="alert puzzle">
<div>

Based on these data,
what airport is the worst to fly in to with regards to arriving on time?
In other words,
which airport (`dest`) is associated with the highest average arrival delays?

<details> <summary> Hints (click here) </summary>
<br> 

Group the tibble by the `dest` variable,
summarise with the mean `arr_delay` for each group,
then use `arrange()` to sort the new variable in descending order,
which can be done with `desc()`.

</details>

<details> <summary> Solution (click here) </summary>
<br>

```{r}
dest_grp <- group_by(flights, dest)

dest_delays <- summarise(dest_grp, mean_delay = mean(arr_delay, na.rm = TRUE))

arrange(dest_delays, desc(mean_delay))
```

</details>

</div></div>

<br>

------------------------------------------------------------------------

## Using Pipes

We did this above...

```{r}
by_dest <- group_by(flights, dest)
delay <- summarise(by_dest,
                   count = n(),
                   dist = mean(distance, na.rm = TRUE),
                   delay = mean(arr_delay, na.rm = TRUE))
```

That code can be rewritten with the pipe as follows...

```{r}
delays <- flights %>% 
  group_by(dest) %>% 
  summarise(
    count = n(),
    dist = mean(distance, na.rm = TRUE),
    delay = mean(arr_delay, na.rm = TRUE)
  )
```

Notice here we didn't have to create intermediate/temporary tibbles.

<br>

------------------------------------------------------------------------

## Breakout Exercises 2

Here we'll redo the same analyses as in the first breakout session,
but this time using the pipe.

### Exercise 1

<div class="alert puzzle">
<div>

Overall, which carrier had the longest mean arrival delay (*arr_delay*) in 2013?
(Note, all these data are from 2013.)

<details> <summary> Solution (click here) </summary>
<br>

```{r}
flights %>% 
  group_by(carrier) %>%
  summarise(mean_delay = mean(arr_delay, na.rm = TRUE))
```    

</details>

</div></div>

------------------------------------------------------------------------

### Exercise 2

<div class="alert puzzle">
<div>

Evaluate arrival delay by carrier again, but this time,
evaluate only for carriers who made at least 10,000 flights.

<details> <summary> Solution (click here) </summary>

```{r}
flights %>%
  group_by(carrier) %>%
  summarise(mean_delay = mean(arr_delay, na.rm = TRUE),
            n = n()) %>%
  filter(n > 10000)
```

</details>

</div></div>

------------------------------------------------------------------------

### Exercise 3

<div class="alert puzzle">
<div>

Based on these data,
what airport is the worst to fly in to with regards to arriving on time?
In other words,
which airport (`dest`) is associated with the highest average arrival delays?

<details> <summary> Solution (click here) </summary>
<br>

```{r}
flights %>%
  group_by(dest) %>%
  summarise(mean_delay = mean(arr_delay, na.rm = TRUE)) %>%
  arrange(desc(mean_delay))
```

</details>

</div></div>

------------------------------------------------------------------------
