---
output: hugodown::md_document
title: "S04E12: R for Data Science - Chapter 5.5: mutate"
subtitle: "Data Transformation with _dplyr_, part IV: `mutate()`"
summary: "Today we will cover the `mutate()` function to create new columns in dataframes. While this function itself is simple enough, we will get to see some interesting data manipulation techniques and operators such as `%%` and `%/%`."
authors: [stephen-opiyo]
tags: [codeclub, r4ds]
date: "`r Sys.Date()`"
lastmod: "`r Sys.Date()`"
toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE, message = FALSE, cache = FALSE)
```

<figure><p align="center">
<img src=img/mutate.jpeg width="70%">
<figcaption>Artwork by Allison Horst</figcaption>
</p></figure>

------------------------------------------------------------------------

## Introduction

Today we are going to cover the dplyr function `mutate()`.
You can today's topic in the R 4 Data Science book at:
<https://r4ds.had.co.nz/transform.html#add-new-variables-with-mutate>.

We will again be using the `nycflights13` and `tidyverse` packages,
so we first need make sure these packages are installed,
and then load them for the current session by doing `library()` commands:

```{r, eval=FALSE}
## Only if you haven't done so before, install the packages
install.packages("tidyverse")
install.packages("nycflights13")
```

```{r}
## Load the packages
library(tidyverse)
library(nycflights13)
```

<br>

------

## The `mutate()` function

The `mutate()` function always _adds_ new columns at the end of your dataset.
Let us create a small dataset so we can see the new variables:

```{r}
flights_sml <- select(flights, 
  year:day, 
  ends_with("delay"), 
  distance, 
  air_time
)
```

As a first example, let's add columns `gain` (delay made up) and `speed`:

```{r, echo=FALSE}
mutate(flights_sml,
  gain = dep_delay - arr_delay,
  speed = distance / air_time * 60
)
```

```{r, eval=FALSE}
mutate(flights_sml,
  gain = dep_delay - arr_delay,
  speed = distance / air_time * 60
)
```

`transmute()` is very similar to `mutate()`,
except the returned dataframe only contains the new variables that were created:

```{r, eval=FALSE}
transmute(flights,
  gain = dep_delay - arr_delay,
  hours = air_time / 60,
  gain_per_hour = gain / hours
)
```

```{r, echo=FALSE}
transmute(flights,
  gain = dep_delay - arr_delay,
  hours = air_time / 60,
  gain_per_hour = gain / hours
)
```

<br>

------

## Useful creation functions to use with mutate

Note that functions must be vectorised before use with `mutate()`.

**Arithmetic operators:**
`+` (addition), `-` (subtraction), `*` (multiplication), `/` (division), `^` (to the power)

**Modular arithmetic:**
`%/%` (integer division) and `%%` (remainder).
Modular arithmetic is a handy tool because it allows you to break integers up into pieces.
For example:

```{r, eval=FALSE}
transmute(flights,
  dep_time,
  hour = dep_time %/% 100,
  minute = dep_time %% 100
)
```

```{r, echo=FALSE}
transmute(flights,
  dep_time,
  hour = dep_time %/% 100,
  minute = dep_time %% 100
)
```

**Logs:** `log ()`, `log2()`, `log10()`.
Logarithms are useful transformation for dealing with data that ranges across
multiple orders of magnitude.

**Offsets:** `lead()` and `lag()`.
These allow you to refer to leading or lagging values:

```{r, eval=FALSE}
(x <- 1:10)
```

```{r, echo=FALSE}
(x <- 1:10)
```

```{r, eval=FALSE}
lag(x)
```

```{r, echo=FALSE}
lag(x)
```

```{r, eval=FALSE}
lead(x)
```

```{r, echo=FALSE}
lead(x)
```

**Cumulative and rolling aggregates:**
R provides functions for sums `cumsum()`, products `cumprod()`, mins `cummin()`,
and maxes `cummax()`;
and dplyr provides function for mean `cummean()`.

```{r, eval=FALSE}
x
```

```{r, echo=FALSE}
x
```

```{r, eval=FALSE}
cumsum(x)
```

```{r, echo=FALSE}
cumsum(x)
```

```{r, eval=FALSE}
cummean(x)
```

```{r, echo=FALSE}
cummean(x)
```

**Logical comparisons:**
`<` (less than), `<=` (equal to or less than), `>` (greater than),
`>=` (equal to or greater than), `!=` (not equal to), and `==` (equal to).

**Ranking:** `min_rank()` gives the smallest values the small rank,
and `desc(x)` to give the largest values the smallest ranks.

```{r, eval=FALSE}
y <- c(1, 2, 2, NA, 3, 4)
```

```{r, echo=FALSE}
y <- c(1, 2, 2, NA, 3, 4)
```

```{r, eval=FALSE}
min_rank(y)
```

```{r, echo=FALSE}
min_rank(y)
```

```{r, eval=FALSE}
min_rank(desc(y))
```

```{r, echo=FALSE}
min_rank(desc(y))
```

There are other ranks functions, too,
e.g., `row_number()`, `dense_rank()`, `percent_rank()`, `cume_dist()`, `ntile()`.

```{r, eval=FALSE}
row_number(y)
```

```{r, echo=FALSE}
row_number(y)
```

```{r, eval=FALSE}
dense_rank(y)
```

```{r, echo=FALSE}
dense_rank(y)
```

```{r, eval=FALSE}
percent_rank(y)
```

```{r, echo=FALSE}
percent_rank(y)
```

<br>

------

## Breakout room exercises

These correspond to the
[exercises in the R4DS book](https://r4ds.had.co.nz/transform.html#exercises-11),
but we're skipping exercises 2 and 6.

### R4DS Exercise 1

Currently `dep_time` and `sched_dep_time` are convenient to look at,
but hard to compute with because they’re not really continuous numbers.
Convert them to a more convenient representation of number of minutes
since midnight.

<details><summary><b>Hints</b> (click here)</summary>
<br>

- You can use the following code structure
  (we first select only relevant columns to more easily check the results):

```{r, eval=FALSE}
flights %>% 
  select(dep_time, sched_dep_time) %>% 
  mutate(
    flights_times,
    dep_time_mins = YOUR_CODE_HERE,
    sched_dep_time_mins = YOUR_CODE_HERE
    )
```

- Get the number of _hours_ since midnight using `%/%`
  (as in the example in the book) and then convert this number to minutes.
  
- Then, get the number of _minutes_ past the hour using `%%`
  (again as in the example in the book), and add this number to the previous
  one to get the total number of minutes past midnight.
  
- Now, you might think you're all done, but there is one remaining problem:
  midnight was originally represented as `2400`,
  which the code described above would convert to `1440` (`24 * 60`).
  But it should be `0` instead.
  This is a tricky one,
  but you can handle it using an additional `%%` calculation.

</details>

<details><summary><b>Solution</b> (click here)</summary>
<br>

- To handle the midnight case, use `%% 1440`:
  this would simply return the original value for all values below 1440,
  and would return `0` only for `1440` (midnight): 

```{r, eval=FALSE}
flights %>%
  select(dep_time, sched_dep_time) %>% 
  mutate(
    dep_time_mins = (dep_time %/% 100 * 60 + dep_time %% 100) %% 1440,
    sched_dep_time_mins = (sched_dep_time %/% 100 * 60 + sched_dep_time %% 100) %% 1440
    )
```

```{r, echo=FALSE}
flights %>%
  select(dep_time, sched_dep_time) %>% 
  mutate(
    dep_time_mins = (dep_time %/% 100 * 60 + dep_time %% 100) %% 1440,
    sched_dep_time_mins = (sched_dep_time %/% 100 * 60 + sched_dep_time %% 100) %% 1440
    )
```

</details>

### R4DS Exercise 3

Compare `dep_time`, `sched_dep_time`, and `dep_delay`.
How would you expect those three numbers to be related?
Test if this is indeed the case.

<details><summary><b>Hints</b> (click here)</summary>
<br>

- We should expect `dep_delay` to equal the difference between
  `sched_dep_time` and `dep_time`.
  
- To check this, the first step is to use your code from exercise 1 to create
  columns with the `sched_dep_time` and the `dep_time` in minutes past midnight.
  
- Then, create a column with the difference in minutes between `sched_dep_time`
  and `dep_time` (e.g. called `our_delay_calc`).

- Next, create a column with the difference between `our_delay_calc` and `dep_delay`
  (e.g. called `delay_diff`).

- Finally, use `filter()` to see if there are any rows where the `delay_diff`
  does not equal 0. Recall that we would expect no differences at all,
  if all is well. So what is going on with those rows?

</details>

<details><summary><b>Solution</b> (click here)</summary>
<br>

```{r, eval=FALSE}
flights %>% 
  select(dep_time, sched_dep_time, dep_delay) %>% 
  mutate(
    dep_time_min = (dep_time %/% 100 * 60 + dep_time %% 100) %% 1440,
    sched_dep_time_min = (sched_dep_time %/% 100 * 60 + sched_dep_time %% 100) %% 1440,
    our_delay_calc = dep_time_min - sched_dep_time_min,
    dep_delay_diff = our_delay_calc - dep_delay
  ) %>% 
  filter(dep_delay_diff != 0)
```

```{r, echo=FALSE}
flights %>% 
  select(dep_time, sched_dep_time, dep_delay) %>% 
  mutate(
    dep_time_min = (dep_time %/% 100 * 60 + dep_time %% 100) %% 1440,
    sched_dep_time_min = (sched_dep_time %/% 100 * 60 + sched_dep_time %% 100) %% 1440,
    our_delay_calc = dep_time_min - sched_dep_time_min,
    dep_delay_diff = our_delay_calc - dep_delay
  ) %>% 
  filter(dep_delay_diff != 0)
```

These `dep_delay_diff` values are all 1440 minutes (= 24 hours),
and as you can see, these are flights that were scheduled to depart one day
but left only the next day (past midnight).

So, for our `dep_time_min` and `sched_dep_time_min` to always work for calculations,
they should also take dates into account -- and should e.g. be minutes past some
date in the past.

</details>

### R4DS Exercise 4

Find the 10 most delayed flights using a ranking function.
How do you want to handle ties?
Carefully read the documentation for `min_rank()`.

<details><summary><b>Hints</b> (click here)</summary>
<br>

- Use `mutate()` to create a new column with the ranks,
  then arrange by this column and/or filter only top ranks to get the most
  delayed flights.
  
- To see the differences in the handling of ties between `row_number()`,
  `min_rank()`, and `dense_rank()`, take a look this:
  
```{r, eval=FALSE}
tibble(delays = c(3, 5, 5, 5, 130, 276),
       rank_rownr = row_number(delays),
       rank_min = min_rank(delays),
       rank_dense = dense_rank(delays))
```

```{r, echo=FALSE}
tibble(delays = c(3, 5, 5, 5, 130, 276),
       rank_rownr = row_number(delays),
       rank_min = min_rank(delays),
       rank_dense = dense_rank(delays))
```

</details>

<details><summary><b>Solution</b> (click here)</summary>
<br>

```{r}
flights %>%
  select(dep_time, sched_dep_time, dep_delay) %>%
  mutate(dep_delay_rank = min_rank(desc(dep_delay))) %>% 
  filter(dep_delay_rank <= 10) %>%
  arrange(dep_delay_rank)
```

</details>

### R4DS Exercise 5

What does `1:3 + 1:10` return? Why?

<details><summary><b>Hints</b> (click here)</summary>
<br>

- First, you should realize that `1:3` expands to a vector with the values
  `1`, `2`, and `3`, and similarly for `1:10`.
  
- Many R operations are _vectorized_, which means that when pairing two vectors
  (including the case where one of those vectors is just a single value),
  the shorter vector will be recycled.
  In the example below, `3` is recycled and added to every single value in the
  vector `1:5`:
  
```{r, eval=FALSE}
3 + 1:5
```

```{r, echo=FALSE}
3 + 1:5
```

</details>

<details><summary><b>Solution</b> (click here)</summary>
<br>

```{r, eval=FALSE}
1:3 + 1:10
```

```{r, echo=FALSE}
1:3 + 1:10
```

R gives a warning because the length of the longer vector isn't a multiple of
the length of the shorter vector: it recycles `1:3` three times and is then
left over with a single value from `1:3` to be paired with `10`.
This kind of thing is usually not intended.

</details>

### Bonus exercise

You can use the `paste0()` function to combine strings of text,
for instance:

```{r, eval=FALSE}
carrier <- "UA"
flight <- 1545
paste0("The full flight number is: ", carrier, flight)
```

```{r, echo=FALSE}
carrier <- "UA"
flight <- 1545
paste0("The full flight number is: ", carrier, flight)
```

Use `paste0()` inside `mutate()` to create a new column `flight_no`
which has the full flight number (carrier followed by flight, like above)
for each flight.

<details><summary><b>Solution</b> (click here)</summary>
<br>

```{r, eval=FALSE}
flights %>%
  select(carrier, flight) %>%
  mutate(flight_no = paste0(carrier, flight))
```

```{r, echo=FALSE}
flights %>%
  select(carrier, flight) %>%
  mutate(flight_no = paste0(carrier, flight))
```

</details>

<br> <br>

-----

## Acknowledgements

We used <https://jrnold.github.io/r4ds-exercise-solutions> to provide hints
and solutions for the exercises.
