---
title: "Code Club S02E05: Intro to the Tidyverse (Part 2)"
summary: "In this session of Code Club, we will be learning some more about data wrangling in the tidyverse."
author: [stephen-opiyo]
date: "2021-09-22"
output: hugodown::md_document
toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

----

## Prep homework 

### Basic computer setup

-   If you didn't already do this, please follow the [Code Club Computer Setup](/codeclub-setup/) instructions, which also has pointers for if you're new to R or RStudio.

-   If you're able to do so, please open RStudio a bit before Code Club starts -- and in case you run into issues, please join the Zoom call early and we'll troubleshoot.

<br>

----

## Introduction

**What will we go over today**

-   We will continue using *dplyr* package of the *tidyverse* introduced last week.
-   Learn using [`arrange()`](https://dplyr.tidyverse.org/reference/arrange.html) - orders the rows of a data frame by the values of selected columns. 
-   Learn using [`mutate()`](https://dplyr.tidyverse.org/reference/mutate.html) - adds new variables and preserves existing ones.

<br>

----

## 1 - What is the dplyr package?

[*dplyr*](https://dplyr.tidyverse.org/index.html) is one of the *tidyverse* packages
that are designed for data science. *dplyr* provides functions for data manipulation. 

Functions for row-wise operations include:

-   [`filter()`](https://dplyr.tidyverse.org/reference/filter.html) - chooses rows based on column values.
-   [`slice()`](https://dplyr.tidyverse.org/reference/slice.html) - chooses rows based on location.
-   [`arrange()`](https://dplyr.tidyverse.org/reference/arrange.html) - orders the rows of a data frame by the values of selected columns.

Functions for column-wise operations include:

-   [`select()`](https://dplyr.tidyverse.org/reference/select.html) - changes whether or not a column is included.
-   [`rename`()](https://dplyr.tidyverse.org/reference/rename.html) - changes the name of columns. 
-   [`mutate()`](https://dplyr.tidyverse.org/reference/mutate.html) - changes the values of columns and creates new columns
-   [`relocate`()](https://dplyr.tidyverse.org/reference/relocate.html) - changes the order of the columns.

Functions for *groups* of rows include:

-   [`summarise()`](https://dplyr.tidyverse.org/reference/summarise.html) - collapses a group into a single row.

Last week, we got introduced to the *tidyverse* and covered the pipe, `select()`, and `filter`.
We also discussed *tidyverse* packages such as [`ggplot2`](https://ggplot2.tidyverse.org/index.html),
a part of the core tidyverse, which we have talked about in previous Code Clubs ([intro](/codeclub/04_ggplot2), [intro2](/codeclub/05_ggplot-round-2), [maps](/codeclub/11_ggplot-maps), and [ggplotly](/codeclub/15_plotly)), and others.

We also saw that packages are basically R add-ons that contain additional functions or datasets we can use.
Using the function [`install.packages()`](https://www.rdocumentation.org/packages/utils/versions/3.6.2/topics/install.packages), we can install packages that are available at the Comprehensive R Archive Network, or [CRAN](https://cran.r-project.org/).  

For those who have not installed the *tidyverse*, let's install it. 
We only need to do this once, so if you did this last week, you don't need to now.

```{r, eval = FALSE}
install.packages("tidyverse")
```

To use the *dplyr* package within the *tidyverse*, we need to call it up using `library()`.

```{r}
library(tidyverse)
```

<br> 

------------------------------------------------------------------------

## 2 - Using the `arrange()` function

We will learn how to use the `arrange()` function from *dplyr* to sort a data frame in multiple ways.
First, we will sort a dataframe by values of a single variable,
and then we will learn how to sort a dataframe by more than one variable in the dataframe.
By default, *dplyr*’s `arrange()` sorts in _ascending_ order.

**Let's get set up and grab some data so that we have some material to work with.**

We will use the same dataset [`palmerpenguins`](https://allisonhorst.github.io/palmerpenguins/)
we used last week.
To get this data, we need to install the *palmerpenguins* package
(no need to do this if you did so last week):
 
```{r, eval = FALSE}
install.packages("palmerpenguins")
```

Then, to use the package, we need to use the function `library()` to load the package in R:

```{r}
library(palmerpenguins)
```

The dataframe we will use today is called `penguins`.
Let's take a look at the structure of the data:

```{r}
# look at the first 10 rows and all columns
head(penguins, 10)

# check the structure of penguins_data
# glimpse() which is a part of dplyr functions 
# similarly to str() and can be used interchangeably
glimpse(penguins)
```

Okay, now we have a sense of what the `penguins` dataset is.

Now we want to sort `penguins` dataframe by body mass to quickly learn about the lightest penguin and its relations to other variables.
We will use the pipe operator `%>%` to feed the data to the `arrange()` function.
We then specify name of the variable that we want to sort the dataframe by.

In this example, we are sorting by variable `body_mass_g`,
so we will see the lightest penguins at the top of the dataframe:

```{r}
penguins %>%           # take the penguins_data
  arrange(body_mass_g) # sort the dataframe in ascending order based on body mass
```

If we wanted to sort descendingly, such that the heaviest penguin are on top,
we can add a `-` in front of the variable:

```{r}
penguins %>%             # take the penguins_data
  arrange(-body_mass_g)  # sort the dataframe in descending order based on body mass
```

We can also pipe the results into the `filter()` function covered last week,
to select only penguins weighing more than 5000 g:

```{r}
penguins_new <-     # assign the results to a dataframe `penguins_new`
  penguins %>%                               # take the penguins data
  arrange(bill_length_mm, bill_depth_mm) %>% # sort by bill length followed by bill depth
  filter(body_mass_g > 5000)                 # select species greater with mass > 5000 g.
```

```{r}
head(penguins_new, 5)     # look at the top 5 
```

Let's check the counts of different species and islands among our new dataset:

```{r}
penguins_new %>%
  count(species, .drop = FALSE)   # .drop=FALSE will keep values of 0
```

```{r}
penguins_new %>%
  count(island, .drop = FALSE)
```

You can see that we have only retained Gentoo Penguins from the island of Biscoe.

<br>

------------------------------------------------------------------------

## Breakout session 1 - `arrange()`

<div class="alert puzzle">
<div>

### Exercise 1

With the `penguins` dataset, answer the following questions:

- Create a new dataset called `penguins_shortflippers` from the `penguins` dataset
  with the 20 penguins with the shortest flippers.

- How many penguins of each species are found in `penguins_shortflippers`? 

- Which islands do they come from?

<details>
<summary>
Hints
</summary>
<br>

- To create `penguins_shortflippers`, first use `arrange()` to sort by flipper lengths,
  and pipe the results into the `head()` function to get the top 20.

- To get the species and island composition, use the `count()` function.

</details>

<details>
<summary>
Solution (click here)
</summary>
<br>

- To create a dataframe with the 20 penguins with the shortest flippers:

```{r, eval = TRUE}
penguins_shortflippers <-        # assign the results
  penguins %>%                   # take penguins_data
  arrange(flipper_length_mm) %>% # sort the data by bill flipper length 
  head(20)                       # take the top 20
```

- To see the species composition in `penguins_shortflippers`:

```{r, eval = TRUE}
penguins_shortflippers %>%
  count(species)
```

- To see the island composition in `penguins_shortflippers`:

```{r, eval = TRUE}
penguins_shortflippers %>%
  count(island)
```

</details>

</div>
</div>

<br>

------------------------------------------------------------------------

## 3 - Using `mutate()`

Besides selecting sets of existing columns, it’s often useful to add new columns that are derived from existing columns. 
The `mutate()` function create new variables, usually by manipulating existing variables.

`mutate()` always adds new columns at the end of the dataframe.
When you use `mutate()`, you need typically to specify 3 things:

- the name of the dataframe you want to modify
- the name of the new column that you'll create
- the values to be inserted in the new column

We will be working with the penguins dataset to learn the `mutate()` function.
We will create a new dataframe called `mutate_penguins`, with a new column called `body_mass_g_new`.

The first argument (dataset to be piped) is the dataframe we're going to modify, `penguins`.
After that, we have the name-value pair for our new variable.

Here, the name of the new variable is `size` and the values are `body_mass_g` multiplied by
`flipper_length_mm`:

```{r}
mutate_penguins <-  # assign the results to a dataframe `mutate_penguins`
  penguins %>%      # take the penguins_data
  mutate(size = body_mass_g * flipper_length_mm) # create a new column
```

```{r}
head(mutate_penguins) %>%
  select(6:9)       # show the first rows of columns 6-9
```

You can see that we created data with a new column called `size`.

<br>

-----

## Breakout session 2 - `mutate()` 

<div class="alert puzzle">
<div>

### Exercise 2

- Create a new dataframe called `penguins_bills` with a new column called `bill_shape` by
  dividing bill length by bill depth.

- What is the species composition of the 20 penguins with the _largest_ values for `bill_shape`?

<details>
<summary>
Hints (click here)
</summary>
<br>

To get the species composition of the top 20,
first use `arrange()` (think about the direction you need to sort in!),
then `head()`, and then `count()`.

</details>

<details>
<summary>
Solution (click here)
</summary>
<br>

- New dataframe with a bill shape variable:

```{r}
penguins_bills <-
  penguins %>%      # take the penguins_data
  mutate(bill_shape = bill_length_mm / bill_depth_mm) # Create a new column `bill_shape`   
```

- Species composition of the 20 penguins with the largest `bill_shape` values:

```{r}
penguins_bills %>%
  arrange(-bill_shape) %>%  # sort by bill_shape in descending order
  head(20) %>%              # take the top 20
  count(species)            # create a frequency table
```

They are all Gentoo penguins!

</details>

</div>
</div>

<div class="alert puzzle">
<div>

### Exercise 3

Create a new dataframe called `penguins_year`:

- with only penguins sampled after 2007,
- with a new column called `year_nr` which has a year number that starts counting
from 2008 (i.e., 2008 = year 1, 2009 = year 2, etc.)
- sorted by `year_nr`.

<details>
<summary>
Hint (click here)
</summary>

Not all values you pass to `mutate()` need to be variables!
You can subtract `year` by a fixed number.

</details>

<details>
<summary>
Solution (click here)
</summary>

```{r}
penguins_year <-
  penguins %>%
  filter(year > 2007) %>%
  mutate(year_nr = year - 2007) %>%
  arrange(year_nr)
```

</details>

</div>
</div>

<br>

----

